---
title: "cruise_data"
author: "Angela Nankabirwa"
date: "2026-03-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


#loading packages

```{r}

library(tidyverse)
library(janitor)
library(stringi)  
library(lubridate)
library(readr)
library(knitr)
library(readxl)
library(ggpattern) 
library(gridExtra)
library(kableExtra) #formatting e,g. colour)
library(zoo) #remove noisy bumps from data #rolling median
library(specleanr) #20 ways to detect outliers
library(tidytext) #visualize outliers from specleanr package



```






#import all cruise ysi files
+   2022

+     MFANGANO SONDE
```{r}
file_paths.csv <- dir("2022_files", pattern = "*.csv", full.names = TRUE)



ysiF_2022 <- read_csv(file_paths.csv[3], skip=2, 
                      col_names = c("date_time", "temp", 
                                    "sp_cond", "salinity", "depth", "chl_ug_l",
                                    "chl_rfu_F", "bga_pc", "bga_rfu_F", "battery"))

#attach time stamps
ysiF_2022_b <- ysiF_2022 %>%
  mutate(date_time = as.POSIXct(date_time, format = "%m/%d/%Y %H:%M"),
         code = case_when(
           (date_time >= as.POSIXct("2022-07-12 8:20:00") & date_time <= as.POSIXct("2022-07-12 8:40:00")) ~ "BG10",
           (date_time >= as.POSIXct("2022-07-12 10:40:00") & date_time <= as.POSIXct("2022-07-12 11:00:00")) ~ "BG13",
           (date_time >= as.POSIXct("2022-07-12 12:50:00") & date_time <= as.POSIXct("2022-07-12 13:00:00")) ~ "BG16",
           (date_time >= as.POSIXct("2022-07-12 15:40:00") & date_time <= as.POSIXct("2022-07-12 15:50:00")) ~ "CNL1",
           (date_time >= as.POSIXct("2022-07-12 21:00:00") & date_time <= as.POSIXct("2022-07-12 21:20:00")) ~ "BG18",
           TRUE ~ NA_character_
         )) %>% 
  filter(!is.na(code))



```

+   checking for outliers

```{r}
#The values are very few. They were recorded every 30 minutes and so we will just do average


ysiF_22_outlier_detect <- multidetect(data = ysiF_2022_b, 
                                var = "chl_rfu_F", 
                                var_col = "code",
                                multiple = TRUE,
                                methods = c("adjbox", "iqr", "hampel","jknife",
                                            "seqfences", "mixediqr",
                                            "distboxplot", "semiqr",
                                            "zscore", "logboxplot", "medianrule"),
                       silence_true_errors = TRUE, 
                       missingness = 0.1,
                       sdm = FALSE,
                       na.inform = TRUE)
  
  

#Visualize the number of outliers detected by each method

ggoutliers(ysiF_22_outlier_detect, select = unique(ysiF_2022_b$code),
           color = "lightblue",
           ncol = 5,
           nrow = 1)



ggsave(file = "plots/2022_plots/2022MF_outlier_per_site23.png", width = 10, height = 5, dpi = 300)


#Obtaining quality controlled dataset using loess method or data labeling


out_qc_loess22F <- extract_clean_data(refdata = ysiF_2022_b, 
            outliers = ysiF_22_outlier_detect, loess = TRUE,
            var_col = "code",
            threshold = 0.7)  #set threshold to 0.8 to remove the very strong outliers only

#clean dataset
nrow(out_qc_loess22F)

#reference data
nrow(ysiF_2022_b)

out_qc_labeled22F <- classify_data(refdata = ysiF_2022_b, 
                                 var_col = "code", 
                                outliers = ysiF_22_outlier_detect)

#Visualize labelled quality controlled dataset

ggenvironmentalspace(out_qc_labeled22F, 
                     type = '1D',
                     ggxangle = 45, 
                     scalecolor = 'viridis',
                     xhjust = 1,
                     legend_position = 'blank',
                     ylab = "Number of records",
                     xlab = "Outlier labels",
                     ncol = 5,
                     nrow = 1)

ggsave(filename = "plots/2022_plots/labelled_outlier_chart_MF.png", width = 8, height = 5, dpi = 300)


#labelled outlier or not 
labeled22F <- classify_data(refdata = ysiF_2022_b, 
                           var_col = "code",
                           outliers = ysiF_22_outlier_detect)



#plot clean dataset

out_qc_loess22F %>% 
mutate(code = as.factor(code)) %>% 
ggplot(aes(x = chl_rfu_F)) +
  geom_histogram(bins=5,binwidth = 1, boundary = 0, fill = "steelblue", color = "black") +
facet_wrap(~ code, scales = "free_y")+
  labs(x = "Chlorophyll (RFU)", y = "Frequency") +
  theme_bw() +
  theme(text = element_text(size = 20, colour = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))+
 scale_x_continuous(
  breaks = scales::pretty_breaks(n = 6)  #ggplot makes breaks based on data
)

ggsave(file="plots/2022_plots/ysi22_loessfilter_MF_freq_distribution.png", width = 10, height = 14, dpi = 300)



#No strong outlier so we will just average the values

ysiF_2022_c <- ysiF_2022_b %>% 
  filter(!is.na(code)) %>% 
  group_by(code) %>% 
  summarise(
    chl_ysi = mean(chl_rfu_F, na.rm = TRUE),   
    bga_ysi = mean(bga_rfu_F, na.rm = TRUE),
    temp_ysi = mean(temp, na.rm = TRUE),
    sp_cond_ysi = mean(sp_cond, na.rm = TRUE)
  ) %>% 
  ungroup() %>% 
  mutate(sample_type = "cruise-2022_MFsonde")

```


+   Dunga sonde 2022

```{r}
#DUNGA sonde
ysi_2022DG <-read_csv(file_paths.csv[2], skip=2, col_names = c("date_time", "temp", 
                                       "sp_cond", "salinity",
                                                           "depth", "chl_ug_l",
                                  "chl_rfu_DG", "bga_pc", "bga_rfu_DG", "battery"))


ysi_2022DG_b <- ysi_2022DG %>%
  mutate(date_time = as.POSIXct(date_time, format = "%m/%d/%Y %H:%M"),
         code = case_when(
           (date_time >= as.POSIXct("2022-07-12 8:20:00") & date_time <= as.POSIXct("2022-07-12 8:40:00")) ~ "BG10",
           (date_time >= as.POSIXct("2022-07-12 10:50:00") & date_time <= as.POSIXct("2022-07-12 11:00:00")) ~ "BG13",
           (date_time >= as.POSIXct("2022-07-12 12:50:00") & date_time <= as.POSIXct("2022-07-12 13:00:00")) ~ "BG16",
           (date_time >= as.POSIXct("2022-07-12 15:40:00") & date_time <= as.POSIXct("2022-07-12 15:50:00")) ~ "CNL1",
           (date_time >= as.POSIXct("2022-07-12 21:00:00") & date_time <= as.POSIXct("2022-07-12 21:20:00")) ~ "BG18",
           TRUE ~ NA_character_
         )) %>% 
  filter(!is.na(code))


#for outliers, same situation as MF sonde, so we will just do the average

ysi_2022DG_c <- ysi_2022DG_b %>% 
  filter(!is.na(code)) %>% 
group_by(code) %>% 
  summarise(
    chl_ysi = mean(chl_rfu_DG, na.rm = TRUE),
    bga_ysi = mean(bga_rfu_DG, na.rm= TRUE),
    temp_ysi = mean(temp, na.rm=TRUE),
    sp_cond_ysi = mean(sp_cond, na.rm = TRUE)
    ) %>% 
  ungroup() %>% 
  mutate(sample_type = "cruise-2022_DGysi")
```

+     combine MF and DG sonde data

```{r}
####bind DG and MF sonde cruise

cruise_sondes_2022 <- bind_rows(ysi_2022DG_c, ysiF_2022_c) 



```



+ import files for 2023 and 2024

```{r}


#YSI files

# Get a list of file paths in the "files" directory

file_paths.csv <- dir("ysi_files", pattern = "*.csv", full.names = TRUE)


# Read the first file and inspect its column names
first_file <- read_csv(file_paths.csv[1])
colnames(first_file)

#function to read and combine csv files
read_and_combine_csv <- function(file_path) {
  tryCatch(
    {
      # Read the CSV file
      data <- read_csv(file_path)
      
      
      # Remove the first row
      data <- data[-1, ]
      
      # Columns to convert to double
      cols_to_convert <- c("Temp", "SpCond", "Sal", "Depth", "Chl...6", "Chl...7", "BGA-PC...8", "BGA-PC...9", "Battery")
      
      # Convert specified columns to double
      data[, cols_to_convert] <- lapply(data[, cols_to_convert], as.numeric)
      
      # Extract the file name without the ".csv" extension
      file_name <- tools::file_path_sans_ext(basename(file_path))
      
      # Add a new column "FileName" to the data frame, remove formats
      data <- data %>% mutate(FileName = file_name)%>% 
        clean_names() #clean headings 
      
      
      return(data)
    },
    error = function(e) {
      cat("Error occurred while processing file:", file_path, "\n")
      message("Error Message:", conditionMessage(e), "\n")
      traceback()
      return(NULL) # Return NULL to indicate failure
    }
  )
}

# Read all csv files and combine them
 df_ysi_cruz <- map_df(file_paths.csv, read_and_combine_csv) %>% 
    rename(chl_rfu = chl_7,
         bga_rfu = bga_pc_9) 
 


```

+   attach time stamps to 2023 cruz

+   **2023**
```{r}



ysi2023 <- df_ysi_cruz %>%
  filter(file_name == "cruise-ysidata-2023_naphtali") %>% 
  
  mutate(date_time = as.POSIXct(date_time, format = "%m/%d/%Y %H:%M"),
         code = case_when(
           (date_time >= as.POSIXct("2023-06-27 11:26:00") & date_time <= as.POSIXct("2023-06-27 11:44:00")) ~ "CHL1",
           (date_time >= as.POSIXct("2023-06-27 11:46:00") & date_time <= as.POSIXct("2023-06-27 12:49:00")) ~ "CHL2",
           (date_time >= as.POSIXct("2023-06-27 12:50:00") & date_time <= as.POSIXct("2023-06-27 14:10:00")) ~ "BG7",
           (date_time >= as.POSIXct("2023-06-27 14:11:00") & date_time <= as.POSIXct("2023-06-27 14:40:00")) ~ "CHL3",
           (date_time >= as.POSIXct("2023-06-27 14:41:00") & date_time <= as.POSIXct("2023-06-27 15:01:00")) ~ "CHL4",
           (date_time >= as.POSIXct("2023-06-27 15:02:00") & date_time <= as.POSIXct("2023-06-27 15:20:00")) ~ "CHL5",
           (date_time >= as.POSIXct("2023-06-27 15:22:00") & date_time <= as.POSIXct("2023-06-27 15:58:00")) ~ "CHL6",
           (date_time >= as.POSIXct("2023-06-27 15:59:00") & date_time <= as.POSIXct("2023-06-27 16:28:00")) ~ "CHL7",
           (date_time >= as.POSIXct("2023-06-27 16:29:00") & date_time <= as.POSIXct("2023-06-27 18:08:00")) ~ "CHL8",
           (date_time >= as.POSIXct("2023-06-27 18:10:00") & date_time <= as.POSIXct("2023-06-27 19:16:00")) ~ "CHL9",
           (date_time >= as.POSIXct("2023-06-27 19:17:00") & date_time <= as.POSIXct("2023-06-27 20:08:00")) ~ "CHL10",
           (date_time >= as.POSIXct("2023-06-28 7:52:00")  & date_time <= as.POSIXct("2023-06-28 8:35:00"))  ~ "CHL11",
           (date_time >= as.POSIXct("2023-06-28 8:37:00")  & date_time <= as.POSIXct("2023-06-28 9:38:00")) ~ "CHL12",
           (date_time >= as.POSIXct("2023-06-28 9:40:00")  & date_time <= as.POSIXct("2023-06-28 10:05:00")) ~ "CHL13",
           (date_time >= as.POSIXct("2023-06-28 10:07:00") & date_time <= as.POSIXct("2023-06-28 11:28:00")) ~ "CHL14",
           (date_time >= as.POSIXct("2023-06-28 11:29:00") & date_time <= as.POSIXct("2023-06-28 12:26:00")) ~ "CHL15",
           (date_time >= as.POSIXct("2023-06-28 12:28:00") & date_time <= as.POSIXct("2023-06-28 13:46:00")) ~ "CHL16",
           (date_time >= as.POSIXct("2023-06-28 13:47:00") & date_time <= as.POSIXct("2023-06-28 15:01:00")) ~ "CHL17",
           (date_time >= as.POSIXct("2023-06-28 15:02:00") & date_time <= as.POSIXct("2023-06-28 15:34:00")) ~ "CHL18",
           (date_time >= as.POSIXct("2023-06-28 15:35:00") & date_time <= as.POSIXct("2023-06-28 16:16:00")) ~ "CHL19",
           (date_time >= as.POSIXct("2023-06-28 16:17:00") & date_time <= as.POSIXct("2023-06-28 17:04:00")) ~ "CHL20",
           (date_time >= as.POSIXct("2023-06-28 17:05:00") & date_time <= as.POSIXct("2023-06-28 17:19:00")) ~ "CHL21",
           (date_time >= as.POSIXct("2023-06-28 17:20:00") & date_time <= as.POSIXct("2023-06-28 18:26:00")) ~ "CHL22",
           (date_time >= as.POSIXct("2023-06-28 18:28:00") & date_time <= as.POSIXct("2023-06-28 20:05:00")) ~ "CHL23",
           (date_time >= as.POSIXct("2023-06-28 20:10:00") & date_time <= as.POSIXct("2023-06-28 20:50:00")) ~ "CHL24",
           (date_time >= as.POSIXct("2023-06-28 20:52:00") & date_time <= as.POSIXct("2023-06-28 21:22:00")) ~ "CHL25",
           (date_time >= as.POSIXct("2023-06-28 21:23:00") & date_time <= as.POSIXct("2023-06-28 21:55:00")) ~ "CHL26",
           
           TRUE ~ NA_character_
         )) %>% 
  filter(!is.na(code)) %>% 
  mutate(
    date_time = as.character(date_time),
    date_time_parsed = parse_date_time(
      date_time,
      orders = c(
  "Ymd HMS",
  "Ymd HM",
  "dmy HMS",
  "dmy HM",
  "dmY HMS",
  "mdy HMS",
  "mdy HM"
))) %>% 
  mutate(year= year(date_time_parsed),
         month = month(date_time_parsed),
         mmyy = format(date_time_parsed, "%m-%y"))

  


```

+   2023 chl frequency distribution

```{r}


#No manual filter

ysi2023 %>% 
  mutate(code = as.factor(code)) %>% 
ggplot(aes(x = chl_rfu)) +
  geom_histogram(bins = 30, binwidth = 2, boundary = 0, fill = "steelblue", color = "black") +
  #facet_wrap(~ code, scales = "free_y") +
  labs(x = "Chlorophyll (RFU)", y = "Frequency") +
  theme_bw() +
  theme(text = element_text(size = 30, colour = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))+
  scale_x_continuous(
    breaks = seq(0, max(ysi2023$chl_rfu, na.rm = TRUE), by = 2)
  ) 

ggsave(file="plots/2023_plots/ysi2023_gen_freq_distribution.png", width = 10, height = 12, dpi = 300)




#Make plots for separate sites

ysi2023 %>% 
  mutate(code = as.factor(code)) %>% 
ggplot(aes(x = chl_rfu)) +
  geom_histogram(bins = 30, binwidth=2, boundary=0, fill = "steelblue", color = "black") +
  facet_wrap(~ code) +
  labs(x = "Chlorophyll (RFU)", y = "Frequency") +
   theme(text = element_text(size = 30, colour = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))+
  theme_bw() +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 4))



ggsave(file="plots/2023_plots/ysi2023_site_plots.png",width = 16, height = 12, dpi = 300)




```

+   2023 filter for outliers

```{r}

ysicrjul_23_outlier_detect <- multidetect(data = ysi2023, 
                                var = "chl_rfu", 
                                var_col = "code",
                                multiple = TRUE,
                                methods = c("adjbox", "iqr", "hampel","jknife",
                                            "seqfences", "mixediqr",
                                            "distboxplot", "semiqr",
                                            "zscore", "logboxplot", "medianrule"),
                       silence_true_errors = TRUE, 
                       missingness = 0.1,
                       sdm = FALSE,
                       na.inform = TRUE)
  
  

#Visualize the number of outliers detected by each method

ggoutliers(ysicrjul_23_outlier_detect, select = unique(ysi2023$code),
           color = "lightblue",
           ncol = 5,
           nrow = 7)



#outlier_per_site24 <- ggoutliers(ysicrjul_23_outlier_detect,
                       # select = c("CHL26", "CHL25", "CHL24","CHL3","CHL4", "CHL5","CHL6", #"CHL7", "BG10","CHL9","CHL10","BG13", "CHL19","CHL18","CHL15","CNL1","CNL2",  "CHL12", "CHL13","BG16","CHL20","CHL21","CHL27","BG5"),
                       # color='purple',
                       # desc = TRUE,
                       # ncol = 5,
                       # nrow = 7)

ggsave(file = "plots/2023_plots/2023_outliersoutlier_per_site23.png", width = 10, height = 20, dpi = 300)


#Obtaining quality controlled dataset using loess method or data labeling


out_qc_loess23 <- extract_clean_data(refdata = ysi2023, 
            outliers = ysicrjul_23_outlier_detect, loess = TRUE,
            var_col = "code",
            threshold = 0.7)  #set threshold to 0.8 to remove the very strong outliers only

#clean dataset
nrow(out_qc_loess23)

#reference data
nrow(ysi2023)

out_qc_labeled23 <- classify_data(refdata = ysi2023, 
                                 var_col = "code", 
                                outliers = ysicrjul_23_outlier_detect)

#Visualize labelled quality controlled dataset

ggenvironmentalspace(out_qc_labeled23, 
                     type = '1D',
                     ggxangle = 45, 
                     scalecolor = 'viridis',
                     xhjust = 1,
                     legend_position = 'blank',
                     ylab = "Number of records",
                     xlab = "Outlier labels",
                     ncol = 5,
                     nrow = 7)

ggsave(filename = "plots/2023_plots/labelled_outlier_chart_site.png", width = 10, height = 26, dpi = 300)


#labelled outlier or not 
labeled23 <- classify_data(refdata = ysi2023, 
                           var_col = "code",
                           outliers = ysicrjul_23_outlier_detect)



#plot clean dataset

out_qc_loess23 %>% 
mutate(code = as.factor(code)) %>% 
ggplot(aes(x = chl_rfu)) +
  geom_histogram(bins = 30, binwidth = 2, boundary = 0, fill = "steelblue", color = "black") +
facet_wrap(~ code, scales = "free_y")+
  labs(x = "Chlorophyll (RFU)", y = "Frequency") +
  theme_bw() +
  theme(text = element_text(size = 10, colour = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))+
 scale_x_continuous(
  breaks = scales::pretty_breaks(n = 6)  #ggplot makes breaks based on data
)

ggsave(file="plots/2023_plots/ysi2023_loessfilter_sites_freq_distribution.png", width = 10, height = 14, dpi = 300)



```


+   final filtering of points for cruz 2023

```{r}

final23_df <- out_qc_labeled23 %>%
  group_by(code) %>%
  slice_max(order_by = chl_rfu, n = 3) %>%
  summarise(
    chl_ysi = mean(chl_rfu, na.rm = TRUE),
    bga_ysi = mean(bga_rfu, na.rm = TRUE),
    temp_ysi = mean(temp, na.rm = TRUE),
    sp_cond_ysi = mean(sp_cond, na.rm = TRUE),
    file_name  = unique(file_name)
  )
    
    

```


+   **2024**
```{r}


crjul_24 <- df_ysi_cruz %>%
  filter(file_name == "CR24JUL_ysi")
  
##cross check to only keep all depths > -0.7 i.e. -0.6 to -0.4 for lake samples

##adding timestamps to cruise (CR) and Dunga mapping data (they have a different date_time structure from the calibration data)

ysicrjul_24 <- crjul_24 %>%
  filter(file_name != c("CALIB24_ysi")) %>% 
  mutate(date_time = as.POSIXct(date_time, format = "%m/%d/%Y %H:%M"),
         code = case_when(
            (date_time >= as.POSIXct("2024-07-15 15:40") & date_time <= as.POSIXct("2024-07-15 15:41")) ~ "DM1",
            (date_time >= as.POSIXct("2024-07-15 15:47") & date_time <= as.POSIXct("2024-07-15 15:49")) ~ "DM2",
                   (date_time >= as.POSIXct("2024-07-15 15:52") & date_time <= as.POSIXct("2024-07-15 15:55")) ~ "DM3",
           (date_time >= as.POSIXct("2024-07-15 15:58") & date_time <= as.POSIXct("2024-07-15 16:01")) ~ "DM4",
           (date_time >= as.POSIXct("2024-07-15 16:04") & date_time <= as.POSIXct("2024-07-15 16:07")) ~ "DM5",
           (date_time >= as.POSIXct("2024-07-15 16:13") & date_time <= as.POSIXct("2024-07-15 16:16")) ~ "DM6",
           (date_time >= as.POSIXct("2024-07-15 16:20") & date_time <= as.POSIXct("2024-07-15 16:23")) ~ "DM7",
           (date_time >= as.POSIXct("2024-07-15 16:27") & date_time <= as.POSIXct("2024-07-15 16:29")) ~ "DM8",
           (date_time >= as.POSIXct("2024-07-15 16:33") & date_time <= as.POSIXct("2024-07-15 16:35")) ~ "DM9",
           (date_time >= as.POSIXct("2024-07-15 16:38") & date_time <= as.POSIXct("2024-07-15 16:41")) ~ "DM10",
           (date_time >= as.POSIXct("2024-07-15 16:43") & date_time <= as.POSIXct("2024-07-15 16:46")) ~ "DM11",
           (date_time >= as.POSIXct("2024-07-15 16:50") & date_time <= as.POSIXct("2024-07-15 16:53")) ~ "DM12",
           (date_time >= as.POSIXct("2024-07-15 16:59") & date_time <= as.POSIXct("2024-07-15 17:01")) ~ "DM13",
           (date_time >= as.POSIXct("2024-07-15 17:07") & date_time <= as.POSIXct("2024-07-15 17:09")) ~ "DM14",
           (date_time >= as.POSIXct("2024-07-15 17:13") & date_time <= as.POSIXct("2024-07-15 17:14")) ~ "DM15",
           (date_time >= as.POSIXct("2024-07-15 17:17") & date_time <= as.POSIXct("2024-07-15 17:20")) ~ "DM16",
          (date_time >= as.POSIXct("2024-07-15 18:17")  & date_time <= as.POSIXct("2024-07-15 18:21")) ~ "PT",
          
          (date_time >= as.POSIXct("2024-07-18 14:28:00") & date_time <= as.POSIXct("2024-07-18 14:48:00")) ~ "CHL26",
           (date_time >= as.POSIXct("2024-07-18 15:37:00") & date_time <= as.POSIXct("2024-07-18 15:42:00")) ~ "CHL25",
           (date_time >= as.POSIXct("2024-07-18 17:14:00") & date_time <= as.POSIXct("2024-07-18 17:31:00")) ~ "CHL24",
           (date_time >= as.POSIXct("2024-07-18 18:58:00") & date_time <= as.POSIXct("2024-07-18 19:13:00")) ~ "CHL3",
           (date_time >= as.POSIXct("2024-07-18 19:36:00") & date_time <= as.POSIXct("2024-07-18 19:50:00")) ~ "CHL4",
           (date_time >= as.POSIXct("2024-07-18 20:16:00") & date_time <= as.POSIXct("2024-07-18 20:27:00")) ~ "CHL5",
           (date_time >= as.POSIXct("2024-07-18 20:48:00") & date_time <= as.POSIXct("2024-07-18 21:04:00")) ~ "CHL6",
           (date_time >= as.POSIXct("2024-07-18 21:51:00") & date_time <= as.POSIXct("2024-07-18 22:12:00")) ~ "CHL7",
           (date_time >= as.POSIXct("2024-07-19 7:48:00") & date_time <= as.POSIXct("2024-07-19 8:05:00")) ~ "BG10",
           (date_time >= as.POSIXct("2024-07-19 9:23:00") & date_time <= as.POSIXct("2024-07-19 9:31:00")) ~ "CHL9",
           (date_time >= as.POSIXct("2024-07-19 10:46:00") & date_time <= as.POSIXct("2024-07-19 11:01:00")) ~ "CHL10",
           (date_time >= as.POSIXct("2024-07-19 11:31:00") & date_time <= as.POSIXct("2024-07-19 11:56:00"))  ~ "BG13",
           (date_time >= as.POSIXct("2024-07-19 14:23:00") & date_time <= as.POSIXct("2024-07-19 14:48:00")) ~ "CHL19",
           (date_time >= as.POSIXct("2024-07-19 16:02:00") & date_time <= as.POSIXct("2024-07-19 16:14:00")) ~ "CHL18",
          (date_time >= as.POSIXct("2024-07-19 16:53:00") & date_time <= as.POSIXct("2024-07-19 16:59:00")) ~ "CHL15",
           (date_time >= as.POSIXct("2024-07-19 17:52:00") & date_time <= as.POSIXct("2024-07-19 18:09:00")) ~ "CNL1",
          (date_time >= as.POSIXct("2024-07-20 8:31:00") & date_time <= as.POSIXct("2024-07-20 8:48:00")) ~ "CNL2",
          (date_time >= as.POSIXct("2024-07-20 9:52:00") & date_time <= as.POSIXct("2024-07-20 10:05:00")) ~ "CHL12",
           (date_time >= as.POSIXct("2024-07-20 10:46:00") & date_time <= as.POSIXct("2024-07-20 10:57:00")) ~ "CHL13",
           (date_time >= as.POSIXct("2024-07-20 12:01:00") & date_time <= as.POSIXct("2024-07-20 12:14:00")) ~ "BG16",
           (date_time >= as.POSIXct("2024-07-20 14:29:00") & date_time <= as.POSIXct("2024-07-20 14:39:00")) ~ "CHL20",
           (date_time >= as.POSIXct("2024-07-20 15:34:00") & date_time <= as.POSIXct("2024-07-20 15:43:00")) ~ "CHL21",
           (date_time >= as.POSIXct("2024-07-20 18:18:00") & date_time <= as.POSIXct("2024-07-20 18:23:00")) ~ "CHL27",
          (date_time >= as.POSIXct("2024-07-20 19:06:00") & date_time <= as.POSIXct("2024-07-20 19:13:00")) ~ "BG5",
          
              TRUE ~ NA_character_
         )) %>% 
  filter(!is.na(code)) %>% 
  mutate(
    date_time = as.character(date_time),
    date_time_parsed = parse_date_time(
      date_time,
      orders = c(
  "Ymd HMS",
  "Ymd HM",
  "dmy HMS",
  "dmy HM",
  "dmY HMS",
  "mdy HMS",
  "mdy HM"
))) %>% 
  mutate(year= year(date_time_parsed),
         month = month(date_time_parsed),
         mmyy = format(date_time_parsed, "%m-%y"))



```


+     Frequency distribution for July 2024 cruz
```{r}



#No manual filter

ysicrjul_24 %>% 
  mutate(code = as.factor(code)) %>% 
ggplot(aes(x = chl_rfu)) +
  geom_histogram(bins = 30, binwidth = 2, boundary = 0, fill = "steelblue", color = "black") +
  #facet_wrap(~ code, scales = "free_y") +
  labs(x = "Chlorophyll (RFU)", y = "Frequency") +
  theme_bw() +
  theme(text = element_text(size = 30, colour = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))+
 scale_x_continuous(
  breaks = scales::pretty_breaks(n = 6)  #ggplot makes breaks based on data
)

ggsave(file="plots/2024_plots/ysi2024_gen_freq_distribution.png", width = 10, height = 12, dpi = 300)




#Make plots for separate sites

ysicrjul_24 %>% 
  mutate(code = as.factor(code)) %>% 
ggplot(aes(x = chl_rfu)) +
  geom_histogram(bins = 30, binwidth=2, boundary=0, fill = "steelblue", color = "black") +
  facet_wrap(~ code, scales = "free_y") +
  labs(x = "Chlorophyll (RFU)", y = "Frequency") +
   theme(text = element_text(size = 30, colour = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))+
  theme_bw() +
  scale_x_continuous(
  breaks = scales::pretty_breaks(n = 6)
)
  

ggsave(file="plots/2024_plots/ysi2024_site_freq_distribution.png",width = 16, height = 12, dpi = 300)


#this data might a manual filter to remove values chl >150






```

+  try outlier filters for July 2024 data

+   manual removal based on error readings
```{r}

ysicrjul_24_filter <- ysicrjul_24 %>% 
  filter(!(chl_rfu <= 0 | chl_rfu >= 100)) 

ysicrjul_24_filter %>% 
mutate(code = as.factor(code)) %>% 
ggplot(aes(x = chl_rfu)) +
  geom_histogram(bins = 30, binwidth = 2, boundary = 0, fill = "steelblue", color = "black") +
facet_wrap(~ code, scales = "free_y") +
  labs(x = "Chlorophyll (RFU)", y = "Frequency") +
  theme_bw() +
  theme(text = element_text(size = 10, colour = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))+
 scale_x_continuous(
  breaks = scales::pretty_breaks(n = 6)  #ggplot makes breaks based on data
)

ggsave(file="plots/2024_plots/ysi2024_manualfilter_sites_freq_distribution.png", width = 10, height = 12, dpi = 300)



```

+   outlier removal using specleanr

```{r}

ysicrjul_24_outlier_detect <- multidetect(data = ysicrjul_24_filter, 
                                var = "chl_rfu", 
                                var_col = "code",
                                multiple = TRUE,
                                methods = c("adjbox", "iqr", "hampel","jknife",
                                            "seqfences", "mixediqr",
                                            "distboxplot", "semiqr",
                                            "zscore", "logboxplot", "medianrule"),
                       silence_true_errors = TRUE, 
                       missingness = 0.1,
                       sdm = FALSE,
                       na.inform = TRUE)
  
  

#Visualize the number of outliers detected by each method

ggoutliers(ysicrjul_24_outlier_detect,
                        select = c("CHL26", "CHL25", "CHL24","CHL3","CHL4", "CHL5","CHL6", "CHL7", "BG10","CHL9","CHL10","BG13", "CHL19","CHL18","CHL15","CNL1","CNL2",  "CHL12", "CHL13","BG16","CHL20","CHL21","CHL27","BG5"),
                        color='purple',
                        desc = TRUE,
                        ncol = 5,
                        nrow = 7)

ggsave(file = "plots/2024_plots/2024_outliers_per_site24_no_chl18_.png", width = 10, height = 20, dpi = 300)


#Obtaining quality controlled dataset using loess method or data labeling


out_qc_loess24 <- extract_clean_data(refdata = ysicrjul_24_filter, 
            outliers = ysicrjul_24_outlier_detect, loess = TRUE,
            var_col = "code",
            threshold = 0.7)  #set threshold to 0.8 to remove the very strong outliers only

#clean dataset
nrow(out_qc_loess24)

#reference data
nrow(ysicrjul_24)

out_qc_labeled24 <- classify_data(refdata = ysicrjul_24_filter, 
                                  var_col = "code",
                                outliers = ysicrjul_24_outlier_detect)

#Visualize labelled quality controlled dataset

ggenvironmentalspace(out_qc_labeled24, 
                     type = '1D',
                     ggxangle = 45, 
                     ncol = 5,
                     nrow = 12,
                     scalecolor = 'viridis',
                     xhjust = 1,
                     legend_position = 'blank',
                     ylab = "Number of records",
                     xlab = "Outlier labels")


ggsave(filename = "plots/2024_plots/filtered_site_data.png", width = 14, height = 14, dpi = 300)



#labelled outlier or not 
labeled24 <- classify_data(refdata = ysicrjul_24_filter, 
                         var_col = "code",
                         outliers = ysicrjul_24_outlier_detect)



#plot clean dataset

out_qc_loess24 %>% 
mutate(code = as.factor(code)) %>% 
ggplot(aes(x = chl_rfu)) +
  geom_histogram(bins = 30, binwidth = 2, boundary = 0, fill = "steelblue", color = "black") +
facet_wrap(~ code, scales = "free_y")+
  labs(x = "Chlorophyll (RFU)", y = "Frequency") +
  theme_bw() +
  theme(text = element_text(size = 10, colour = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))+
 scale_x_continuous(
  breaks = scales::pretty_breaks(n = 6)  #ggplot makes breaks based on data
)

ggsave(file="plots/2024_plots/ysi2024_loessfilter_Sites_freq_distribution.png", width = 10, height = 12, dpi = 300)





```

+   final filtering of top points for ysi 
```{r}


final24_df <- out_qc_labeled24 %>%
  group_by(code, year) %>%
  slice_max(order_by = chl_rfu, n = 3) %>%
  summarise(
    chl_ysi = mean(chl_rfu, na.rm = TRUE),
    bga_ysi = mean(bga_rfu, na.rm = TRUE),
    temp_ysi = mean(temp, na.rm = TRUE),
    sp_cond_ysi = mean(sp_cond, na.rm = TRUE),
    file_name  = unique(file_name),
  ) 


```


+   Extracted chlorophyll files


+ 2022

```{r}


data_2022 <- read_csv(file_paths.csv[1])




data_2022 <- data_2022 %>% 
  rename(date_time = date) 

data_2022 <- data_2022%>% 
  mutate(date = date(date_time),
         time = paste(hour(date_time), minute(date_time), second(date_time), 
                      sep = ":"))  # Extract time components and combine)

cruise_sites_22 <- data_2022 %>% 
  dplyr::select(station, fu_chl) %>%
  rename(code = station,
         ext_chl = fu_chl) %>% 
  filter(code %in% c("BG10", "BG13", "BG16", "CNL1", "BG18")) 
  
#cruise_sites_22 contains extracted chl for matching the 2022 cruise_compar sondes

ysidata_2022 <- full_join(cruise_sondes_2022, cruise_sites_22, by = "code") %>% 
  rename(file_name = sample_type)



```


+   2023


```{r}


file_paths.csv <- dir("extracted_chls", pattern = "*.csv", full.names = TRUE)


chl_df_23 <- read_csv(file_paths.csv[2]) %>% 
  clean_names() %>% 
  rename(code =site)

cr_jul23_merge <- full_join(chl_df_23, final23_df)

cr_jul23_merge_subset <- cr_jul23_merge %>% 
  dplyr::select(code, ext_chl, chl_ysi, bga_ysi, sp_cond_ysi, temp_ysi, file_name)

```

+   2024
```{r}


chl_dat_24 <- read_csv(file_paths.csv[1]) %>% 
  clean_names() %>% 
  mutate(dilution_ratio = as.character(dilution_ratio)) %>% 
  rename(index_code = chl_index)


# Define the function to compute the dilution factor (DF)
df_func <- function(total_vol, sample_vol) {
  DF = total_vol / sample_vol
  return(DF)
}

# Apply the function to the defined columns to get DF
chl_dat$DF <- df_func(chl_dat$total_vol_in_cuvette, chl_dat$sample_vol)


#convert final_FU to chl concentration using Pete's formula (for formula refer to WG_2022 analysis R file)

#Chl a (ug/l) = 0.0003*F*dilution factor*extract volume (mL) / volume filtered (L)
#where dilution factor is whether you needed to dilute in order to get a reading.

###Make a function to convert florescence units to chlorophyll concentration
convert_fluorescence_to_chlorophyll <- function(fluoresence_units, dilution_factor, extract_volume, volume_filtered) {
 ext_chl <- 0.0003 * fluorescence_units * dilution_factor * extract_volume / volume_filtered
 return(ext_chl)
}


fluorescence_units <- chl_dat$fu_after_dilution
dilution_factor  <- chl_dat$DF
extract_volume <- chl_dat$vol_extract
volume_filtered <- chl_dat$vol_filtered/1000


chl_dat$ext_chl <- convert_fluorescence_to_chlorophyll(fluoresence_units, dilution_factor, extract_volume, volume_filtered)


chl_dat_subset <- chl_dat[-36, ]

chl_df_24 <- chl_dat_subset %>% 
  select(index_code,site, ext_chl, sample_type) %>% 
  filter_out(sample_type == "expt-jul_24") %>% 
  rename(code = site)
 


```


+   Join 2022, 2023 and 2024 cruz datasets

```{r}


#join ext_chl with july 2024 ysi data

cr_jul24_merge <- full_join(final24_df, chl_df_24, by = "code")
     

#join ext_chl 2023 cruz with july 2023 ysi data


cr_jul24_merge_subset <- cr_jul24_merge %>% 
  dplyr::select(code, ext_chl, chl_ysi, bga_ysi, sp_cond_ysi, temp_ysi, file_name)




ysi_field_data <- bind_rows(cr_jul23_merge_subset, cr_jul24_merge_subset, ysidata_2022)



```










#   **c3 files**
```{r}


#YSI files

# Get a list of file paths in the "files" directory

file_paths.csv <- dir("c3_files", pattern = "*.csv", full.names = TRUE)


# Read the first file and inspect its column names
first_file <- read_csv(file_paths.csv[1])
colnames(first_file)

#function to read and combine csv files
read_and_combine_csv <- function(file_path) {
  tryCatch(
    {
      # Read the CSV file
      data <- read_csv(file_path)
      
      
      # Remove the first row
      data <- data[-1, ]
      
      # Columns to convert to double
      cols_to_convert <- c("Temp", "SpCond", "Sal", "Depth", "Chl...6", "Chl...7", "BGA-PC...8", "BGA-PC...9", "Battery")
      
      # Convert specified columns to double
      data[, cols_to_convert] <- lapply(data[, cols_to_convert], as.numeric)
      
      # Extract the file name without the ".csv" extension
      file_name <- tools::file_path_sans_ext(basename(file_path))
      
      # Add a new column "FileName" to the data frame, remove formats
      data <- data %>% mutate(FileName = file_name)%>% 
        clean_names() #clean headings 
      
      
      return(data)
    },
    error = function(e) {
      cat("Error occurred while processing file:", file_path, "\n")
      message("Error Message:", conditionMessage(e), "\n")
      traceback()
      return(NULL) # Return NULL to indicate failure
    }
  )
}

# Read all csv files and combine them
 df_ysi_cruz <- map_df(file_paths.csv, read_and_combine_csv) %>% 
    rename(chl_rfu = chl_7,
         bga_rfu = bga_pc_9) 
 




```

+     2025 March-April

```{r}

crap25 <- file_paths.csv[4] %>% 
    rename(chl_rfu = chl_7,
         bga_rfu = bga_pc_9) 
 




fifth_file <- 

```

