---
title: "cruise_data"
author: "Angela Nankabirwa"
date: "2026-03-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


#loading packages

```{r}

library(tidyverse)
library(janitor)
library(stringi)  
library(lubridate)
library(readr)
library(knitr)
library(readxl)
library(ggpattern) 
library(gridExtra)
library(kableExtra) #formatting e,g. colour)
library(zoo) #remove noisy bumps from data #rolling median
library(specleanr) #20 ways to detect outliers
library(tidytext) #visualize outliers from specleanr package
library(ggpubr) ##adding equations to the plot
library(purrr) #to read csv files with map_dfr 
library(hms) #convert time to proper hms
library(mgcv) #For GAM models



```






#import all cruise ysi files
+   2022

+     MFANGANO SONDE
```{r}

file_paths.csv <- dir("2022_files", pattern = "*.csv", full.names = TRUE)



ysiF_2022 <- read_csv(file_paths.csv[2], skip=2, 
                      col_names = c("date_time", "temp", 
                                    "sp_cond", "salinity", "depth", "chl_ug_l",
                                    "chl_rfu_F", "bga_pc", "bga_rfu_F", "battery"))

#attach time stamps
ysiF_2022_b <- ysiF_2022 %>%
  mutate(date_time = as.POSIXct(date_time, format = "%m/%d/%Y %H:%M"),
         code = case_when(
           (date_time >= as.POSIXct("2022-07-12 8:20:00") & date_time <= as.POSIXct("2022-07-12 8:40:00")) ~ "BG10",
           (date_time >= as.POSIXct("2022-07-12 10:40:00") & date_time <= as.POSIXct("2022-07-12 11:00:00")) ~ "BG13",
           (date_time >= as.POSIXct("2022-07-12 12:50:00") & date_time <= as.POSIXct("2022-07-12 13:00:00")) ~ "BG16",
           (date_time >= as.POSIXct("2022-07-12 15:40:00") & date_time <= as.POSIXct("2022-07-12 15:50:00")) ~ "CNL1",
           (date_time >= as.POSIXct("2022-07-12 21:00:00") & date_time <= as.POSIXct("2022-07-12 21:20:00")) ~ "BG18",
           TRUE ~ NA_character_
         )) %>% 
  filter(!is.na(code))



```

+   checking for MF sonde outliers

```{r}
#The values are very few. They were recorded every 30 minutes and so we will just do average


ysiF_22_outlier_detect <- multidetect(data = ysiF_2022_b, 
                                var = "chl_rfu_F", 
                                var_col = "code",
                                multiple = TRUE,
                                methods = c("adjbox", "iqr", "hampel","jknife",
                                            "seqfences", "mixediqr",
                                            "distboxplot", "semiqr",
                                            "zscore", "logboxplot", "medianrule"),
                       silence_true_errors = TRUE, 
                       missingness = 0.1,
                       sdm = FALSE,
                       na.inform = TRUE)
  
  

#Visualize the number of outliers detected by each method

ggoutliers(ysiF_22_outlier_detect, select = unique(ysiF_2022_b$code),
           color = "lightblue",
           ncol = 5,
           nrow = 1)



ggsave(file = "plots/2022_plots/2022MF_outlier_per_site23.png", width = 10, height = 5, dpi = 300)


#Obtaining quality controlled dataset using loess method or data labeling


out_qc_loess22F <- extract_clean_data(refdata = ysiF_2022_b, 
            outliers = ysiF_22_outlier_detect, loess = TRUE,
            var_col = "code",
            threshold = 0.7)  #set threshold to 0.8 to remove the very strong outliers only

#clean dataset
nrow(out_qc_loess22F)

#reference data
nrow(ysiF_2022_b)

out_qc_labeled22F <- classify_data(refdata = ysiF_2022_b, 
                                 var_col = "code", 
                                outliers = ysiF_22_outlier_detect)

#Visualize labelled quality controlled dataset

ggenvironmentalspace(out_qc_labeled22F, 
                     type = '1D',
                     ggxangle = 45, 
                     scalecolor = 'viridis',
                     xhjust = 1,
                     legend_position = 'blank',
                     ylab = "Number of records",
                     xlab = "Outlier labels",
                     ncol = 5,
                     nrow = 1)

ggsave(filename = "plots/2022_plots/labelled_outlier_chart_MF.png", width = 8, height = 5, dpi = 300)


#labelled outlier or not 
labeled22F <- classify_data(refdata = ysiF_2022_b, 
                           var_col = "code",
                           outliers = ysiF_22_outlier_detect)



#plot clean dataset

out_qc_loess22F %>% 
mutate(code = as.factor(code)) %>% 
ggplot(aes(x = chl_rfu_F)) +
  geom_histogram(bins=5,binwidth = 1, boundary = 0, fill = "steelblue", color = "black") +
facet_wrap(~ code, scales = "free_y")+
  labs(x = "Chlorophyll (RFU)", y = "Frequency") +
  theme_bw() +
  theme(text = element_text(size = 20, colour = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))+
 scale_x_continuous(
  breaks = scales::pretty_breaks(n = 6)  #ggplot makes breaks based on data
)

ggsave(file="plots/2022_plots/ysi22_loessfilter_MF_freq_distribution.png", width = 10, height = 14, dpi = 300)



#No strong outlier so we will just average the values

ysiF_2022_c <- ysiF_2022_b %>% 
  filter(!is.na(code)) %>% 
  group_by(code) %>% 
  summarise(
    chl_ysi = mean(chl_rfu_F, na.rm = TRUE),   
    bga_ysi = mean(bga_rfu_F, na.rm = TRUE),
    temp_ysi = mean(temp, na.rm = TRUE),
    sp_cond_ysi = mean(sp_cond, na.rm = TRUE)
  ) %>% 
  ungroup() %>% 
  mutate(sample_type = "cruise-2022_MFsonde")

```


+   Dunga sonde 2022

```{r}
#DUNGA sonde
ysi_2022DG <-read_csv(file_paths.csv[1], skip=2, col_names = c("date_time", "temp", 
                                       "sp_cond", "salinity",
                                                           "depth", "chl_ug_l",
                                  "chl_rfu_DG", "bga_pc", "bga_rfu_DG", "battery"))


ysi_2022DG_b <- ysi_2022DG %>%
  mutate(date_time = as.POSIXct(date_time, format = "%m/%d/%Y %H:%M"),
         code = case_when(
           (date_time >= as.POSIXct("2022-07-12 8:20:00") & date_time <= as.POSIXct("2022-07-12 8:40:00")) ~ "BG10",
           (date_time >= as.POSIXct("2022-07-12 10:50:00") & date_time <= as.POSIXct("2022-07-12 11:00:00")) ~ "BG13",
           (date_time >= as.POSIXct("2022-07-12 12:50:00") & date_time <= as.POSIXct("2022-07-12 13:00:00")) ~ "BG16",
           (date_time >= as.POSIXct("2022-07-12 15:40:00") & date_time <= as.POSIXct("2022-07-12 15:50:00")) ~ "CNL1",
           (date_time >= as.POSIXct("2022-07-12 21:00:00") & date_time <= as.POSIXct("2022-07-12 21:20:00")) ~ "BG18",
           TRUE ~ NA_character_
         )) %>% 
  filter(!is.na(code))


#for outliers, same situation as MF sonde, so we will just do the average

ysi_2022DG_c <- ysi_2022DG_b %>% 
  filter(!is.na(code)) %>% 
group_by(code) %>% 
  summarise(
    chl_ysi = mean(chl_rfu_DG, na.rm = TRUE),
    bga_ysi = mean(bga_rfu_DG, na.rm= TRUE),
    temp_ysi = mean(temp, na.rm=TRUE),
    sp_cond_ysi = mean(sp_cond, na.rm = TRUE)
    ) %>% 
  ungroup() %>% 
  mutate(sample_type = "cruise-2022_DGysi")
```

+     combine MF and DG sonde data

```{r}
####bind DG and MF sonde cruise

cruise_sondes_2022 <- bind_rows(ysi_2022DG_c, ysiF_2022_c) 



```



+ import files for 2023 and 2024

```{r}


#YSI files

# Get a list of file paths in the "files" directory

file_paths.csv <- dir("ysi_files", pattern = "*.csv", full.names = TRUE)


# Read the first file and inspect its column names
first_file <- read_csv(file_paths.csv[1])
colnames(first_file)

#function to read and combine csv files
read_and_combine_csv <- function(file_path) {
  tryCatch(
    {
      # Read the CSV file
      data <- read_csv(file_path)
      
      
      # Remove the first row
      data <- data[-1, ]
      
      # Columns to convert to double
      cols_to_convert <- c("Temp", "SpCond", "Sal", "Depth", "Chl...6", "Chl...7", "BGA-PC...8", "BGA-PC...9", "Battery")
      
      # Convert specified columns to double
      data[, cols_to_convert] <- lapply(data[, cols_to_convert], as.numeric)
      
      # Extract the file name without the ".csv" extension
      file_name <- tools::file_path_sans_ext(basename(file_path))
      
      # Add a new column "FileName" to the data frame, remove formats
      data <- data %>% mutate(FileName = file_name)%>% 
        clean_names() #clean headings 
      
      
      return(data)
    },
    error = function(e) {
      cat("Error occurred while processing file:", file_path, "\n")
      message("Error Message:", conditionMessage(e), "\n")
      traceback()
      return(NULL) # Return NULL to indicate failure
    }
  )
}

# Read all csv files and combine them
 df_ysi_cruz <- map_df(file_paths.csv, read_and_combine_csv) %>% 
    rename(chl_rfu = chl_7,
         bga_rfu = bga_pc_9) 
 


```

+   attach time stamps to 2023 cruz

+   **2023**
```{r}



ysi2023 <- df_ysi_cruz %>%
  filter(file_name == "cruise-ysidata-2023_naphtali") %>% 
  
  mutate(date_time = as.POSIXct(date_time, format = "%m/%d/%Y %H:%M"),
         code = case_when(
           (date_time >= as.POSIXct("2023-06-27 11:26:00") & date_time <= as.POSIXct("2023-06-27 11:44:00")) ~ "CHL1",
           (date_time >= as.POSIXct("2023-06-27 11:46:00") & date_time <= as.POSIXct("2023-06-27 12:49:00")) ~ "CHL2",
           (date_time >= as.POSIXct("2023-06-27 12:50:00") & date_time <= as.POSIXct("2023-06-27 14:10:00")) ~ "BG7",
           (date_time >= as.POSIXct("2023-06-27 14:11:00") & date_time <= as.POSIXct("2023-06-27 14:40:00")) ~ "CHL3",
           (date_time >= as.POSIXct("2023-06-27 14:41:00") & date_time <= as.POSIXct("2023-06-27 15:01:00")) ~ "CHL4",
           (date_time >= as.POSIXct("2023-06-27 15:02:00") & date_time <= as.POSIXct("2023-06-27 15:20:00")) ~ "CHL5",
           (date_time >= as.POSIXct("2023-06-27 15:22:00") & date_time <= as.POSIXct("2023-06-27 15:58:00")) ~ "CHL6",
           (date_time >= as.POSIXct("2023-06-27 15:59:00") & date_time <= as.POSIXct("2023-06-27 16:28:00")) ~ "CHL7",
           (date_time >= as.POSIXct("2023-06-27 16:29:00") & date_time <= as.POSIXct("2023-06-27 18:08:00")) ~ "CHL8",
           (date_time >= as.POSIXct("2023-06-27 18:10:00") & date_time <= as.POSIXct("2023-06-27 19:16:00")) ~ "CHL9",
           (date_time >= as.POSIXct("2023-06-27 19:17:00") & date_time <= as.POSIXct("2023-06-27 20:08:00")) ~ "CHL10",
           (date_time >= as.POSIXct("2023-06-28 7:52:00")  & date_time <= as.POSIXct("2023-06-28 8:35:00"))  ~ "CHL11",
           (date_time >= as.POSIXct("2023-06-28 8:37:00")  & date_time <= as.POSIXct("2023-06-28 9:38:00")) ~ "CHL12",
           (date_time >= as.POSIXct("2023-06-28 9:40:00")  & date_time <= as.POSIXct("2023-06-28 10:05:00")) ~ "CHL13",
           (date_time >= as.POSIXct("2023-06-28 10:07:00") & date_time <= as.POSIXct("2023-06-28 11:28:00")) ~ "CHL14",
           (date_time >= as.POSIXct("2023-06-28 11:29:00") & date_time <= as.POSIXct("2023-06-28 12:26:00")) ~ "CHL15",
           (date_time >= as.POSIXct("2023-06-28 12:28:00") & date_time <= as.POSIXct("2023-06-28 13:46:00")) ~ "CHL16",
           (date_time >= as.POSIXct("2023-06-28 13:47:00") & date_time <= as.POSIXct("2023-06-28 15:01:00")) ~ "CHL17",
           (date_time >= as.POSIXct("2023-06-28 15:02:00") & date_time <= as.POSIXct("2023-06-28 15:34:00")) ~ "CHL18",
           (date_time >= as.POSIXct("2023-06-28 15:35:00") & date_time <= as.POSIXct("2023-06-28 16:16:00")) ~ "CHL19",
           (date_time >= as.POSIXct("2023-06-28 16:17:00") & date_time <= as.POSIXct("2023-06-28 17:04:00")) ~ "CHL20",
           (date_time >= as.POSIXct("2023-06-28 17:05:00") & date_time <= as.POSIXct("2023-06-28 17:19:00")) ~ "CHL21",
           (date_time >= as.POSIXct("2023-06-28 17:20:00") & date_time <= as.POSIXct("2023-06-28 18:26:00")) ~ "CHL22",
           (date_time >= as.POSIXct("2023-06-28 18:28:00") & date_time <= as.POSIXct("2023-06-28 20:05:00")) ~ "CHL23",
           (date_time >= as.POSIXct("2023-06-28 20:10:00") & date_time <= as.POSIXct("2023-06-28 20:50:00")) ~ "CHL24",
           (date_time >= as.POSIXct("2023-06-28 20:52:00") & date_time <= as.POSIXct("2023-06-28 21:22:00")) ~ "CHL25",
           (date_time >= as.POSIXct("2023-06-28 21:23:00") & date_time <= as.POSIXct("2023-06-28 21:55:00")) ~ "CHL26",
           
           TRUE ~ NA_character_
         )) %>% 
  filter(!is.na(code)) %>% 
  mutate(
    date_time = as.character(date_time),
    date_time_parsed = parse_date_time(
      date_time,
      orders = c(
  "Ymd HMS",
  "Ymd HM",
  "dmy HMS",
  "dmy HM",
  "dmY HMS",
  "mdy HMS",
  "mdy HM"
))) %>% 
  mutate(year= year(date_time_parsed),
         month = month(date_time_parsed),
         mmyy = format(date_time_parsed, "%m-%y"))

  


```

+   2023 chl frequency distribution

```{r}


#No manual filter

ysi2023 %>% 
  mutate(code = as.factor(code)) %>% 
ggplot(aes(x = chl_rfu)) +
  geom_histogram(bins = 30, binwidth = 2, boundary = 0, fill = "steelblue", color = "black") +
  #facet_wrap(~ code, scales = "free_y") +
  labs(x = "Chlorophyll (RFU)", y = "Frequency") +
  theme_bw() +
  theme(text = element_text(size = 30, colour = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))+
  scale_x_continuous(
    breaks = seq(0, max(ysi2023$chl_rfu, na.rm = TRUE), by = 2)
  ) 

ggsave(file="plots/2023_plots/ysi2023_gen_freq_distribution.png", width = 10, height = 12, dpi = 300)




#Make plots for separate sites

ysi2023 %>% 
  mutate(code = as.factor(code)) %>% 
ggplot(aes(x = chl_rfu)) +
  geom_histogram(bins = 30, binwidth=2, boundary=0, fill = "steelblue", color = "black") +
  facet_wrap(~ code) +
  labs(x = "Chlorophyll (RFU)", y = "Frequency") +
   theme(text = element_text(size = 30, colour = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))+
  theme_bw() +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 4))



ggsave(file="plots/2023_plots/ysi2023_site_plots.png",width = 16, height = 12, dpi = 300)




```

+   2023 filter for outliers

```{r}

ysicrjul_23_outlier_detect <- multidetect(data = ysi2023, 
                                var = "chl_rfu", 
                                var_col = "code",
                                multiple = TRUE,
                                methods = c("adjbox", "iqr", "hampel","jknife",
                                            "seqfences", "mixediqr",
                                            "distboxplot", "semiqr",
                                            "zscore", "logboxplot", "medianrule"),
                       silence_true_errors = TRUE, 
                       missingness = 0.1,
                       sdm = FALSE,
                       na.inform = TRUE)
  
  

#Visualize the number of outliers detected by each method

ggoutliers(ysicrjul_23_outlier_detect, select = unique(ysi2023$code),
           color = "lightblue",
           ncol = 5,
           nrow = 7)



#outlier_per_site24 <- ggoutliers(ysicrjul_23_outlier_detect,
                       # select = c("CHL26", "CHL25", "CHL24","CHL3","CHL4", "CHL5","CHL6", #"CHL7", "BG10","CHL9","CHL10","BG13", "CHL19","CHL18","CHL15","CNL1","CNL2",  "CHL12", "CHL13","BG16","CHL20","CHL21","CHL27","BG5"),
                       # color='purple',
                       # desc = TRUE,
                       # ncol = 5,
                       # nrow = 7)

ggsave(file = "plots/2023_plots/2023_outliersoutlier_per_site23.png", width = 10, height = 20, dpi = 300)


#Obtaining quality controlled dataset using loess method or data labeling


out_qc_loess23 <- extract_clean_data(refdata = ysi2023, 
            outliers = ysicrjul_23_outlier_detect, loess = TRUE,
            var_col = "code",
            threshold = 0.7)  #set threshold to 0.8 to remove the very strong outliers only

#clean dataset
nrow(out_qc_loess23)

#reference data
nrow(ysi2023)

out_qc_labeled23 <- classify_data(refdata = ysi2023, 
                                 var_col = "code", 
                                outliers = ysicrjul_23_outlier_detect)

#Visualize labelled quality controlled dataset

ggenvironmentalspace(out_qc_labeled23, 
                     type = '1D',
                     ggxangle = 45, 
                     scalecolor = 'viridis',
                     xhjust = 1,
                     legend_position = 'blank',
                     ylab = "Number of records",
                     xlab = "Outlier labels",
                     ncol = 5,
                     nrow = 7)

ggsave(filename = "plots/2023_plots/labelled_outlier_chart_site.png", width = 10, height = 26, dpi = 300)


#labelled outlier or not 
labeled23 <- classify_data(refdata = ysi2023, 
                           var_col = "code",
                           outliers = ysicrjul_23_outlier_detect)



#plot clean dataset

out_qc_loess23 %>% 
mutate(code = as.factor(code)) %>% 
ggplot(aes(x = chl_rfu)) +
  geom_histogram(bins = 30, binwidth = 2, boundary = 0, fill = "steelblue", color = "black") +
facet_wrap(~ code, scales = "free_y")+
  labs(x = "Chlorophyll (RFU)", y = "Frequency") +
  theme_bw() +
  theme(text = element_text(size = 10, colour = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))+
 scale_x_continuous(
  breaks = scales::pretty_breaks(n = 6)  #ggplot makes breaks based on data
)

ggsave(file="plots/2023_plots/ysi2023_loessfilter_sites_freq_distribution.png", width = 10, height = 14, dpi = 300)



```


+   final filtering of points for cruz 2023

```{r}

final23_df <- out_qc_labeled23 %>%
  group_by(code) %>%
  slice_max(order_by = chl_rfu, n = 3) %>%
  summarise(
    chl_ysi = mean(chl_rfu, na.rm = TRUE),
    bga_ysi = mean(bga_rfu, na.rm = TRUE),
    temp_ysi = mean(temp, na.rm = TRUE),
    sp_cond_ysi = mean(sp_cond, na.rm = TRUE),
    file_name  = unique(file_name)
  )
    
    

```


+   **2024**
```{r}


crjul_24 <- df_ysi_cruz %>%
  filter(file_name == "CR24JUL_ysi")
  
##cross check to only keep all depths > -0.7 i.e. -0.6 to -0.4 for lake samples

##adding timestamps to cruise (CR) and Dunga mapping data (they have a different date_time structure from the calibration data)

ysicrjul_24 <- crjul_24 %>%
  filter(file_name != c("CALIB24_ysi")) %>% 
  mutate(date_time = as.POSIXct(date_time, format = "%m/%d/%Y %H:%M"),
         code = case_when(
          
          (date_time >= as.POSIXct("2024-07-18 14:28:00") & date_time <= as.POSIXct("2024-07-18 14:48:00")) ~ "CHL26",
           (date_time >= as.POSIXct("2024-07-18 15:37:00") & date_time <= as.POSIXct("2024-07-18 15:42:00")) ~ "CHL25",
           (date_time >= as.POSIXct("2024-07-18 17:14:00") & date_time <= as.POSIXct("2024-07-18 17:31:00")) ~ "CHL24",
           (date_time >= as.POSIXct("2024-07-18 18:58:00") & date_time <= as.POSIXct("2024-07-18 19:13:00")) ~ "CHL3",
           (date_time >= as.POSIXct("2024-07-18 19:36:00") & date_time <= as.POSIXct("2024-07-18 19:50:00")) ~ "CHL4",
           (date_time >= as.POSIXct("2024-07-18 20:16:00") & date_time <= as.POSIXct("2024-07-18 20:27:00")) ~ "CHL5",
           (date_time >= as.POSIXct("2024-07-18 20:48:00") & date_time <= as.POSIXct("2024-07-18 21:04:00")) ~ "CHL6",
           (date_time >= as.POSIXct("2024-07-18 21:51:00") & date_time <= as.POSIXct("2024-07-18 22:12:00")) ~ "CHL7",
           (date_time >= as.POSIXct("2024-07-19 7:48:00") & date_time <= as.POSIXct("2024-07-19 8:05:00")) ~ "BG10",
           (date_time >= as.POSIXct("2024-07-19 9:23:00") & date_time <= as.POSIXct("2024-07-19 9:31:00")) ~ "CHL9",
           (date_time >= as.POSIXct("2024-07-19 10:46:00") & date_time <= as.POSIXct("2024-07-19 11:01:00")) ~ "CHL10",
           (date_time >= as.POSIXct("2024-07-19 11:31:00") & date_time <= as.POSIXct("2024-07-19 11:56:00"))  ~ "BG13",
           (date_time >= as.POSIXct("2024-07-19 14:23:00") & date_time <= as.POSIXct("2024-07-19 14:48:00")) ~ "CHL19",
           (date_time >= as.POSIXct("2024-07-19 16:02:00") & date_time <= as.POSIXct("2024-07-19 16:14:00")) ~ "CHL18",
          (date_time >= as.POSIXct("2024-07-19 16:53:00") & date_time <= as.POSIXct("2024-07-19 16:59:00")) ~ "CHL15",
           (date_time >= as.POSIXct("2024-07-19 17:52:00") & date_time <= as.POSIXct("2024-07-19 18:09:00")) ~ "CNL1",
          (date_time >= as.POSIXct("2024-07-20 8:31:00") & date_time <= as.POSIXct("2024-07-20 8:48:00")) ~ "CNL2",
          (date_time >= as.POSIXct("2024-07-20 9:52:00") & date_time <= as.POSIXct("2024-07-20 10:05:00")) ~ "CHL12",
           (date_time >= as.POSIXct("2024-07-20 10:46:00") & date_time <= as.POSIXct("2024-07-20 10:57:00")) ~ "CHL13",
           (date_time >= as.POSIXct("2024-07-20 12:01:00") & date_time <= as.POSIXct("2024-07-20 12:14:00")) ~ "BG16",
           (date_time >= as.POSIXct("2024-07-20 14:29:00") & date_time <= as.POSIXct("2024-07-20 14:39:00")) ~ "CHL20",
           (date_time >= as.POSIXct("2024-07-20 15:34:00") & date_time <= as.POSIXct("2024-07-20 15:43:00")) ~ "CHL21",
           (date_time >= as.POSIXct("2024-07-20 18:18:00") & date_time <= as.POSIXct("2024-07-20 18:23:00")) ~ "CHL27",
          (date_time >= as.POSIXct("2024-07-20 19:06:00") & date_time <= as.POSIXct("2024-07-20 19:13:00")) ~ "BG5",
          
              TRUE ~ NA_character_
         )) %>% 
  filter(!is.na(code)) %>% 
  mutate(
    date_time = as.character(date_time),
    date_time_parsed = parse_date_time(
      date_time,
      orders = c(
  "Ymd HMS",
  "Ymd HM",
  "dmy HMS",
  "dmy HM",
  "dmY HMS",
  "mdy HMS",
  "mdy HM"
))) %>% 
  mutate(year= year(date_time_parsed),
         month = month(date_time_parsed),
         mmyy = format(date_time_parsed, "%m-%y")) 



```


+     Frequency distribution for July 2024 cruz
```{r}



#No manual filter

ysicrjul_24 %>% 
  mutate(code = as.factor(code)) %>% 
ggplot(aes(x = chl_rfu)) +
  geom_histogram(bins = 30, binwidth = 2, boundary = 0, fill = "steelblue", color = "black") +
  #facet_wrap(~ code, scales = "free_y") +
  labs(x = "Chlorophyll (RFU)", y = "Frequency") +
  theme_bw() +
  theme(text = element_text(size = 30, colour = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))+
 scale_x_continuous(
  breaks = scales::pretty_breaks(n = 6)  #ggplot makes breaks based on data
)

ggsave(file="plots/2024_plots/ysi2024_gen_freq_distribution.png", width = 10, height = 12, dpi = 300)




#Make plots for separate sites

ysicrjul_24 %>% 
  mutate(code = as.factor(code)) %>% 
ggplot(aes(x = chl_rfu)) +
  geom_histogram(bins = 30, binwidth=2, boundary=0, fill = "steelblue", color = "black") +
  facet_wrap(~ code, scales = "free_y") +
  labs(x = "Chlorophyll (RFU)", y = "Frequency") +
   theme(text = element_text(size = 30, colour = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))+
  theme_bw() +
  scale_x_continuous(
  breaks = scales::pretty_breaks(n = 6)
)
  

ggsave(file="plots/2024_plots/ysi2024_site_freq_distribution.png",width = 16, height = 12, dpi = 300)


#this data might a manual filter to remove values chl >150






```

+  try outlier filters for July 2024 data

+   manual removal based on error readings
```{r}

ysicrjul_24_filter <- ysicrjul_24 %>% 
  filter(!(chl_rfu <= 0 | chl_rfu >= 100)) 

ysicrjul_24_filter %>% 
mutate(code = as.factor(code)) %>% 
ggplot(aes(x = chl_rfu)) +
  geom_histogram(bins = 30, binwidth = 2, boundary = 0, fill = "steelblue", color = "black") +
facet_wrap(~ code, scales = "free_y") +
  labs(x = "Chlorophyll (RFU)", y = "Frequency") +
  theme_bw() +
  theme(text = element_text(size = 10, colour = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))+
 scale_x_continuous(
  breaks = scales::pretty_breaks(n = 6)  #ggplot makes breaks based on data
)

ggsave(file="plots/2024_plots/ysi2024_manualfilter_sites_freq_distribution.png", width = 10, height = 12, dpi = 300)



```

+   outlier removal using specleanr

```{r}

ysicrjul_24_outlier_detect <- multidetect(data = ysicrjul_24_filter, 
                                var = "chl_rfu", 
                                var_col = "code",
                                multiple = TRUE,
                                methods = c("adjbox", "iqr", "hampel","jknife",
                                            "seqfences", "mixediqr",
                                            "distboxplot", "semiqr",
                                            "zscore", "logboxplot", "medianrule"),
                       silence_true_errors = TRUE, 
                       missingness = 0.1,
                       sdm = FALSE,
                       na.inform = TRUE)
  
  

#Visualize the number of outliers detected by each method

ggoutliers(ysicrjul_24_outlier_detect,
                        select = c("CHL26", "CHL25", "CHL24","CHL3","CHL4", "CHL5","CHL6", "CHL7", "BG10","CHL9","CHL10","BG13", "CHL19","CHL18","CHL15","CNL1","CNL2",  "CHL12", "CHL13","BG16","CHL20","CHL21","CHL27","BG5"),
                        color='purple',
                        desc = TRUE,
                        ncol = 5,
                        nrow = 7)

ggsave(file = "plots/2024_plots/2024_outliers_per_site24_no_chl18_.png", width = 10, height = 20, dpi = 300)


#Obtaining quality controlled dataset using loess method or data labeling


out_qc_loess24 <- extract_clean_data(refdata = ysicrjul_24_filter, 
            outliers = ysicrjul_24_outlier_detect, loess = TRUE,
            var_col = "code",
            threshold = 0.7)  #set threshold to 0.8 to remove the very strong outliers only

#clean dataset
nrow(out_qc_loess24)

#reference data
nrow(ysicrjul_24)

out_qc_labeled24 <- classify_data(refdata = ysicrjul_24_filter, 
                                  var_col = "code",
                                outliers = ysicrjul_24_outlier_detect)

#Visualize labelled quality controlled dataset

ggenvironmentalspace(out_qc_labeled24, 
                     type = '1D',
                     ggxangle = 45, 
                     ncol = 5,
                     nrow = 12,
                     scalecolor = 'viridis',
                     xhjust = 1,
                     legend_position = 'blank',
                     ylab = "Number of records",
                     xlab = "Outlier labels")


ggsave(filename = "plots/2024_plots/filtered_site_data.png", width = 14, height = 14, dpi = 300)



#labelled outlier or not 
labeled24 <- classify_data(refdata = ysicrjul_24_filter, 
                         var_col = "code",
                         outliers = ysicrjul_24_outlier_detect)



#plot clean dataset

out_qc_loess24 %>% 
mutate(code = as.factor(code)) %>% 
ggplot(aes(x = chl_rfu)) +
  geom_histogram(bins = 30, binwidth = 2, boundary = 0, fill = "steelblue", color = "black") +
facet_wrap(~ code, scales = "free_y")+
  labs(x = "Chlorophyll (RFU)", y = "Frequency") +
  theme_bw() +
  theme(text = element_text(size = 10, colour = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))+
 scale_x_continuous(
  breaks = scales::pretty_breaks(n = 6)  #ggplot makes breaks based on data
)

ggsave(file="plots/2024_plots/ysi2024_loessfilter_Sites_freq_distribution.png", width = 10, height = 12, dpi = 300)





```

+   final filtering of top points for ysi 
```{r}


final24_df <- out_qc_labeled24 %>%
  group_by(code, year) %>%
  slice_max(order_by = chl_rfu, n = 3) %>%
  summarise(
    chl_ysi = mean(chl_rfu, na.rm = TRUE),
    bga_ysi = mean(bga_rfu, na.rm = TRUE),
    temp_ysi = mean(temp, na.rm = TRUE),
    sp_cond_ysi = mean(sp_cond, na.rm = TRUE),
    file_name  = unique(file_name),
  ) 


```


+   Extracted chlorophyll files


+   2023


```{r}


file_paths.csv <- dir("extracted_chls", pattern = "*.csv", full.names = TRUE)


chl_df_23 <- read_csv(file_paths.csv[2]) %>% 
  clean_names() %>% 
  rename(code =site)

cr_jul23_merge <- full_join(chl_df_23, final23_df)

cr_jul23_merge_subset <- cr_jul23_merge %>% 
  dplyr::select(code, ext_chl, chl_ysi, bga_ysi, sp_cond_ysi, temp_ysi, file_name)

```

+   2024
```{r}


chl_dat_24 <- read_csv(file_paths.csv[1]) %>% 
  clean_names() %>% 
  mutate(dilution_ratio = as.character(dilution_ratio)) %>% 
  rename(index_code = chl_index)


# Define the function to compute the dilution factor (DF)
df_func <- function(total_vol, sample_vol) {
  DF = total_vol / sample_vol
  return(DF)
}

# Apply the function to the defined columns to get DF
chl_dat$DF <- df_func(chl_dat$total_vol_in_cuvette, chl_dat$sample_vol)


#convert final_FU to chl concentration using Pete's formula (for formula refer to WG_2022 analysis R file)

#Chl a (ug/l) = 0.0003*F*dilution factor*extract volume (mL) / volume filtered (L)
#where dilution factor is whether you needed to dilute in order to get a reading.

###Make a function to convert florescence units to chlorophyll concentration
convert_fluorescence_to_chlorophyll <- function(fluoresence_units, dilution_factor, extract_volume, volume_filtered) {
 ext_chl <- 0.0003 * fluorescence_units * dilution_factor * extract_volume / volume_filtered
 return(ext_chl)
}


fluorescence_units <- chl_dat$fu_after_dilution
dilution_factor  <- chl_dat$DF
extract_volume <- chl_dat$vol_extract
volume_filtered <- chl_dat$vol_filtered/1000


chl_dat$ext_chl <- convert_fluorescence_to_chlorophyll(fluoresence_units, dilution_factor, extract_volume, volume_filtered)


chl_dat_subset <- chl_dat[-36, ]

chl_df_24 <- chl_dat_subset %>% 
  select(index_code,site, ext_chl, sample_type) %>% 
  filter_out(sample_type == "expt-jul_24") %>% 
  rename(code = site)
 


```


+ 2022

```{r}


data_2022 <- read_csv(file_paths.csv[4])

#this file has 2022 cruise ext_chl


data_2022 <- data_2022 %>% 
  rename(date_time = date) 

data_2022 <- data_2022%>% 
  mutate(date = date(date_time),
         time = paste(hour(date_time), minute(date_time), second(date_time), 
                      sep = ":"))  # Extract time components and combine)

cruise_sites_22 <- data_2022 %>% 
  dplyr::select(station, fu_chl) %>%
  rename(code = station,
         ext_chl = fu_chl) %>% 
  filter(code %in% c("BG10", "BG13", "BG16", "CNL1", "BG18")) 
  
#cruise_sites_22 contains extracted chl for matching the 2022 cruise_compar sondes

ysidata_2022 <- full_join(cruise_sondes_2022, cruise_sites_22, by = "code") %>% 
  rename(file_name = sample_type)



```


+   Join 2022, 2023 and 2024 cruz datasets

```{r}


#join ext_chl with july 2024 ysi data

cr_jul24_merge <- full_join(final24_df, chl_df_24, by = "code")
     

#join ext_chl 2023 cruz with july 2023 ysi data


cr_jul24_merge_subset <- cr_jul24_merge %>% 
  dplyr::select(code, ext_chl, chl_ysi, bga_ysi, sp_cond_ysi, temp_ysi, file_name)




ysi_field_data <- bind_rows(cr_jul23_merge_subset, cr_jul24_merge_subset, ysidata_2022)  %>% 
  mutate(data_type = "field")



```


# Experiments data

+   2022 experiments

+   MF sonde
```{r}
#MF sonde

file_paths_csv <- dir("ysi_experiments", pattern = "*.csv", full.names = TRUE)


ysi_expt_files <- purrr::map_dfr(file_paths_csv, read_and_combine_csv) 



MF_EXPER <- ysi_expt_files %>% 
  filter(file_name == "MF_EXPER_2022") 

MF_EXPER_b <- MF_EXPER %>%
  mutate(date_time = as.POSIXct(date_time, format = "%m/%d/%Y %H:%M"),
         code = case_when(
           (date_time >= as.POSIXct("2022-07-10 12:01:00") & date_time <= as.POSIXct("2022-07-10 13:01:00")) ~ "X",
           (date_time >= as.POSIXct("2022-07-10 13:11:00") & date_time <= as.POSIXct("2022-07-10 14:21:00")) ~ "0.5X",
           (date_time >= as.POSIXct("2022-07-10 14:31:00") & date_time <= as.POSIXct("2022-07-10 15:21:00")) ~ "0.25X",
           (date_time >= as.POSIXct("2022-07-10 15:31:00") & date_time <= as.POSIXct("2022-07-10 16:21:00")) ~ "0.125X",
           (date_time >= as.POSIXct("2022-07-10 16:31:00") & date_time <= as.POSIXct("2022-07-10 21:01:00")) ~ "0.062X",
           (date_time >= as.POSIXct("2022-07-10 21:11:00") & date_time <= as.POSIXct("2022-07-10 22:41:00")) ~ "0.031X",
           (date_time >= as.POSIXct("2022-07-10 22:51:00") & date_time <= as.POSIXct("2022-07-11 06:01:00")) ~ "0.015X",
           (date_time >= as.POSIXct("2022-07-11 06:11:00") & date_time <= as.POSIXct("2022-07-11 12:01:00")) ~ "0.007X",
           TRUE ~ NA_character_
         )) %>% 
  filter(!is.na(code))



```

+     remove outliers_MF sonde
```{r}

ysiF_expt_outlier_detect <- multidetect(data = MF_EXPER_b, 
                                var = "chl_rfu_F", 
                                var_col = "code",
                                multiple = TRUE,
                                methods = c("adjbox", "iqr", "hampel","jknife",
                                            "seqfences", "mixediqr",
                                            "distboxplot", "semiqr",
                                            "zscore", "logboxplot", "medianrule"),
                       silence_true_errors = TRUE, 
                       missingness = 0.1,
                       sdm = FALSE,
                       na.inform = TRUE)
  
  

#Visualize the number of outliers detected by each method

ggoutliers(ysiF_expt_outlier_detect, select = unique(MF_EXPER_b$code),
           color = "lightblue",
           ncol = 5,
           nrow = 10)



ggsave(file = "plots/2022_plots/expts/2022MFsonde_Expt_outlier_graph.png", width = 10, height = 5, dpi = 300)


#Obtaining quality controlled dataset using loess method or data labeling


out_qc_loess22F <- extract_clean_data(refdata = MF_EXPER_b, 
            outliers = ysiF_expt_outlier_detect, loess = TRUE,
            var_col = "code",
            threshold = 0.7)  #set threshold to 0.8 to remove the very strong outliers only

#clean dataset
nrow(MF_EXPER_b)

#reference data
nrow(ysiF_2022_b)

out_qc_labeled22F <- classify_data(refdata = MF_EXPER_b, 
                                 var_col = "code", 
                                outliers = ysiF_expt_outlier_detect)

#Visualize labelled quality controlled dataset

ggenvironmentalspace(ysiF_expt_outlier_detect, 
                     type = '1D',
                     ggxangle = 45, 
                     scalecolor = 'viridis',
                     xhjust = 1,
                     legend_position = 'blank',
                     ylab = "Number of records",
                     xlab = "Outlier labels",
                     ncol = 5,
                     nrow = 2)  #investigate why it's not working#############################

#ggsave(filename = "plots/2022_plots/labelled_outlier_chart_MF.png", width = 8, height = 5, dpi = 300)


#labelled outlier or not 
labeled22F <- classify_data(refdata = MF_EXPER_b, 
                           var_col = "code",
                           outliers = ysiF_expt_outlier_detect)



#plot clean dataset

out_qc_loess22F %>% 
mutate(code = as.factor(code)) %>% 
ggplot(aes(x = chl_rfu_F)) +
  geom_histogram(bins=5,binwidth = 1, boundary = 0, fill = "steelblue", color = "black") +
#facet_wrap(~ code, scales = "free_y")+
  labs(x = "Chlorophyll (RFU)", y = "Frequency") +
  theme_bw() +
  theme(text = element_text(size = 20, colour = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))+
 scale_x_continuous(
  breaks = scales::pretty_breaks(n = 6)  #ggplot makes breaks based on data
)

ggsave(file="plots/2022_plots/expts/MF_expt_loessfilter_general_freq_distribution.png", width = 10, height = 14, dpi = 300)


#final dataset

MF_EXPER_c <- out_qc_loess22F %>% filter(!is.na(code)) %>% 
  group_by(code) %>%
  slice_max(order_by = chl_rfu_F, n = 3) %>%
  summarise(
    chl_ysi = mean(chl_rfu_F, na.rm = TRUE),
    bga_ysi = mean(bga_rfu_F, na.rm= TRUE),
    temp_ysi = mean(temp, na.rm=TRUE),
    sp_cond_ysi = mean(sp_cond, na.rm = TRUE)
    ) %>% 
  ungroup() %>% 
  mutate(file_name = "MFysi-2022_expt")

```


+     DG sonde
```{r}


DU_EXPER <- ysi_expt_files %>% 
  filter(file_name == "DG_EXPER_2022") 


DU_EXPER_b <- DU_EXPER %>%
  mutate(date_time = as.POSIXct(date_time, format = "%m/%d/%Y %H:%M"),
         code = case_when(
           (date_time >= as.POSIXct("2022-07-08 19:10:00") & date_time <= as.POSIXct("2022-07-10 13:00:00")) ~ "X",
           (date_time >= as.POSIXct("2022-07-10 13:10:00") & date_time <= as.POSIXct("2022-07-10 14:20:00")) ~ "0.5X",
           (date_time >= as.POSIXct("2022-07-10 14:30:00") & date_time <= as.POSIXct("2022-07-10 15:20:00")) ~ "0.25X",
           (date_time >= as.POSIXct("2022-07-10 15:30:00") & date_time <= as.POSIXct("2022-07-10 16:20:00")) ~ "0.125X",
           (date_time >= as.POSIXct("2022-07-10 16:30:00") & date_time <= as.POSIXct("2022-07-10 21:00:00")) ~ "0.062X",
           (date_time >= as.POSIXct("2022-07-10 21:10:00") & date_time <= as.POSIXct("2022-07-10 22:40:00")) ~ "0.031X",
           (date_time >= as.POSIXct("2022-07-10 22:50:00") & date_time <= as.POSIXct("2022-07-11 06:00:00")) ~ "0.015X",
           (date_time >= as.POSIXct("2022-07-11 06:10:00") & date_time <= as.POSIXct("2022-07-11 11:50:00")) ~ "0.007X",
           TRUE ~ NA_character_
         ))
```


+ remove outliers from the DG sonde 
```{r}


DG22_expt_outlier_detect <- multidetect(data = DU_EXPER_b, 
                                var = "chl_rfu_DG", 
                                var_col = "code",
                                multiple = TRUE,
                                methods = c("adjbox", "iqr", "hampel","jknife",
                                            "seqfences", "mixediqr",
                                            "distboxplot", "semiqr",
                                            "zscore", "logboxplot", "medianrule"),
                       silence_true_errors = TRUE, 
                       missingness = 0.1,
                       sdm = FALSE,
                       na.inform = TRUE)
  
  

#Visualize the number of outliers detected by each method

ggoutliers(DG22_expt_outlier_detect, select = unique(DU_EXPER_b$code),
           color = "lightblue",
           ncol = 5,
           nrow = 10)



ggsave(file = "plots/2022_plots/expts/2022MFsonde_Expt_outlier_graph.png", width = 10, height = 5, dpi = 300)


#Obtaining quality controlled dataset using loess method or data labeling


out_qc_loess22DG <- extract_clean_data(refdata = DU_EXPER_b, 
            outliers = DG22_expt_outlier_detect, loess = TRUE,
            var_col = "code",
            threshold = 0.8)  #set threshold to 0.8 to remove the very strong outliers only

#clean dataset
nrow(out_qc_loess22DG)

#reference data
nrow(DU_EXPER_b)

out_qc_labeled22DU <- classify_data(refdata = DU_EXPER_b, 
                                 var_col = "code", 
                                outliers = DG22_expt_outlier_detect)

#Visualize labelled quality controlled dataset

ggenvironmentalspace(DG22_expt_outlier_detect, 
                     type = '1D',
                     ggxangle = 45, 
                     scalecolor = 'viridis',
                     xhjust = 1,
                     legend_position = 'blank',
                     ylab = "Number of records",
                     xlab = "Outlier labels",
                     ncol = 5,
                     nrow = 3)  #investigate why it's not working#############################

#ggsave(filename = "plots/2022_plots/labelled_outlier_chart_MF.png", width = 8, height = 5, dpi = 300)


#labelled outlier or not 
labeled22DG <- classify_data(refdata = DU_EXPER_b, 
                           var_col = "code",
                           outliers = DG22_expt_outlier_detect)



#plot clean dataset

out_qc_labeled22DU %>% 
mutate(code = as.factor(code)) %>% 
ggplot(aes(x = chl_rfu_DG)) +
  geom_histogram(bins=5,binwidth = 1, boundary = 0, fill = "steelblue", color = "black") +
#facet_wrap(~ code, scales = "free_y")+
  labs(x = "Chlorophyll (RFU)", y = "Frequency") +
  theme_bw() +
  theme(text = element_text(size = 20, colour = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))+
 scale_x_continuous(
  breaks = scales::pretty_breaks(n = 6)  #ggplot makes breaks based on data
)

ggsave(file="plots/2022_plots/expts/DG_expt_loessfilter_general_freq_distribution.png", width = 10, height = 14, dpi = 300)


##################################################333
#Final dataset

DU_EXPER_c <- out_qc_labeled22DU %>% 
  group_by(code) %>%
  slice_max(order_by = chl_rfu_DG, n = 3) %>%
  summarise(
    chl_ysi = mean(chl_rfu_DG, na.rm = TRUE),
    bga_ysi = mean(bga_rfu_DG, na.rm= TRUE),
    temp_ysi = mean(temp, na.rm=TRUE),
    sp_cond_ysi = mean(sp_cond, na.rm = TRUE)
    ) %>% 
  ungroup() %>% 
  mutate(file_name = "DGysi-2022_expt")

###joining the expt readings for ysi_DG and ysi_MF
expt22_ysi <- full_join(DU_EXPER_c, MF_EXPER_c)




```


+ 2024 spiking MF sonde sensor and c3 experiments with ext_chl

This file was generated from a previous RMD file.

```{r}


exptsjul_24 <- read_csv("2024_spiking_expts_combined_data.csv") 


MF_EXPERjul_24 <- expts_24 %>% 
  filter(sos == "ysi") %>% 
  mutate(code = site,
         chl_ysi = top_chl,
         bga_ysi = top_bga,
         file_name = "2024_MFsonde_experiment",
         data_type = "experiment") %>% 
  dplyr::select(code, chl_ysi, bga_ysi, ext_chl, file_name, data_type)







```




+   extracted chl for the dilution expt

```{r}

ext_chl_expt22 <- read_csv(file_paths.csv[4])


ext_chl_expt22$DF <- df_func(ext_chl_expt22$total_vol, ext_chl_expt22$sample_vol)

fluorescence_units <- ext_chl_expt22$FU
dilution_factor  <- ext_chl_expt22$DF
extract_volume <- ext_chl_expt22$vol_ext_ml
volume_filtered <- ext_chl_expt22$vol_fil_ml/1000

ext_chl_expt22$ext_chl <- convert_fluorescence_to_chlorophyll(fluoresence_units, dilution_factor, extract_volume, volume_filtered)


ext_chl_expt22_b <- ext_chl_expt22 %>% 
  dplyr::select(code, ext_chl)


#join the ext_chl with the ysi readings

ysi_dilution_expt <- full_join(expt22_ysi, ext_chl_expt22_b, by = "code") %>% 
  mutate(data_type = "experiment") %>% 
  bind_rows(MF_EXPERjul_24)






```



+ join ysi dilution expt data with field data

```{r}


ysi_calib_data <- bind_rows(ysi_field_data, ysi_dilution_expt) %>% 
  filter(if_all(c(ext_chl, chl_ysi), ~ !is.na(.)))  
  






```





+ Linear regression for chl_ysi ~ ext_chl

+   check slopes across years

```{r}
#Field only
ysi_calib_data_field <- calib_data  %>% 
  filter(data_type %in% "field") 


ysi_calib_data %>% 
  filter(data_type %in% "field") %>% 
  arrange(file_name, chl_ysi) %>% 
  ggplot(aes(x = chl_ysi, y = ext_chl)) +
  geom_point(color = "steelblue") +
  geom_line(aes(group = 1)) +
  facet_wrap(~ file_name, scales= "free_y") 

ggsave(filename = "plots/CALIB/rship_ysi~ext_field.png", width = 10, height = 14, dpi = 300)


##log transformed

ysi_calib_data %>% 
  #filter(data_type %in% "experiment") %>% 
  ggplot(aes(log10(chl_ysi), log10(ext_chl), color = file_name)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  facet_wrap(~ file_name, scales = "free_y")

ggsave(filename = "plots/CALIB/log_rship_ysi~ext_field.png", width = 10, height = 14, dpi = 300)



###lm for different field data

###let's try a lm with the DGsonde included

ysi_calib_data_field_withdg  <- ysi_calib_data_field %>% 
arrange(file_name, chl_ysi) 

calib_data_field_withdg %>% 
ggplot(aes(log10(chl_ysi), log10(ext_chl), colour = file_name)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  theme(axis.text = element_text(size=14))+
stat_regline_equation(
    aes(label =  paste(after_stat(eq.label), ..rr.label.., 
                       sep = "~~~~")), show.legend = FALSE)


ggsave(filename = "plots/CALIB/log_rship_ysi~ext_field_withDGysi_facet.png", width = 10, height = 10, dpi = 300)


###model with DGsonde no interaction

model_1 <- lm(
  log10(ext_chl + 1) ~log10(chl_ysi + 1),
  data = ysi_calib_data_field_withdg
)

summary(model_1)



#no DGsonde
ysi_calib_data_field_nodg <- ysi_calib_data_field %>% 
filter(file_name != "cruise-2022_DGysi") %>% 
arrange(file_name, chl_ysi)

ysi_calib_data_field_nodg %>% 
ggplot(aes(log10(chl_ysi), log10(ext_chl), colour = file_name)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
stat_regline_equation(
    aes(label =  paste(after_stat(eq.label), ..rr.label.., 
                       sep = "~~~~")), show.legend = FALSE)

ggsave(filename = "plots/CALIB/log_rship_ysi~ext_field_noDGysi_facetted.png", width = 10, height = 10, dpi = 300)

###model with no DGsonde no interaction

model_1 <- lm(
  log10(ext_chl + 1) ~log10(chl_ysi + 1),
  data = ysi_calib_data_field_nodg
)

summary(model_1)



```

+ lets try models with interaction

```{r}

###model with DG with interaction

model_1 <- lm(
  log10(ext_chl + 1) ~log10(chl_ysi + 1)  * file_name,
  data = ysi_calib_data_field_withdg
)

summary(model_1)



###model with no DG with interaction

model_1 <- lm(
  log10(ext_chl + 1) ~log10(chl_ysi + 1)  * file_name,
  data = calib_data_field_nodg
)


summary(model_1)


```

+   let's add some environmental variables
```{r}

#with DG

model_1 <- lm(
  log10(ext_chl + 1) ~ 
    log10(chl_ysi + 1) * file_name +
    temp_ysi + sp_cond_ysi,
  data = calib_data_field_withdg
)

summary(model_1)

#no DG

model_1 <- lm(
  log10(ext_chl + 1) ~ 
    log10(chl_ysi + 1) * file_name +
    temp_ysi + sp_cond_ysi,
  data = calib_data_field_nodg
)

summary(model_1)



```







#   **c3 files**

+   2023 cruise
```{r}

file_paths_csv <- dir("c3_files", pattern = "*.csv", full.names = TRUE)


#function to read files

read_c3_file <- function(file) {
  
  readr::read_csv(
    file,
    skip = 5,
    col_names = c("date_time", "PC", "chl", "depth", "temp")
  ) %>%
    dplyr::slice(4:n()) %>%
    dplyr::select(date_time, PC, chl, depth, temp) %>%
    
    # Force numeric conversion 
    dplyr::mutate(
      dplyr::across(
        c(PC, chl, depth, temp),
        ~ as.numeric(.)    #readr::parse_number alternaitve code
      )
    ) %>%
    
    dplyr::mutate(source_file = basename(file))
}


#read files

c3_cruise_files <- purrr::map_dfr(file_paths_csv, read_c3_file) 

c3_cruise_files_b <- c3_cruise_files  %>% 
  mutate(
    date_time = as.character(date_time),
    date_time_parsed = parse_date_time(
      date_time,
      orders = c(
  "Ymd HMS",
  "Ymd HM",
  "dmy HMS",
  "dmy HM",
  "dmY HMS",
  "mdy HMS",
  "mdy HM"
))) %>%
  mutate(year= year(date_time_parsed),
         month = month(date_time_parsed),
         mmyy = format(date_time_parsed, "%m-%y")) %>% 
  arrange(date_time) %>% 
  clean_names()


#check files
table(c3_cruise_files_b$source_file)

```


+   attach timestamps to different files

+   27 June 2023 cruz
```{r}

c327jun_2023 <- c3_cruise_files_b %>%
  filter(source_file %in% "c3_CR_27_June_2023.csv") %>% 
  mutate(date_time = as.POSIXct(date_time, format = "%m/%d/%y %H:%M:%S"),
         code = case_when(
           (date_time >= as.POSIXct("2023-06-27 11:26:00") & date_time <= as.POSIXct("2023-06-27 11:44:00")) ~ "CHL1",
           (date_time >= as.POSIXct("2023-06-27 11:53:00") & date_time <= as.POSIXct("2023-06-27 12:49:00")) ~ "CHL2",
           (date_time >= as.POSIXct("2023-06-27 12:51:00") & date_time <= as.POSIXct("2023-06-27 14:08:00")) ~ "BG7",
           (date_time >= as.POSIXct("2023-06-27 14:26:00") & date_time <= as.POSIXct("2023-06-27 14:40:00")) ~ "CHL3",
           (date_time >= as.POSIXct("2023-06-27 14:42:00") & date_time <= as.POSIXct("2023-06-27 14:59:00")) ~ "CHL4",
           (date_time >= as.POSIXct("2023-06-27 15:02:00") & date_time <= as.POSIXct("2023-06-27 15:18:00")) ~ "CHL5",
           (date_time >= as.POSIXct("2023-06-27 15:22:00") & date_time <= as.POSIXct("2023-06-27 15:58:00")) ~ "CHL6",
           (date_time >= as.POSIXct("2023-06-27 15:59:00") & date_time <= as.POSIXct("2023-06-27 16:28:00")) ~ "CHL7",
           (date_time >= as.POSIXct("2023-06-27 16:29:00") & date_time <= as.POSIXct("2023-06-27 18:08:00")) ~ "CHL8",
           (date_time >= as.POSIXct("2023-06-27 18:10:00") & date_time <= as.POSIXct("2023-06-27 19:14:00")) ~ "CHL9",
           (date_time >= as.POSIXct("2023-06-27 19:17:00") & date_time <= as.POSIXct("2023-06-27 20:08:00")) ~ "CHL10",
  
           TRUE ~ NA_character_
         ))

c327jun_2023_clean <- c327jun_2023 %>% 
  filter(!is.na(code)) %>% 
  arrange(date_time)


```

+ 28 June 2023 data
```{r}



c328jun_2023 <- c3_cruise_files_b %>%
  filter(source_file %in% "Turner_C3_28June2023_cruise.csv") %>% 
  arrange(date_time) %>% 
  mutate(date_time = as.POSIXct(date_time, format = "%m/%d/%y %H:%M:%S"),
         code = case_when(
          (date_time >= as.POSIXct("2023-06-28 7:52:00")  & date_time <= as.POSIXct("2023-06-28 8:35:00"))  ~ "CHL11",
          (date_time >= as.POSIXct("2023-06-28 8:37:00")  & date_time <= as.POSIXct("2023-06-28 9:36:00")) ~ "CHL12",
          (date_time >= as.POSIXct("2023-06-28 9:40:00")  & date_time <= as.POSIXct("2023-06-28 10:05:00")) ~ "CHL13",
         (date_time >= as.POSIXct("2023-06-28 10:09:00") & date_time <= as.POSIXct("2023-06-28 11:12:00")) ~ "CHL14",
           (date_time >= as.POSIXct("2023-06-28 11:29:00") & date_time <= as.POSIXct("2023-06-28 12:24:00")) ~ "CHL15",
        (date_time >= as.POSIXct("2023-06-28 12:28:00") & date_time <= as.POSIXct("2023-06-28 13:46:00")) ~ "CHL16",
           (date_time >= as.POSIXct("2023-06-28 13:47:00") & date_time <= as.POSIXct("2023-06-28 14:58:00")) ~ "CHL17",
           (date_time >= as.POSIXct("2023-06-28 15:02:00") & date_time <= as.POSIXct("2023-06-28 15:34:00")) ~ "CHL18",
           (date_time >= as.POSIXct("2023-06-28 15:36:00") & date_time <= as.POSIXct("2023-06-28 16:16:00")) ~ "CHL19",
           (date_time >= as.POSIXct("2023-06-28 16:19:00") & date_time <= as.POSIXct("2023-06-28 17:04:00")) ~ "CHL20",
           (date_time >= as.POSIXct("2023-06-28 17:05:00") & date_time <= as.POSIXct("2023-06-28 17:19:00")) ~ "CHL21",
           (date_time >= as.POSIXct("2023-06-28 17:22:00") & date_time <= as.POSIXct("2023-06-28 18:26:00")) ~ "CHL22",   
           (date_time >= as.POSIXct("2023-06-28 18:28:00") & date_time <= as.POSIXct("2023-06-28 20:05:00")) ~ "CHL23",
            (date_time >= as.POSIXct("2023-06-28 20:10:00") & date_time <= as.POSIXct("2023-06-28 20:30:00")) ~ "CHL24",
           (date_time >= as.POSIXct("2023-06-28 20:52:00") & date_time <= as.POSIXct("2023-06-28 21:15:00")) ~ "CHL25",
           (date_time >= as.POSIXct("2023-06-28 21:23:00") & date_time <= as.POSIXct("2023-06-28 21:55:00")) ~ "CHL26",
           TRUE ~ NA_character_
         ))


 

```



+   Join 27 and 28 June cruise data
```{r}


 c3_2023_clean <- bind_rows(c327jun_2023,c328jun_2023)  %>% 
    filter(!is.na(code))

#write.csv(c3_2023_clean, file="c3_2023_clean.csv") 



```

+ Frequency distribution for c3 cuise dataset


```{r}


c3_2023_clean %>% 
  filter(chl > 0) %>% 
  mutate(code = as.factor(code)) %>% 
ggplot(aes(x = chl)) +
  geom_histogram(bins = 30, binwidth = 2, boundary = 0, fill = "steelblue", color = "black") +
  #facet_wrap(~ code) +
  labs(x = "Chlorophyll (RFU)", y = "Frequency") +
  theme_bw() +
  theme(text = element_text(size = 15, colour = "black"),
        axis.text.x = element_text(size=20, angle = 45, hjust = 1, vjust = 1))+
 scale_x_continuous(
  breaks = scales::pretty_breaks(n = 6)  #ggplot makes breaks based on data
)


ggsave(file="plots/c3/2023_plots/c3_2023cruz_gen_freq_distribution.png", width = 10, height = 12, dpi = 300)







```




+   Remove outliers from 2023 cruise c3 data

```{r}

c3_23_outlier_detect <- multidetect(data = c3_2023_clean, 
                                var = "chl", 
                                var_col = "code",
                                multiple = TRUE,
                                methods = c("adjbox", "iqr", "hampel","jknife",
                                            "seqfences", "mixediqr",
                                            "distboxplot", "semiqr",
                                            "zscore", "logboxplot", "medianrule"),
                       silence_true_errors = TRUE, 
                       missingness = 0.1,
                       sdm = FALSE,
                       na.inform = TRUE)
  
  

#Visualize the number of outliers detected by each method

ggoutliers(c3_23_outlier_detect,
                        select = unique(c3_2023_clean$code),
                        color='purple',
                        desc = TRUE,
                        ncol = 10,
                        nrow = 10)

ggsave(file = "plots/c3/2023_plots/outliers_per_site.png", width = 10, height = 10, dpi = 300)


#Obtaining quality controlled dataset using loess method or data labeling


out_qc_loess23 <- extract_clean_data(refdata = c3_2023_clean, 
            outliers = c3_23_outlier_detect, loess = TRUE,
            var_col = "code"
            )  #set threshold to 0.8 to remove the very strong outliers only

#clean dataset
nrow(out_qc_loess23)

#reference data
nrow(c3_2023_clean)

out_qc_labeled23 <- classify_data(refdata = c3_2023_clean, 
                                  var_col = "code",
                                outliers = c3_23_outlier_detect)


#Visualize labelled quality controlled dataset

ggenvironmentalspace(out_qc_labeled23, 
                     type = '1D',
                     ggxangle = 45, 
                     ncol = 10,
                     nrow = 15,
                     scalecolor = 'viridis',
                     xhjust = 1,
                     legend_position = 'blank',
                     ylab = "Number of records",
                     xlab = "Outlier labels")


ggsave(filename = "plots/c3/2023_plots/filtered_site_data.png", width = 14, height = 10, dpi = 300)



#labelled outlier or not 
labeled23 <- classify_data(refdata = c3_2023_clean, 
                         var_col = "code",
                         outliers = c3_23_outlier_detect)



#plot clean dataset

out_qc_loess23 %>% 
mutate(code = as.factor(code)) %>% 
ggplot(aes(x = chl)) +
  geom_histogram(bins = 30, binwidth = 2, boundary = 0, fill = "steelblue", color = "black") +
facet_wrap(~ code, scales = "free_y")+
  labs(x = "Chlorophyll (ug/l)", y = "Frequency") +
  theme_bw() +
  theme(text = element_text(size = 10, colour = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))+
 scale_x_continuous(
  breaks = scales::pretty_breaks(n = 6)  #ggplot makes breaks based on data
)

ggsave(file="plots/2023_plots/JUN2023_loessfilter_Sites_freq_distribution.png", width = 12, height = 10, dpi = 300)








```

The low value of chl at CHL22 is 97ug/l. It is not removed but still is high enough compared to many other sites... The threshold for this cruise is flexible

Chl at chl4 is also somewhat variable but the values are over all high 

+ Final filtering of points for Jun 2023 cruise

```{r}

final_crjun_2023_df <- out_qc_loess23 %>%
  group_by(code, year) %>%
  slice_max(order_by = chl, n = 3) %>%
  summarise(
    chl = mean(chl, na.rm = TRUE),
    pc = mean(pc, na.rm = TRUE),
    temp = mean(temp, na.rm = TRUE),
    depth = mean(depth, na.rm = TRUE),
    source_file  = unique(source_file),
  ) 




```







+     July 2024

```{r}

c3_jul2024 <- c3_cruise_files_b %>%
  filter(source_file %in% "c3_CR_24JUL2024.csv") %>% 
  mutate(date_time = as.POSIXct(date_time, format = "%m/%d/%y %H:%M:%S"),
         code = case_when(
        (date_time >= as.POSIXct("2024-07-18 14:29:00") & date_time <= as.POSIXct("2024-07-18 15:00:00")) ~ "CHL26",
        (date_time >= as.POSIXct("2024-07-18 15:37:00") & date_time <= as.POSIXct("2024-07-18 15:46:00")) ~ "CHL25",
        (date_time >= as.POSIXct("2024-07-18 17:12:00") & date_time <= as.POSIXct("2024-07-18 17:31:00")) ~ "CHL24",
        (date_time >= as.POSIXct("2024-07-18 18:57:00") & date_time <= as.POSIXct("2024-07-18 19:15:00")) ~ "CHL3",
           (date_time >= as.POSIXct("2024-07-18 19:37:00") & date_time <= as.POSIXct("2024-07-18 19:50:00")) ~ "CHL4",
           (date_time >= as.POSIXct("2024-07-18 20:15:00") & date_time <= as.POSIXct("2024-07-18 20:30:00")) ~ "CHL5",
           (date_time >= as.POSIXct("2024-07-18 20:48:00") & date_time <= as.POSIXct("2024-07-18 21:05:00")) ~ "CHL6",
           (date_time >= as.POSIXct("2024-07-18 21:51:00") & date_time <= as.POSIXct("2024-07-18 22:15:00")) ~ "CHL7",
        
           (date_time >= as.POSIXct("2024-07-19 07:48:00") & date_time <= as.POSIXct("2024-07-19 08:09:00")) ~ "BG10",
           (date_time >= as.POSIXct("2024-07-19 09:27:00") & date_time <= as.POSIXct("2024-07-19 09:50:00")) ~ "CHL9",
          (date_time >= as.POSIXct("2024-07-19 10:45:00") & date_time <= as.POSIXct("2024-07-19 11:02:00")) ~ "CHL10",
        (date_time >= as.POSIXct("2024-07-19 11:45:00")  & date_time <= as.POSIXct("2024-07-19 11:57:00"))  ~ "BG13",
        (date_time >= as.POSIXct("2024-07-19 14:20:00")  & date_time <= as.POSIXct("2024-07-19 14:40:00")) ~ "CHL19",
        (date_time >= as.POSIXct("2024-07-19 16:02:00")  & date_time <= as.POSIXct("2024-07-19 16:13:00")) ~ "CHL18",
        (date_time >= as.POSIXct("2024-07-19 16:52:00") & date_time <= as.POSIXct("2024-07-19 17:13:00")) ~ "CHL15",
        (date_time >= as.POSIXct("2024-07-19 17:50:00") & date_time <= as.POSIXct("2024-07-19 18:10:00")) ~ "CNL1",
        
        (date_time >= as.POSIXct("2024-07-20 08:31:00") & date_time <= as.POSIXct("2024-07-20 08:51:00")) ~ "CNL2",
        (date_time >= as.POSIXct("2024-07-20 09:52:00") & date_time <= as.POSIXct("2024-07-20 10:06:00")) ~ "CHL12",
        (date_time >= as.POSIXct("2024-07-20 10:46:00") & date_time <= as.POSIXct("2024-07-20 11:00:00")) ~ "CHL13",
        (date_time >= as.POSIXct("2024-07-20 12:01:00") & date_time <= as.POSIXct("2024-07-20 12:15:00")) ~ "BG16",
        (date_time >= as.POSIXct("2024-07-20 14:29:00") & date_time <= as.POSIXct("2024-07-20 14:39:00")) ~ "CHL20",
        (date_time >= as.POSIXct("2024-07-20 15:35:00") & date_time <= as.POSIXct("2024-07-20 15:41:00")) ~ "CHL21",
        (date_time >= as.POSIXct("2024-07-20 18:16:00") & date_time <= as.POSIXct("2024-07-20 18:25:00")) ~ "CHL27",
        (date_time >= as.POSIXct("2024-07-20 19:00:00") & date_time <= as.POSIXct("2024-07-20 19:15:00")) ~ "BG5",
           
           
           TRUE ~ NA_character_
         ))

c3_jul2024_clean <- c3_jul2024 %>% 
  filter(!is.na(code))





```





+   Frequency distribution for c3 2024 July cruise dataset



```{r}

#Keeping chl values > 0

c3_jul2024_clean %>% 
  filter(chl > 0) %>% 
  mutate(code = as.factor(code)) %>% 
ggplot(aes(x = chl)) +
  geom_histogram(bins = 30, binwidth = 2, boundary = 0, fill = "steelblue", color = "black") +
  #facet_wrap(~ code) +
  labs(x = "Chlorophyll (RFU)", y = "Frequency") +
  theme_bw() +
  theme(text = element_text(size = 15, colour = "black"),
        axis.text.x = element_text(size=14, angle = 45, hjust = 1, vjust = 1))+
 scale_x_continuous(
  breaks = scales::pretty_breaks(n = 6)  #ggplot makes breaks based on data
)


ggsave(file="plots/c3/2024_plots/c3_2024JULcruz_site_freq_distribution.png", width = 10, height = 12, dpi = 300)








```

+   Remove outliers from 2024 july cruise c3 data

```{r}


c3_j24_outlier_detect <- multidetect(data = c3_jul2024_clean, 
                                var = "chl", 
                                var_col = "code",
                                multiple = TRUE,
                                methods = c("adjbox", "iqr", "hampel","jknife",
                                            "seqfences", "mixediqr",
                                            "distboxplot", "semiqr",
                                            "zscore", "logboxplot", "medianrule"),
                       silence_true_errors = TRUE, 
                       missingness = 0.1,
                       sdm = FALSE,
                       na.inform = TRUE)
  
  

#Visualize the number of outliers detected by each method

ggoutliers(c3_j24_outlier_detect,
                        select = unique(c3_jul2024_clean$code),
                        color='purple',
                        desc = TRUE,
                        ncol = 5,
                        nrow = 10)

ggsave(file = "plots/c3/2024_plots/outliers_per_site.png", width = 10, height = 20, dpi = 300)


#Obtaining quality controlled dataset using loess method or data labeling


out_qc_loessj24 <- extract_clean_data(refdata = c3_jul2024_clean, 
            outliers = c3_j24_outlier_detect, loess = TRUE,
            var_col = "code"
            )  #set threshold to 0.8 to remove the very strong outliers only

#clean dataset
nrow(out_qc_loessj24)

#reference data
nrow(c3_jul2024_clean)

out_qc_labeled24 <- classify_data(refdata = c3_jul2024_clean, 
                                  var_col = "code",
                                outliers = c3_j24_outlier_detect)


#Visualize labelled quality controlled dataset

ggenvironmentalspace(out_qc_labeled24, 
                     type = '1D',
                     ggxangle = 45, 
                     ncol = 5,
                     nrow = 15,
                     scalecolor = 'viridis',
                     xhjust = 1,
                     legend_position = 'blank',
                     ylab = "Number of records",
                     xlab = "Outlier labels")


ggsave(filename = "plots/c3/2024_plots/filtered_site_data.png", width = 14, height = 14, dpi = 300)



#labelled outlier or not 
labeled24 <- classify_data(refdata = c3_jul2024_clean, 
                         var_col = "code",
                         outliers = c3_j24_outlier_detect)



#plot clean dataset

out_qc_loessj24 %>% 
mutate(code = as.factor(code)) %>% 
ggplot(aes(x = chl)) +
  geom_histogram(bins = 30, binwidth = 2, boundary = 0, fill = "steelblue", color = "black") +
facet_wrap(~ code)+
  labs(x = "Chlorophyll (ug/l)", y = "Frequency") +
  theme_bw() +
  theme(text = element_text(size = 10, colour = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))+
 scale_x_continuous(
  breaks = scales::pretty_breaks(n = 6)  #ggplot makes breaks based on data
)

ggsave(file="plots/2024_plots/c3_2024_loessfilter_Sites_freq_distribution.png", width = 10, height = 12, dpi = 300)






```

Notes: CHL21 is problematic... high variation



+ Final July cruz 2024 dataset

```{r}


final_crjul_2024_df <- out_qc_loessj24 %>%
  group_by(code, year) %>%
  summarise(
    chl = mean(chl, na.rm = TRUE),
    pc = mean(pc, na.rm = TRUE),
    temp = mean(temp, na.rm = TRUE),
    depth = mean(depth, na.rm = TRUE),
    source_file  = unique(source_file),
  ) 



final_crjul_2024_df2 <- out_qc_loessj24 %>%
  group_by(code, year) %>%
  slice_max(order_by = chl, n = 3) %>%
  summarise(
    chl = mean(chl, na.rm = TRUE),
    pc = mean(pc, na.rm = TRUE),
    temp = mean(temp, na.rm = TRUE),
    depth = mean(depth, na.rm = TRUE),
    source_file  = unique(source_file),
  ) 





```

For final dataset, I tried the top average vs averaging all cleaned values. The former over estimates chl in gulf areas that are more turbid while the latter averages better to provide a better comparison across neigbouring sites. So, I will just do the average of the cleaned values after removing outliers.




+     Nov 2024

```{r}

c3_nov2024 <- c3_cruise_files_b %>%
  filter(source_file %in% "c3_CR_NOV2024.csv") %>% 
  mutate(date_time = as.POSIXct(date_time, format = "%m/%d/%Y %H:%M"),
         code = case_when(
        (date_time >= as.POSIXct("2024-11-07 09:35:00") & date_time <= as.POSIXct("2024-11-07 09:43:00")) ~ "CHL26",
        (date_time >= as.POSIXct("2024-11-07 11:06:00") & date_time <= as.POSIXct("2024-11-07 11:35:00")) ~ "CHL25",
        (date_time >= as.POSIXct("2024-11-07 12:27:00") & date_time <= as.POSIXct("2024-11-07 12:55:00")) ~ "CHL24",
        (date_time >= as.POSIXct("2024-11-07 14:46:00") & date_time <= as.POSIXct("2024-11-07 14:56:00")) ~ "BG5",
        (date_time >= as.POSIXct("2024-11-07 16:28:00") & date_time <= as.POSIXct("2024-11-07 16:48:00")) ~ "CHL27",
        (date_time >= as.POSIXct("2024-11-07 17:40:00") & date_time <= as.POSIXct("2024-11-07 18:04:00")) ~ "CHL3",
        (date_time >= as.POSIXct("2024-11-07 18:30:00") & date_time <= as.POSIXct("2024-11-07 18:56:00")) ~ "CHL4",
        (date_time >= as.POSIXct("2024-11-07 19:24:00") & date_time <= as.POSIXct("2024-11-07 19:40:00")) ~ "CHL5",
        (date_time >= as.POSIXct("2024-11-07 20:05:00") & date_time <= as.POSIXct("2024-11-07 20:30:00")) ~ "CHL6",
        
           (date_time >= as.POSIXct("2024-11-08 08:06:00") & date_time <= as.POSIXct("2024-11-08 08:24:00")) ~ "BG10",
           
        (date_time >= as.POSIXct("2024-11-08 09:29:00")  & date_time <= as.POSIXct("2024-11-08 09:50:00")) ~ "CHL9",
        (date_time >= as.POSIXct("2024-11-08 10:07:00")  & date_time <= as.POSIXct("2024-11-08 10:26:00"))  ~ "CHL7",
        (date_time >= as.POSIXct("2024-11-08 11:22:00")  & date_time <= as.POSIXct("2024-11-08 11:42:00")) ~ "CHL10",
        (date_time >= as.POSIXct("2024-11-08 12:20:00")  & date_time <= as.POSIXct("2024-11-08 12:34:00")) ~ "BG13",
        (date_time >= as.POSIXct("2024-11-08 14:00:00") & date_time <= as.POSIXct("2024-11-08 14:23:00")) ~ "CHL19",
        (date_time >= as.POSIXct("2024-11-08 14:59:00") & date_time <= as.POSIXct("2024-11-08 15:20:00")) ~ "CHL18",
        (date_time >= as.POSIXct("2024-11-08 15:51:00") & date_time <= as.POSIXct("2024-11-08 16:05:00")) ~ "BG16",
        (date_time >= as.POSIXct("2024-11-08 17:13:00") & date_time <= as.POSIXct("2024-11-08 17:29:00")) ~ "CHL15",
        (date_time >= as.POSIXct("2024-11-08 17:55:00") & date_time <= as.POSIXct("2024-11-08 18:14:00")) ~ "CNL1",
        
        
        (date_time >= as.POSIXct("2024-11-09 07:40:00") & date_time <= as.POSIXct("2024-11-09 08:14:00")) ~ "CNL2",
        (date_time >= as.POSIXct("2024-11-09 09:20:00") & date_time <= as.POSIXct("2024-11-09 09:47:00")) ~ "CHL12",
        (date_time >= as.POSIXct("2024-11-09 10:56:00") & date_time <= as.POSIXct("2024-11-09 11:20:00")) ~ "CHL13",
        (date_time >= as.POSIXct("2024-11-09 15:52:00") & date_time <= as.POSIXct("2024-11-09 16:09:00")) ~ "CHL20",
        (date_time >= as.POSIXct("2024-11-09 17:03:00") & date_time <= as.POSIXct("2024-11-09 17:14:00")) ~ "CHL21",
        (date_time >= as.POSIXct("2024-11-09 17:35:00") & date_time <= as.POSIXct("2024-11-09 17:52:00")) ~ "BG18",
        (date_time >= as.POSIXct("2024-11-09 19:33:00") & date_time <= as.POSIXct("2024-11-09 19:50:00")) ~ "BG7",
           
           
           TRUE ~ NA_character_
         ))

c3_nov2024_clean <- c3_nov2024 %>% 
  filter(!is.na(code))




```

CHL21 data is still quite funky.... 


+   Frequency distribution for Nov 2024 cruz C3 data

```{r}

c3_nov2024_clean %>% 
  filter(chl > 0) %>% 
  mutate(code = as.factor(code)) %>% 
ggplot(aes(x = chl)) +
  geom_histogram(bins = 30, binwidth = 2, boundary = 0, fill = "steelblue", color = "black") +
  facet_wrap(~ code) +
  labs(x = "Chlorophyll (RFU)", y = "Frequency") +
  theme_bw() +
  theme(text = element_text(size = 15, colour = "black"),
        axis.text.x = element_text(size=14, angle = 45, hjust = 1, vjust = 1))+
 scale_x_continuous(
  breaks = scales::pretty_breaks(n = 6)  #ggplot makes breaks based on data
)


ggsave(file="plots/c3/2024_plots/c3_2024NOVcruz_site_freq_distribution.png", width = 10, height = 12, dpi = 300)

```


+ Removing outliers from Nov cruise data

```{r}
c3_n24_outlier_detect <- multidetect(data = c3_nov2024_clean, 
                                var = "chl", 
                                var_col = "code",
                                multiple = TRUE,
                                methods = c("adjbox", "iqr", "hampel","jknife",
                                            "seqfences", "mixediqr",
                                            "distboxplot", "semiqr",
                                            "zscore", "logboxplot", "medianrule"),
                       silence_true_errors = TRUE, 
                       missingness = 0.1,
                       sdm = FALSE,
                       na.inform = TRUE)
  
  

#Visualize the number of outliers detected by each method

ggoutliers(c3_n24_outlier_detect,
                        select = unique(c3_nov2024_clean$code),
                        color='purple',
                        desc = TRUE,
                        ncol = 10,
                        nrow = 10)

ggsave(file = "plots/c3/2024_plots/outliersNOV_per_site.png", width = 10, height = 20, dpi = 300)


#Obtaining quality controlled dataset using loess method or data labeling


out_qc_loessN24 <- extract_clean_data(refdata = c3_nov2024_clean, 
            outliers = c3_n24_outlier_detect, loess = TRUE,
            var_col = "code"
            )  #set threshold to 0.8 to remove the very strong outliers only

#clean dataset
nrow(out_qc_loessN24)

#reference data
nrow(c3_nov2024_clean)

out_qc_labeledN24 <- classify_data(refdata = c3_nov2024_clean, 
                                  var_col = "code",
                                outliers = c3_n24_outlier_detect)


#Visualize labelled quality controlled dataset

ggenvironmentalspace(out_qc_labeledN24, 
                     type = '1D',
                     ggxangle = 45, 
                     ncol = 10,
                     nrow = 15,
                     scalecolor = 'viridis',
                     xhjust = 1,
                     legend_position = 'blank',
                     ylab = "Number of records",
                     xlab = "Outlier labels")


ggsave(filename = "plots/c3/2024_plots/filtered_siteNOV_data.png", width = 14, height = 14, dpi = 300)



#labelled outlier or not 
labeledN24 <- classify_data(refdata = c3_nov2024_clean, 
                         var_col = "code",
                         outliers = c3_n24_outlier_detect)



#plot clean dataset

out_qc_loessN24 %>% 
mutate(code = as.factor(code)) %>% 
ggplot(aes(x = chl)) +
  geom_histogram(bins = 30, binwidth = 2, boundary = 0, fill = "steelblue", color = "black") +
facet_wrap(~ code, scales = "free_y")+
  labs(x = "Chlorophyll (ug/l)", y = "Frequency") +
  theme_bw() +
  theme(text = element_text(size = 10, colour = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))+
 scale_x_continuous(
  breaks = scales::pretty_breaks(n = 6)  #ggplot makes breaks based on data
)

ggsave(file="plots/2024_plots/c3_2024_loessfilter_SitesNOV_freq_distribution.png", width = 10, height = 12, dpi = 300)



```

CHL21 is also very problematic here



+ Get final dataset

```{r}


final_crnov_2024_df <- out_qc_loessN24 %>%
  group_by(code, year) %>%
  summarise(
    chl = mean(chl, na.rm = TRUE),
    pc = mean(pc, na.rm = TRUE),
    temp = mean(temp, na.rm = TRUE),
    depth = mean(depth, na.rm = TRUE),
    source_file  = unique(source_file),
  ) 


```


+ April 2025 time stamps

```{r}

c3_apr2025 <- c3_cruise_files_b %>%
  filter(source_file %in% "c3_CR_APR2025.csv") %>% 
  mutate(date_time = as.POSIXct(date_time, format = "%m/%d/%y %H:%M:%S"),
         code = case_when(
        (date_time >= as.POSIXct("2025-04-05 08:37:00") & date_time <= as.POSIXct("2025-04-05 08:45:00")) ~ "BG5",
        (date_time >= as.POSIXct("2025-04-05 10:40:00") & date_time <= as.POSIXct("2025-04-05 11:00:00")) ~ "CHL27",
        
        (date_time >= as.POSIXct("2025-04-05 11:33:00") & date_time <= as.POSIXct("2025-04-05 11:36:00")) ~ "CHL1",
        (date_time >= as.POSIXct("2025-04-05 11:38:00") & date_time <= as.POSIXct("2025-04-05 11:42:00")) ~ "CHL2",
        (date_time >= as.POSIXct("2025-04-05 13:08:00") & date_time <= as.POSIXct("2025-04-05 13:40:00")) ~ "BG7",
        (date_time >= as.POSIXct("2025-04-05 14:54:00") & date_time <= as.POSIXct("2025-04-05 15:02:00")) ~ "CHL3",
        (date_time >= as.POSIXct("2025-04-05 15:52:00") & date_time <= as.POSIXct("2025-04-05 16:00:00")) ~ "CHL4",
        (date_time >= as.POSIXct("2025-04-05 16:40:00") & date_time <= as.POSIXct("2025-04-05 17:00:00")) ~ "CHL5",
        (date_time >= as.POSIXct("2025-04-05 17:24:00")  & date_time <= as.POSIXct("2025-04-05 17:44:00")) ~ "CHL6",
        (date_time >= as.POSIXct("2025-04-05 18:19:00")  & date_time <= as.POSIXct("2025-04-05 18:37:00"))  ~ "CHL7",
        
        (date_time >= as.POSIXct("2025-04-06 17:55:00")  & date_time <= as.POSIXct("2025-04-06 18:15:00")) ~ "BG10",
        (date_time >= as.POSIXct("2025-04-06 19:03:00")  & date_time <= as.POSIXct("2025-04-06 19:30:00")) ~ "CHL9",
        
        (date_time >= as.POSIXct("2025-04-07 09:16:00") & date_time <= as.POSIXct("2025-04-07 09:28:00")) ~ "CHL19",
        (date_time >= as.POSIXct("2025-04-07 10:12:00") & date_time <= as.POSIXct("2025-04-07 10:24:00")) ~ "CHL18",
        (date_time >= as.POSIXct("2025-04-07 11:14:00") & date_time <= as.POSIXct("2025-04-07 11:26:00")) ~ "CHL15",
        (date_time >= as.POSIXct("2025-04-07 12:06:00") & date_time <= as.POSIXct("2025-04-07 12:30:00")) ~ "CNL1",
        (date_time >= as.POSIXct("2025-04-07 15:18:00") & date_time <= as.POSIXct("2025-04-07 15:38:00")) ~ "CNL2",
        (date_time >= as.POSIXct("2025-04-07 17:00:00") & date_time <= as.POSIXct("2025-04-07 17:30:00")) ~ "CHL12",
        (date_time >= as.POSIXct("2025-04-07 07:09:00") & date_time <= as.POSIXct("2025-04-07 07:17:00")) ~ "BG13",
        (date_time >= as.POSIXct("2025-04-07 17:47:00") & date_time <= as.POSIXct("2025-04-07 18:05:00")) ~ "CHL13",
        
        
        (date_time >= as.POSIXct("2025-04-08 07:20:00") & date_time <= as.POSIXct("2025-04-08 07:43:00")) ~ "BG16",
        (date_time >= as.POSIXct("2025-04-08 10:37:00") & date_time <= as.POSIXct("2025-04-08 10:45:00")) ~ "CHL20",
        (date_time >= as.POSIXct("2025-04-08 11:53:00") & date_time <= as.POSIXct("2025-04-08 12:03:00")) ~ "CHL21",
        (date_time >= as.POSIXct("2025-04-08 12:58:00") & date_time <= as.POSIXct("2025-04-08 13:10:00")) ~ "CHL22",
        (date_time >= as.POSIXct("2025-04-08 13:28:00") & date_time <= as.POSIXct("2025-04-08 13:40:00")) ~ "BG18",
        (date_time >= as.POSIXct("2025-04-08 14:25:00") & date_time <= as.POSIXct("2025-04-08 14:35:00")) ~ "CHL23B",
        (date_time >= as.POSIXct("2025-04-08 16:12:00") & date_time <= as.POSIXct("2025-04-08 16:29:00")) ~ "CHL24",
           
           
           TRUE ~ NA_character_
         ))

c3_apr2025_clean <- c3_apr2025 %>% 
  filter(!is.na(code))







```



+   Frequency distribution for April 2025 cruz C3 data

```{r}

c3_apr2025_clean %>% 
  filter(chl > 0) %>% 
  mutate(code = as.factor(code)) %>% 
ggplot(aes(x = chl)) +
  geom_histogram(bins = 30, binwidth = 2, boundary = 0, fill = "steelblue", color = "black") +
  facet_wrap(~ code) +
  labs(x = "Chlorophyll (RFU)", y = "Frequency") +
  theme_bw() +
  theme(text = element_text(size = 15, colour = "black"),
        axis.text.x = element_text(size=14, angle = 45, hjust = 1, vjust = 1))+
 scale_x_continuous(
  breaks = scales::pretty_breaks(n = 6)  #ggplot makes breaks based on data
)


ggsave(file="plots/c3/2025_plots/c3_2025APRcruz_site_freq_distribution.png", width = 10, height = 10, dpi = 300)

```



+ Removing outliers from April cruise data

```{r}

c3_Apr25_outlier_detect <- multidetect(data = c3_apr2025_clean, 
                                var = "chl", 
                                var_col = "code",
                                multiple = TRUE,
                                methods = c("adjbox", "iqr", "hampel","jknife",
                                            "seqfences", "mixediqr",
                                            "distboxplot", "semiqr",
                                            "zscore", "logboxplot", "medianrule"),
                       silence_true_errors = TRUE, 
                       missingness = 0.1,
                       sdm = FALSE,
                       na.inform = TRUE)
  
  

#Visualize the number of outliers detected by each method

ggoutliers(c3_Apr25_outlier_detect,
                        select = unique(c3_apr2025_clean$code),
                        color='purple',
                        desc = TRUE,
                        ncol = 10,
                        nrow = 10)

ggsave(file = "plots/c3/2025_plots/outliersAPR25_per_site.png", width = 10, height = 10, dpi = 300)


#Obtaining quality controlled dataset using loess method or data labeling


out_qc_loessAPR25 <- extract_clean_data(refdata = c3_apr2025_clean, 
            outliers = c3_Apr25_outlier_detect, loess = TRUE,
            var_col = "code"
            )  #set threshold to 0.8 to remove the very strong outliers only

#clean dataset
nrow(out_qc_loessAPR25)

#reference data
nrow(c3_apr2025_clean)

out_qc_labeledAPR25 <- classify_data(refdata = c3_apr2025_clean, 
                                  var_col = "code",
                                outliers = c3_Apr25_outlier_detect)


#Visualize labelled quality controlled dataset

ggenvironmentalspace(out_qc_labeledAPR25, 
                     type = '1D',
                     ggxangle = 45, 
                     ncol = 10,
                     nrow = 10,
                     scalecolor = 'viridis',
                     xhjust = 1,
                     legend_position = 'blank',
                     ylab = "Number of records",
                     xlab = "Outlier labels")


ggsave(filename = "plots/c3/2025_plots/filtered_siteAPR25_data.png", width = 14, height = 10, dpi = 300)



#labelled outlier or not 
labeledAPR25 <- classify_data(refdata = c3_apr2025_clean, 
                         var_col = "code",
                         outliers = c3_Apr25_outlier_detect)



#plot clean dataset

out_qc_loessAPR25 %>% 
mutate(code = as.factor(code)) %>% 
ggplot(aes(x = chl)) +
  geom_histogram(bins = 30, binwidth = 2, boundary = 0, fill = "steelblue", color = "black") +
facet_wrap(~ code, scales = "free_y")+
  labs(x = "Chlorophyll (ug/l)", y = "Frequency") +
  theme_bw() +
  theme(text = element_text(size = 10, colour = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))+
 scale_x_continuous(
  breaks = scales::pretty_breaks(n = 6)  #ggplot makes breaks based on data
)

ggsave(file="plots/2025_plots/loessfilter_siteAPR_freq_distribution.png", width = 10, height = 10, dpi = 300)



```





+ Get final dataset

```{r}


final_crapr_2025_df <- out_qc_loessAPR25 %>% 
  group_by(code, year) %>%
  summarise(
    chl = mean(chl, na.rm = TRUE),
    pc = mean(pc, na.rm = TRUE),
    temp = mean(temp, na.rm = TRUE),
    depth = mean(depth, na.rm = TRUE),
    source_file  = unique(source_file),
  ) 


```


+ Jun 2025


```{r}


c3_jun2025 <- c3_cruise_files_b %>%
  filter(source_file %in% "c3_CR_JUN2025.csv") %>% 
  mutate(date_time = as.POSIXct(date_time, format = "%m/%d/%Y %H:%M"),
         code = case_when(
       
        (date_time >= as.POSIXct("2025-06-23 07:34:00") & date_time <= as.POSIXct("2025-06-23 08:00:00")) ~ "CHL26",
        (date_time >= as.POSIXct("2025-06-23 09:06:00") & date_time <= as.POSIXct("2025-06-23 09:17:00")) ~ "CHL25",
        (date_time >= as.POSIXct("2025-06-23 10:31:00") & date_time <= as.POSIXct("2025-06-23 10:46:00")) ~ "BG5",
        (date_time >= as.POSIXct("2025-06-23 12:30:00") & date_time <= as.POSIXct("2025-06-23 12:42:00")) ~ "CHL27",
        
        (date_time >= as.POSIXct("2025-06-23 13:42:00") & date_time <= as.POSIXct("2025-06-23 13:55:00")) ~ "CHL3",
        (date_time >= as.POSIXct("2025-06-23 14:18:00") & date_time <= as.POSIXct("2025-06-23 14:23:00")) ~ "BG7",
        (date_time >= as.POSIXct("2025-06-23 15:12:00") & date_time <= as.POSIXct("2025-06-23 15:17:00")) ~ "CHL2",
        (date_time >= as.POSIXct("2025-06-23 15:40:00") & date_time <= as.POSIXct("2025-06-23 16:00:00")) ~ "CHL1",
        (date_time >= as.POSIXct("2025-06-23 17:00:00") & date_time <= as.POSIXct("2025-06-23 17:15:00")) ~ "CHL4",
        (date_time >= as.POSIXct("2025-06-23 17:29:00") & date_time <= as.POSIXct("2025-06-23 17:38:00")) ~ "CHL5",
        
        (date_time >= as.POSIXct("2025-06-23 18:02:00") & date_time <= as.POSIXct("2025-06-23 18:15:00")) ~ "CHL6",
        (date_time >= as.POSIXct("2025-06-23 18:46:00") & date_time <= as.POSIXct("2025-06-23 18:55:00")) ~ "CHL7",
        
        (date_time >= as.POSIXct("2025-06-24 07:30:00") & date_time <= as.POSIXct("2025-06-24 07:48:00")) ~ "BG10",
        
        (date_time >= as.POSIXct("2025-06-24 08:52:00")  & date_time <= as.POSIXct("2025-06-24 09:02:00")) ~ "CHL9",
        (date_time >= as.POSIXct("2025-06-24 10:07:00")  & date_time <= as.POSIXct("2025-06-24 10:20:00"))  ~ "CHL10",
        (date_time >= as.POSIXct("2025-06-24 10:50:00")  & date_time <= as.POSIXct("2025-06-24 10:58:00")) ~ "BG13",
        
        (date_time >= as.POSIXct("2025-06-24 10:50:00") & date_time <= as.POSIXct("2025-06-24 10:58:00")) ~ "BG13",
        (date_time >= as.POSIXct("2025-06-24 12:23:00") & date_time <= as.POSIXct("2025-06-24 12:33:00")) ~ "CHL19",
        (date_time >= as.POSIXct("2025-06-24 13:18:00") & date_time <= as.POSIXct("2025-06-24 13:25:00")) ~ "CHL18",
        (date_time >= as.POSIXct("2025-06-24 14:18:00") & date_time <= as.POSIXct("2025-06-24 14:29:00")) ~ "CHL15",
        
     
        (date_time >= as.POSIXct("2025-06-24 15:09:00") & date_time <= as.POSIXct("2025-06-24 15:28:00")) ~ "CNL1",
        (date_time >= as.POSIXct("2025-06-24 17:13:00") & date_time <= as.POSIXct("2025-06-24 17:28:00")) ~ "CNL2",
        (date_time >= as.POSIXct("2025-06-24 18:22:00") & date_time <= as.POSIXct("2025-06-24 18:32:00")) ~ "CHL12",
        (date_time >= as.POSIXct("2025-06-24 19:53:00") & date_time <= as.POSIXct("2025-06-24 19:59:00")) ~ "CHL13",
        
        (date_time >= as.POSIXct("2025-06-25 09:33:00") & date_time <= as.POSIXct("2025-06-25 09:53:00")) ~ "BG16",
        (date_time >= as.POSIXct("2025-06-25 15:05:00") & date_time <= as.POSIXct("2025-06-25 15:17:00")) ~ "CHL20",
        (date_time >= as.POSIXct("2025-06-25 16:02:00") & date_time <= as.POSIXct("2025-06-25 16:15:00")) ~ "CHL21",
        (date_time >= as.POSIXct("2025-06-25 16:26:00") & date_time <= as.POSIXct("2025-06-25 16:34:00")) ~ "CHL22",
        (date_time >= as.POSIXct("2025-06-25 16:49:00") & date_time <= as.POSIXct("2025-06-25 16:54:00")) ~ "BG18",
        (date_time >= as.POSIXct("2025-06-25 18:29:00") & date_time <= as.POSIXct("2025-06-25 18:39:00")) ~ "CHL23B",
        (date_time >= as.POSIXct("2025-06-25 19:15:00") & date_time <= as.POSIXct("2025-06-25 19:30:00")) ~ "CHL24",
           
           
           TRUE ~ NA_character_
         ))

c3_jun2025_clean <- c3_jun2025 %>% 
  filter(!is.na(code))




```




+   Frequency distribution for June 2025 cruz C3 data

```{r}

c3_jun2025_clean %>% 
  filter(chl > 0) %>% 
  mutate(code = as.factor(code)) %>% 
ggplot(aes(x = chl)) +
  geom_histogram(bins = 30, binwidth = 2, boundary = 0, fill = "steelblue", color = "black") +
  #facet_wrap(~ code, scales="free_y") +
  labs(x = "Chlorophyll (RFU)", y = "Frequency") +
  theme_bw() +
  theme(text = element_text(size = 15, colour = "black"),
        axis.text.x = element_text(size=14, angle = 45, hjust = 1, vjust = 1))+
 scale_x_continuous(
  breaks = scales::pretty_breaks(n = 6)  #ggplot makes breaks based on data
)


ggsave(file="plots/c3/2025_plots/c3_2025JUNcruz_gen_freq_distribution.png", width = 10, height = 12, dpi = 300)

```



+ Removing outliers from June cruise data

```{r}

c3_Jun25_outlier_detect <- multidetect(data = c3_jun2025_clean, 
                                var = "chl", 
                                var_col = "code",
                                multiple = TRUE,
                                methods = c("adjbox", "iqr", "hampel","jknife",
                                            "seqfences", "mixediqr",
                                            "distboxplot", "semiqr",
                                            "zscore", "logboxplot", "medianrule"),
                       silence_true_errors = TRUE, 
                       missingness = 0.1,
                       sdm = FALSE,
                       na.inform = TRUE)
  
  

#Visualize the number of outliers detected by each method

ggoutliers(c3_Jun25_outlier_detect,
                        select = unique(c3_jun2025_clean$code),
                        color='purple',
                        desc = TRUE,
                        ncol = 10,
                        nrow = 10)

ggsave(file = "plots/c3/2025_plots/outliersJUN25_per_site.png", width = 10, height = 10, dpi = 300)


#Obtaining quality controlled dataset using loess method or data labeling


out_qc_loessJUN25 <- extract_clean_data(refdata = c3_jun2025_clean, 
            outliers = c3_Jun25_outlier_detect, loess = TRUE,
            var_col = "code"
            )  #set threshold to 0.8 to remove the very strong outliers only

#clean dataset
nrow(out_qc_loessJUN25)

#reference data
nrow(c3_jun2025_clean)

out_qc_labeledJUN25 <- classify_data(refdata = c3_jun2025_clean, 
                                  var_col = "code",
                                outliers = c3_Jun25_outlier_detect)


#Visualize labelled quality controlled dataset

ggenvironmentalspace(out_qc_labeledJUN25, 
                     type = '1D',
                     ggxangle = 45, 
                     ncol = 10,
                     nrow = 10,
                     scalecolor = 'viridis',
                     xhjust = 1,
                     legend_position = 'blank',
                     ylab = "Number of records",
                     xlab = "Outlier labels")


ggsave(filename = "plots/c3/2025_plots/filtered_siteJUN25_data.png", width = 14, height = 10, dpi = 300)



#labelled outlier or not 
labeledJUN25 <- classify_data(refdata = c3_jun2025_clean, 
                         var_col = "code",
                         outliers = c3_Jun25_outlier_detect)



#plot clean dataset

out_qc_loessJUN25 %>% 
mutate(code = as.factor(code)) %>% 
ggplot(aes(x = chl)) +
  geom_histogram(bins = 30, binwidth = 2, boundary = 0, fill = "steelblue", color = "black") +
facet_wrap(~ code)+
  labs(x = "Chlorophyll (ug/l)", y = "Frequency") +
  theme_bw() +
  theme(text = element_text(size = 10, colour = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))+
 scale_x_continuous(
  breaks = scales::pretty_breaks(n = 6)  #ggplot makes breaks based on data
)

ggsave(file="plots/2025_plots/c3_2025_loessfilter_sitesJUN_freq_distribution.png", width = 10, height = 10, dpi = 300)



```



+ Get final dataset

```{r}


final_crjun_2025_df <- out_qc_loessJUN25 %>% 
  group_by(code, year) %>%
  summarise(
    chl = mean(chl, na.rm = TRUE),
    pc = mean(pc, na.rm = TRUE),
    temp = mean(temp, na.rm = TRUE),
    depth = mean(depth, na.rm = TRUE),
    source_file  = unique(source_file),
  ) 

```



+ Join c3 cruise datasets

```{r}

c3_cruises_data <- bind_rows(final_crjun_2023_df, final_crjul_2024_df, final_crnov_2024_df,    final_crapr_2025_df, final_crjun_2025_df) %>% 
  arrange(source_file) %>% 
  mutate(source_file = str_remove(source_file, "\\.csv$"))

```


+ c3 files with extracted chl

```{r}

file_paths.csv_extc3 <- dir("c3_extracted chls", "*.csv", full.names = TRUE)

c3_j2023_extchl <- read_csv(file_paths.csv_extc3[1]) %>%
  select(code, ext_chl, sd, srp_ug_l, nitrates_ug_l, source_file) %>% 
  filter(!is.na(ext_chl))

  


c3_2024_25_extchl <- read_csv(file_paths.csv_extc3[2]) %>% 
  clean_names() %>% 
  filter(study %in% "CR") %>% 
  rename(code = site,
         ext_chl = florometer_chl_ug_l) %>% 
  select(code, ext_chl, sd, srp_ug_l,nitrates_ug_l,source_file) %>% 
  mutate(sd = as.numeric(sd)) 


c3_extchl_files <- bind_rows(c3_j2023_extchl, c3_2024_25_extchl) %>% 
  mutate(source_file = str_remove(source_file, "\\.csv$")) # remove csv extension from file name, and $ means at the end of name to avoid accidental removals


```

+ joining ext_chl with c3 sonde files

```{r}


c3_calib_data <- full_join(c3_cruises_data, c3_extchl_files, by = c("code", "source_file")) %>% 
  mutate(
    source_file_clean = ifelse(
      source_file %in% c("c3_CR_27_June_2023", "Turner_C3_28June2023_cruise"),
      "cruise-c3data-2023",
      source_file
    )
  ) 


 

```



# 2024 experiments

 + c3

```{r}


c3_spiking_expt24 <- expts_24 %>% 
  filter(sos == "c3") %>% 
  mutate(code = site,
         chl = top_chl,
         pc = top_bga,
         file_name = "2024_c3sonde_experiment",
         data_type = "experiment") %>% 
  mutate(year = year(date_time)) %>% 
  dplyr::select(code, date_time, chl, pc, ext_chl, file_name, data_type)



```


+ Linear regression for chl_c3 ~ ext_chl

+   check slopes across years

```{r}

c3_calib_data %>% 
  filter(!is.na(code), !is.na(chl)) %>% 
  arrange(source_file_clean, chl) %>% 
  ggplot(aes(x = chl, y = ext_chl)) +
  #facet_wrap(~source_file_clean) +  
geom_point(color = "steelblue", size = 3.5) +
  geom_smooth(method = "lm", se = FALSE, color = "red", linewidth = 1.5) +
  stat_regline_equation(
    aes(label = paste(after_stat(eq.label), ..rr.label.., sep = "~~~~")),
    size = 15,
    show.legend = FALSE
  ) +
  theme_bw() +
  theme(
    panel.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_line(colour = "gray90", linetype = "dotted", linewidth = 0.5),
    legend.position = "bottom",
      legend.text = element_text(size = 20),   # 👈 increases legend labels
  legend.title = element_text(size = 18),
  
    axis.text = element_text(size = 30, margin = margin(t = 15)),
    axis.title = element_text(size = 30),
    axis.text.x = element_text(vjust = 1, hjust = 1),
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
  ) +
  guides(color = guide_legend(override.aes = list(size = 8)))
  

ggsave(filename = "plots/CALIBc3/rship_c3chl~ext_gen_origdatachl60.png", width = 20, height = 15, dpi = 300)
  
  
#model
model_orig <- lm(ext_chl ~ chl, data = c3_calib_data)


plot(model_orig)


#Check prediction error explicitly with the original data

c3_calib_data$pred <- predict(model_orig, newdata = c3_calib_data)

ggplot(c3_calib_data, aes(chl, ext_chl - pred)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed")



```

In the original data model, the residuals vs fiited plot shows points skewed to the left and not spread out. The  qqnorm line is more flattened. The scale location plot increasing variance or spread. The leverage plot shows some influential points. However, the ext_chl~ c3chl model shows: the intercept of .... 

The residuals/error plot shows that, low chl, residuals are somewhat tight (but still spread), at high chl, residuals explode (very large positive & negative errors). There is heteroscedasticity or non-constant variance of data points.

**Systematic bias**
Many residuals at low chl are slightly negative
→ model tends to overpredict at low values
Some very large positive residuals
→ model underpredicts high ext_chl

This suggests the relationship is not linear. Predictions get worse as chl increases. 
The nonlinearity is largely due to scale


try a residual plot
```{r}

c3_calib_data_clean <- c3_calib_data %>% 
  filter(!is.na(ext_chl), !is.na(chl))




model_log <- lm(log10(ext_chl) ~ log10(chl), data = c3_calib_data_clean)
model_sqrt <- lm(sqrt(ext_chl) ~ sqrt(chl), data = c3_calib_data_clean)


c3_calib_data$pred_log <- predict(model_log, newdata = c3_calib_data_clean)
c3_calib_data$pred_sqrt <- predict(model_sqrt, newdata = c3_calib_data_clean)

c3_calib_data$resid_log <- log10(c3_calib_data$ext_chl) - c3_calib_data_clean$pred_log
c3_calib_data$resid_sqrt <- sqrt(c3_calib_data$ext_chl) - c3_calib_data_clean$pred_sqrt


#log
ggplot(c3_calib_data_clean, aes(log10(chl), resid_log)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed")


#sqrt
ggplot(c3_calib_data_clean, aes(sqrt(chl), resid_sqrt)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed")



```


**Data transformation**



+   log transformed

```{r}


c3_calib_data_clean %>% 
  ggplot(aes(log10(chl), log10(ext_chl))) +
  geom_point(size = 3.5) +
  geom_smooth(method = "lm", se = FALSE, color = "red", linewidth = 1.5) +
  stat_regline_equation(
    aes(label = paste(after_stat(eq.label), ..rr.label.., sep = "~~~~")),
    size = 15,
    show.legend = FALSE
  ) +
  theme_bw() +
  theme(
    panel.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_line(colour = "gray90", linetype = "dotted", linewidth = 0.5),
    legend.position = "bottom",
      legend.text = element_text(size = 20),   # 👈 increases legend labels
  legend.title = element_text(size = 18),
  
    axis.text = element_text(size = 30, margin = margin(t = 15)),
    axis.title = element_text(size = 30),
    axis.text.x = element_text(vjust = 1, hjust = 1),
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
  ) +
  guides(color = guide_legend(override.aes = list(size = 8)))
  

ggsave(filename = "plots/CALIBc3/rship_c3chl~ext_transformed_gen.png", width = 20, height = 20, dpi = 300)




#facetted
c3_calib_data %>% 
   filter(source_file_clean != "c3_CR_JUN2025") %>% 
  ggplot(aes(log10(chl), log10(ext_chl), color = source_file_clean)) +
  geom_point(size = 4) +
  geom_smooth(method = "lm", se = FALSE, color = "red", linewidth = 1.5) +
  facet_wrap(~ source_file_clean, scales = "free_y") +
  stat_regline_equation(
    aes(label = paste(after_stat(eq.label), ..rr.label.., sep = "~~~~")),
    size = 6,
    show.legend = FALSE
  ) +
  theme_bw(base_size = 18) +    #for scaling facets so they all look the same
  theme(
    panel.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_line(colour = "gray90", linetype = "dotted", linewidth = 0.5),
    legend.position = "bottom",
    legend.text = element_text(size = 16),
    legend.title = element_text(size = 18),
    axis.text = element_text(size = 18, margin = margin(t = 15)),
    axis.title = element_text(size = 20),
    axis.text.x = element_text(vjust = 1, hjust = 1),
    strip.text = element_text(size = 16, face = "bold"),
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold")
  ) +
  guides(color = guide_legend(override.aes = list(size = 5)))
  
  

ggsave(filename = "plots/CALIBc3/rship_c3chl~ext_transformedfacet_nojune.png", width = 30, height = 20, dpi = 300)

```

+   sqrt transformed

```{r}

c3_calib_data %>% 
  ggplot(aes(sqrt(chl), sqrt(ext_chl))) +
  geom_point(size = 3.5) +
  geom_smooth(method = "lm", se = FALSE, color = "red", linewidth = 1.5) +
  stat_regline_equation(
    aes(label = paste(after_stat(eq.label), ..rr.label.., sep = "~~~~")),
    size = 15,
    show.legend = FALSE
  ) +
  theme_bw() +
  theme(
    panel.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_line(colour = "gray90", linetype = "dotted", linewidth = 0.5),
    legend.position = "bottom",
      legend.text = element_text(size = 20),   # 👈 increases legend labels
  legend.title = element_text(size = 18),
  
    axis.text = element_text(size = 30, margin = margin(t = 15)),
    axis.title = element_text(size = 30),
    axis.text.x = element_text(vjust = 1, hjust = 1),
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
  ) +
  guides(color = guide_legend(override.aes = list(size = 8)))
  

ggsave(filename = "plots/CALIBc3/rship_c3chl~ext_sqrttransformedfacet.png", width = 30, height = 20, dpi = 300)



```



c3 log model with no env _ log transformed chl

```{r}

model_log <- lm(log(ext_chl) ~ log(chl), data = c3_calib_data_clean)
  
  
summary(model_log)

par(mfrow = c(2, 2))
plot(model_log)


```

Check diagnostics

```{r}
library(car)


#check for multicollinearity
vif(model_2) #If VIF > 5–10 → variables are overlapping. Which is not the case here


#check residuals normality..etc
par(mfrow = c(2, 2))
plot(model_2)

```



c3 model with env: Does the slope of chl → ext_chl change with environment?

Is my sonde calibration stable across environmental gradients?

+ check visuals with env
```{r}

### check r/ships with ext_chl_Need for data transformation?

ggplot(c3_calib_data, aes(temp, log10(ext_chl))) + geom_point()
ggplot(c3_calib_data, aes(pc, log10(ext_chl))) + geom_point()
ggplot(c3_calib_data, aes(sd, log10(ext_chl))) + geom_point()


#sd relationship _Yes, log trasnform for sd and pc. No need for temp
ggplot(c3_calib_data, aes(sd, log10(ext_chl))) +
  geom_point() +
  geom_smooth(method = "loess", se = FALSE, color = "red")



#Plot slopes by environment....... sd seems to show some interaction coz the slope at high sd changes
c3_calib_data %>% 
  filter(!is.na(sd)) %>% 
ggplot(aes(log10(chl), log10(ext_chl))) +
  geom_point() +
  geom_smooth(method = "lm") +
  facet_wrap(~ cut(log10(sd), 3))

```

+ models with interaction

```{r}


#This avoids collinearity issues and makes interactions interpretable.


c3_model_data <- c3_calib_data_clean %>%
  mutate(
    log_ext_chl = log10(ext_chl),
    log_chl     = log10(chl),
    log_sd      = log10(sd),
    log_pc      = log10(pc)
  ) %>%
  mutate(
    log_chl_c = scale(log_chl, scale = FALSE),
    log_sd_c  = scale(log_sd,  scale = FALSE),
    log_pc_c  = scale(log_pc,  scale = FALSE),
    temp_c    = scale(temp,    scale = FALSE)
  )


#Fit FULL interaction model (Main effects: chl, pc, temp and sd).. plus interaction chl*pc, chl*sd, chl*temp

model_int <- lm(
  log_ext_chl ~ log_chl_c * (log_sd_c + log_pc_c + temp_c),
  data = c3_model_data
)


summary(model_int)


#check for multicollinaerity
vif(model_int)


#Only significant variables
model_int2 <- lm(
  log_ext_chl ~ log_chl_c * (log_pc_c),
  data = c3_model_data)


summary(model_int2)


```

check diagnostics
```{r}

par(mfrow = c(2, 2))
plot(model_int)


```


Compare models with the simpler model

```{r}

anova(model_orig, model_log, model_sqrt, model_int)

AIC(model_orig, model_log, model_sqrt, model_int)

```


The AICs in the interaction model and sqrt model are higher than in the simpler model so it doesn't improve our chl prediction.


+ Try a GAM
```{r}


gam_model <- gam(ext_chl ~ s(chl), data = c3_calib_data)

summary(gam_model)
plot(gam_model)


gam_log <- gam(log(ext_chl) ~ s(log(chl)), data = c3_calib_data)

summary(gam_log)
plot(gam_log)







```
GAM is detecting the same issue of nonlinearity but since log transformation fixed the diagnostics for linear data, then we can just use the lm from log transformation.

**effective degrees of freedom**
edf ≈ 1 → relationship is linear
edf > 1 → relationship is nonlinear (curved)
edf >> 1 (e.g., 4–8) → very wiggly → possible overfitting


Because edf ~ 1, more evidence to stick with the LM

**P-value**
The pvalue is < 0.05, which would imply that the chl explains the ext_chl

The deviance explained is 21.8% (model fit).





#   Loading longterm data series for model testing and calibration

+   c3 data
```{r}

file_paths.csv <- dir("c3_Dunga_time_series", pattern = "*.csv", full.names = TRUE)

first_file <- read_csv(file_paths.csv[1])
colnames(first_file)

#use function to read files

#read files

c3_dunga_files <- purrr::map_dfr(file_paths.csv, read_c3_file) 


c3_dunga_files_clean <- c3_dunga_files  %>% 
  mutate(
    date_time = as.character(date_time),
    date_time_parsed = parse_date_time(
      date_time,
      orders = c(
  "Ymd HMS",
  "Ymd HM",
  "dmy HMS",
  "dmy HM",
  "dmY HMS",
  "mdy HMS",
  "mdy HM"
))) %>%
  mutate(year= year(date_time_parsed),
         month = month(date_time_parsed),
         mmyy = format(date_time_parsed, "%m-%y")) %>% 
  arrange(date_time) %>% 
  clean_names() %>% 
filter(!(chl <= 0 | chl > 500))

#check files
table(c3_dunga_files_clean$source_file)




```




+ frequency distribution


```{r}

c3_dunga_files_clean %>% 
  mutate(code = as.factor(source_file)) %>% 
ggplot(aes(x = chl)) +
  geom_histogram(bins = 30, binwidth = 2, boundary = 0, fill = "steelblue", color = "black") +
  #facet_wrap(~ code, scales="free_y") +
  labs(x = "Chlorophyll (RFU)", y = "Frequency") +
  theme_bw() +
  theme(text = element_text(size = 15, colour = "black"),
        axis.text.x = element_text(size=14, angle = 45, hjust = 1, vjust = 1))+
 scale_x_continuous(
  breaks = scales::pretty_breaks(n = 6)  #ggplot makes breaks based on data
)


ggsave(file="plots/c3/DG_plots/gen_freq_distribution.png", width = 10, height = 12, dpi = 300)

```

+ monthly plots for raw data

```{r}

# get unique mmyy values (ordered chronologically if needed)
mmyy_levels_raw <- c3_dunga_files_clean %>%
  mutate(mmyy_date = lubridate::my(mmyy)) %>%
  arrange(mmyy_date) %>%
  pull(mmyy) %>%
  unique()

# loop
for (m in mmyy_levels_raw) {
  
  df_sub <- c3_dunga_files_clean %>%
    filter(mmyy == m) %>%
    arrange(date_time)
  
  p <- ggplot(df_sub, aes(x = date_time, y = chl)) +
    geom_line(color = "steelblue") +
    geom_point(size = 0.5, alpha = 0.5) +
   # geom_point(aes(color = chl > 200), size = 0.7)+
    geom_smooth(se = FALSE, color = "red", span = 0.2)+
    labs(
      title = paste("Chlorophyll Time Series -", m),
      x = "Date-Time",
      y = "Chlorophyll (µg/L)"
    ) +
    theme_bw() +
    theme(
      text = element_text(size = 10, colour = "black"),
      axis.text.x = element_text(angle = 45, hjust = 1)
    )
  
  ggsave(
    filename = paste0("DG_plots/raw_monthly_chl_plots/chl_timeseries_", m, ".png"),
    plot = p,
    width = 8,
    height = 4,
    dpi = 300
  )
}




```




+   remove outliers

```{r}
c3_dg_outlier_detect <- multidetect(data = c3_dunga_files_clean, 
                                var = "chl", 
                                var_col = "mmyy",
                                multiple = TRUE,
                                methods = c("adjbox", "iqr", "hampel","jknife",
                                            "seqfences", "mixediqr",
                                            "distboxplot", "semiqr",
                                            "zscore", "logboxplot", "medianrule"),
                       silence_true_errors = TRUE, 
                       missingness = 0.1,
                       sdm = FALSE,
                       na.inform = TRUE)
  
  

#Visualize the number of outliers detected by each method

ggoutliers(c3_dg_outlier_detect,
                        select = unique(c3_dunga_files_clean$mmyy),
                        color='purple',
                        desc = TRUE,
                        ncol = 10,
                        nrow = 10)

ggsave(file = "plots/c3/DG_plots/outliers_per_month.png", width = 25, height = 15, dpi = 300)


#Obtaining quality controlled dataset using loess method or data labeling


out_qc_loessdgc3 <- extract_clean_data(refdata = c3_dunga_files_clean, 
            outliers = c3_dg_outlier_detect, loess = TRUE,
            var_col = "mmyy"
            )  #set threshold to 0.8 to remove the very strong outliers only

#clean dataset
nrow(out_qc_loessdgc3)

#reference data
nrow(c3_dunga_files_clean)

out_qc_labeledgdc3 <- classify_data(refdata = c3_dunga_files_clean, 
                                  var_col = "mmyy",
                                outliers = c3_dg_outlier_detect)


#Visualize labelled quality controlled dataset

ggenvironmentalspace(out_qc_labeledgdc3, 
                     type = '1D',
                     ggxangle = 45, 
                     ncol = 10,
                     nrow = 10,
                     scalecolor = 'viridis',
                     xhjust = 1,
                     legend_position = 'blank',
                     ylab = "Number of records",
                     xlab = "Outlier labels")


ggsave(filename = "plots/c3/DG_plots/filtered_month_data.png", width = 25, height = 15, dpi = 300)



#labelled outlier or not 
labeleddgc3 <- classify_data(refdata = c3_dunga_files_clean, 
                         var_col = "mmyy",
                         outliers = c3_dg_outlier_detect)



#plot clean dataset

out_qc_loessdgc3 %>% 
  mutate(mmyy_date = my(mmyy),   # converts "01-25" → Jan 2025
  mmyy = factor(mmyy, levels = unique(mmyy[order(mmyy_date)]))       
  ) %>% 
mutate(code = as.factor(source_file)) %>% 
  filter(mmyy == "01-25") %>% 
ggplot(aes(x = chl)) +
  geom_histogram(bins = 30, binwidth = 2, boundary = 0, fill = "steelblue", color = "black") +
facet_wrap(~ mmyy, scales= "free_y")+
  labs(x = "Chlorophyll (ug/l)", y = "Frequency") +
  theme_bw() +
  theme(text = element_text(size = 10, colour = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))+
 scale_x_continuous(
  breaks = scales::pretty_breaks(n = 6)  #ggplot makes breaks based on data
)

ggsave(file="plots/c3/DG_plots/loessfilter_freq_distribution.png", width = 10, height = 10, dpi = 300)



#Plots frequency probability

out_qc_loessdgc3 %>% 
  ggplot(aes(x = chl)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30, binwidth = 2, boundary = 0,
                 fill = "steelblue", color = "black") +
  facet_wrap(~ mmyy, scales = "free_y")

```

monthly distribtution plots for clean data

```{r}


out_qc_loessdgc3 %>% 
  mutate(month = as_factor(month))%>%
  arrange(date_time) %>% 
  group_by(month, year) %>%
  summarise(avg_chl = mean(chl)) %>% 
  ggplot(aes(x = month, y = avg_chl)) +
  geom_point(color = "steelblue", size=4) +
  facet_wrap(~ year, scales = "free_y") +
  labs(
    x = "Months",
    y = "Chlorophyll (µg/L)"
  ) +
 
theme_bw() +
  theme(
    panel.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_line(colour = "gray90", linetype = "dotted", linewidth = 0.5),
    legend.position = "bottom",
      legend.text = element_text(size = 20),   # 👈 increases legend labels
  legend.title = element_text(size = 18),
  
    axis.text = element_text(size = 30, margin = margin(t = 15)),
    axis.title = element_text(size = 30),
    axis.text.x = element_text(vjust = 1, hjust = 1),
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
  ) +
  guides(color = guide_legend(override.aes = list(size = 8)))
  


ggsave(file="plots/c3/DG_plots/av_rawc3_chl.png", width = 16, height = 10, dpi = 300)



# create output folder (if it doesn’t exist)
#dir.create("monthly_chl_plots", showWarnings = FALSE)

# get unique mmyy values (ordered chronologically if needed)


mmyy_levels <- out_qc_loessdgc3 %>%
  mutate(mmyy_date = lubridate::my(mmyy)) %>%
  arrange(mmyy_date) %>%
  pull(mmyy) %>%
  unique



# loop
for (m in mmyy_levels) {
  
  df_sub <- out_qc_loessdgc3 %>%
    filter(mmyy == m) %>%
    arrange(date_time)
  
  
  p <- ggplot() +
   
    #geom_point(size = 0.5, alpha = 0.5, color="red") +
    #geom_point(aes(y=df_sub_2$chl), size = 0.1, alpha = 0.5, color="purple") +
    
    
    
  geom_point(data = df_sub, aes(x = date_time, y = chl), color = "red", size=0.1) +
    
    
    labs(
      title = paste("Chlorophyll Time Series -", m),
      x = "Date-Time",
      y = "Chlorophyll (µg/L)"
    ) +
    theme_bw() +
    theme(
      text = element_text(size = 10, colour = "black"),
      axis.text.x = element_text(angle = 45, hjust = 1)
    )
  
  ggsave(
    filename = paste0("DG_plots/monthly_chl_plots/chl_timeseries_", m, ".png"),
    plot = p,
    width = 8,
    height = 4,
    dpi = 300
  )
}






```



#   Testing the model on longterm c3 series

+ bias correction on the c3 data we used for calibration.

    + Log transformation compresses large values
    + When you back-transform, you don’t fully recover the spread
    + So predictions are biased low
    + The correction inflates them slightly to compensate so that the high values are not under   estimated
    

+  correct the bias to the series using Duan's smearing estimator or sigma
    It is more robust and doesn't assume perfect normal residues
    

```{r}

sigma22 <- summary(model_log)$sigma^2

c3_calib_data_clean$pred_ext_chl <- 10^(c3_calib_data_clean$pred_log) * exp(sigma2 / 2) #corrects log bias


# Plot observed vs predicted
ggplot(c3_calib_data_clean, aes(ext_chl, pred_ext_chl)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, color = "red")


# Calculate metrics

library(Metrics)

rmse(c3_calib_data_clean$ext_chl, c3_calib_data_clean$pred_ext_chl)
mae(c3_calib_data_clean$ext_chl, c3_calib_data_clean$pred_ext_chl)





```



# Model validation

+   Cross-validation

```{r}

#library(boot)

#cv.glm(c3_calib_data_clean, model_log, K = 10)$delta


library(dplyr)

set.seed(123)

K <- 10
folds <- sample(rep(1:K, length.out = nrow(c3_calib_data_clean)))

cv_results <- data.frame()

for (k in 1:K) {
  
  train <- c3_calib_data_clean[folds != k, ]
  test  <- c3_calib_data_clean[folds == k, ]
  
  model <- lm(log10(ext_chl) ~ log10(chl), data = train)
  
  # Predict
  test$pred_log <- predict(model, newdata = test)
  
  # Bias correction
  sigma2 <- summary(model)$sigma^2
  test$pred_ext_chl <- 10^(test$pred_log) * exp(sigma2 / 2)
  
  # Store results
  cv_results <- rbind(cv_results, test)
}

# Final RMSE
rmse(cv_results$ext_chl, cv_results$pred_ext_chl)



```


+ Apply to long term data series

```{r}


#log transform chl
out_qc_loessdgc3$log_chl <- log10(out_qc_loessdgc3$chl)

#predict new chl log transformed
out_qc_loessdgc3$pred_log <- predict(model_log, newdata = out_qc_loessdgc3)

#add the log bias correction term: converting predictions from log space back to real chlorophyll units, with bias correction

sigma_log <- summary(model_log)$sigma^2

out_qc_loessdgc3$pred_ext_chl <- 10^(out_qc_loessdgc3$pred_log) * exp(sigma_log / 2)


par(mfrow = c(1, 2))
plot(out_qc_loessdgc3$pred_log, out_qc_loessdgc3$chl, main = "No correction" )
plot(out_qc_loessdgc3$pred_ext_chl, out_qc_loessdgc3$chl, main = "With correction" )







#check range before applying

range(c3_calib_data_clean$chl)
range(out_qc_loessdgc3$chl)



```
Notes: 
10^(pred_log) → gives you the median prediction
exp(sigma²/2) → adjusts to the mean (true expected value). Exponentiate the predictions to return to the original scale

The longterm data series are within (0.13 - 200.73) calibration range compared to the c3_calibration data (2.55 - 218.5)


# MSPE 

  + this is the MSPE (Mean Square Prediction Error). average of the squared differences between observed and predicted values and we use it to validate the model
  
  + Lower is better

```{r}

ggplot() +
  geom_point(data = c3_calib_data_clean, aes(chl, ext_chl), alpha = 0.5) +
  geom_line(data = out_qc_loessdgc3, aes(chl, pred_ext_chl), color = "red")


#mspe = (yi - yhat)^2, na.rm = TRUE)

#Work on this function later

#calculate_rmspe <- function(actual, log_pred) {
 # pred <- exp(log_pred)
 # val <- sqrt(mean(((yi - yhat)^2, na.rm = TRUE))
 # return(val)
#}
#my_rmspe <- calculate_rmspe(c3_calib_data_clean$ext_chl, c3_calib_data_clean$pred_ext_chl)



yi = cv_results$ext_chl

yhat = cv_results$pred_ext_chl

mspe_model = mean((yi - yhat)^2, na.rm = TRUE)


# Root Mean Squared Prediction Error (RMSPE): Back to original units

rmspe_model = sqrt(mspe_model)


Percentage_error <- mean(abs((cv_results$ext_chl - cv_results$pred_ext_chl) / 
          cv_results$ext_chl), na.rm = TRUE) * 100

```
MSPE for the model is high `r mspe_model` but it is sensitive to outliers. Even if we cleaned the data, a few points are still controlling the fit. RMSPE is less sensitive to outliers and is the square root of mspe. So the RMSPE is `r rmspe_model`ug/l means that the predicted ext_chl by the c3 is off by rmspe_model ug/l. 

The general percentage error (How big is my prediction error compared to the true value) is `r Percentage_error`. 

**The calibration model exhibited moderate predictive performance (RMSE = 11 µg L⁻¹), with substantial uncertainty at low chlorophyll concentrations and improved accuracy at higher concentrations.**



+ Let's try to break down the chl ranges


```{r}

median_chl <- median(cv_results$chl, na.rm = TRUE)

cv_results <- cv_results %>%
  mutate(chl_group = ifelse(chl <= median_chl, "Low", "High"))


# compare rmse per group

library(dplyr)
library(Metrics)

cv_results %>%
  group_by(chl_group) %>%
  summarise(
    RMSE = rmse(ext_chl, pred_ext_chl),
    MAE  = mae(ext_chl, pred_ext_chl),
    n    = n()
  )


```


At low chl (less than median chl), `r median_chl`, the RMSE is 6.1 while at higher chl, the RMSE increases to 15.


Leave one cruise out cross validation (LOOCV)

```{r}

cruises <- unique(c3_calib_data_clean$source_file)

#Run LOOCV by cruise files

library(dplyr)


cv_date_results <- data.frame()

for (d in cruises) {
  
  train <- c3_calib_data_clean %>% filter(cruises != d)
  test  <- c3_calib_data_clean %>% filter(cruises == d)
  
  
  test$fold <- d
  
  cv_date_results <- bind_rows(cv_cruise_results, test)
}


#Overall performance
rmse(cv_date_results$ext_chl, cv_date_results$pred_ext_chl)


#Perfomance by cruise dates     

cv_cruise_results %>%
  group_by(fold) %>%
  summarise(
    RMSE = rmse(ext_chl, pred_ext_chl),
    n = n()
  )


#Combine both ideas

cv_date_results %>%
  mutate(chl_group = ifelse(chl <= median(chl, na.rm = TRUE), "Low", "High")) %>%
  group_by(chl_group) %>%
  summarise(RMSE = rmse(ext_chl, pred_ext_chl))

```


```{r}


library(dplyr)

sites <- unique(c3_calib_data$code)

cv_date_results <- data.frame()

for (d in sites) {
  
  train <- c3_calib_data_clean %>% filter(sites != d)
  test  <- c3_calib_data_clean %>% filter(sites == d)
  
  model <- lm(log10(ext_chl) ~ log10(chl), data = train)
  
  # Predict
  test$pred_log <- predict(model, newdata = test)
  
  # Bias correction
  sigma2 <- summary(model)$sigma^2
  test$pred_ext_chl <- 10^(test$pred_log) * exp(sigma2 / 2)
  
  test$fold <- d
  
  cv_date_results <- bind_rows(cv_date_results, test)
}

#General performance

rmse(cv_date_results$ext_chl, cv_date_results$pred_ext_chl)

#performace by date

cv_date_results %>%
  group_by(fold) %>%
  summarise(
    RMSE = rmse(ext_chl, pred_ext_chl),
    n = n()
  )


#Combine both ideas

cv_date_results %>%
  mutate(chl_group = ifelse(chl <= median(chl, na.rm = TRUE), "Low", "High")) %>%
  group_by(chl_group) %>%
  summarise(RMSE = rmse(ext_chl, pred_ext_chl))




```




