---
title: "cruise_data"
author: "Angela Nankabirwa"
date: "2026-03-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


#loading packages

```{r}

library(tidyverse)
library(janitor)
library(stringi)  
library(lubridate)
library(readr)
library(knitr)
library(readxl)
library(ggpattern) 
library(gridExtra)
library(kableExtra) #formatting e,g. colour)
library(zoo) #remove noisy bumps from data #rolling median
library(specleanr) #20 ways to detect outliers
library(tidytext) #visualize outliers from specleanr package
library(ggpubr) ##adding equations to the plot
library(purrr) #to read csv files with map_dfr 
library(hms) #convert time to proper hms
library(mgcv) #For GAM models



```






#import all cruise ysi files

+ Function to import ysi files
```{r}

#function to read and combine csv files
read_and_combine_csv <- function(file_path) {
  tryCatch(
    {
      # Read the CSV file
      data <- read_csv(file_path)
      
      
      # Remove the first row
      data <- data[-1, ]
      
      # Columns to convert to double
      cols_to_convert <- c("Temp", "SpCond", "Sal", "Depth", "Chl...6", "Chl...7", "BGA-PC...8", "BGA-PC...9", "Battery")
      
      # Convert specified columns to double
      data[, cols_to_convert] <- lapply(data[, cols_to_convert], as.numeric)
      
      # Extract the file name without the ".csv" extension
      file_name <- tools::file_path_sans_ext(basename(file_path))
      
      # Add a new column "FileName" to the data frame, remove formats
      data <- data %>% mutate(FileName = file_name)%>% 
        clean_names() #clean headings 
      
      
      return(data)
    },
    error = function(e) {
      cat("Error occurred while processing file:", file_path, "\n")
      message("Error Message:", conditionMessage(e), "\n")
      traceback()
      return(NULL) # Return NULL to indicate failure
    }
  )
}



```

# Loading Lake data fro calibration

+   2022

+     MFANGANO chl sensor

```{r}


file_paths_csv <- dir("ysi_2022_files", pattern = "*.csv", full.names = TRUE)


ysi_cruz22_files <- purrr::map_dfr(file_paths_csv, read_and_combine_csv) 


ysiF_2022 <- ysi_cruz22_files %>% 
  filter(file_name == "L_CMPF") 


#attach time stamps
ysiF_2022_b <- ysiF_2022 %>%
  mutate(date_time = as.POSIXct(date_time, format = "%m/%d/%Y %H:%M"),
         code = case_when(
           (date_time >= as.POSIXct("2022-07-12 8:20:00") & date_time <= as.POSIXct("2022-07-12 8:40:00")) ~ "BG10",
           (date_time >= as.POSIXct("2022-07-12 10:40:00") & date_time <= as.POSIXct("2022-07-12 11:00:00")) ~ "BG13",
           (date_time >= as.POSIXct("2022-07-12 12:50:00") & date_time <= as.POSIXct("2022-07-12 13:00:00")) ~ "BG16",
           (date_time >= as.POSIXct("2022-07-12 15:40:00") & date_time <= as.POSIXct("2022-07-12 15:50:00")) ~ "CNL1",
           (date_time >= as.POSIXct("2022-07-12 21:00:00") & date_time <= as.POSIXct("2022-07-12 21:20:00")) ~ "BG18",
           TRUE ~ NA_character_
         )) %>% 
  filter(!is.na(code))





```

+   checking for MF sonde outliers

```{r}
#The values are very few. They were recorded every 30 minutes and so we will just do average


ysiF_22_outlier_detect <- multidetect(data = ysiF_2022_b, 
                                var = "chl_rfu_F", 
                                var_col = "code",
                                multiple = TRUE,
                                methods = c("adjbox", "iqr", "hampel","jknife",
                                            "seqfences", "mixediqr",
                                            "distboxplot", "semiqr",
                                            "zscore", "logboxplot", "medianrule"),
                       silence_true_errors = TRUE, 
                       missingness = 0.1,
                       sdm = FALSE,
                       na.inform = TRUE)
  
  

#Visualize the number of outliers detected by each method

ggoutliers(ysiF_22_outlier_detect, select = unique(ysiF_2022_b$code),
           color = "lightblue",
           ncol = 5,
           nrow = 1)



ggsave(file = "plots/2022_plots/2022MF_outlier_per_site23.png", width = 10, height = 5, dpi = 300)


#Obtaining quality controlled dataset using loess method or data labeling


out_qc_loess22F <- extract_clean_data(refdata = ysiF_2022_b, 
            outliers = ysiF_22_outlier_detect, loess = TRUE,
            var_col = "code",
            threshold = 0.7)  #set threshold to 0.8 to remove the very strong outliers only

#clean dataset
nrow(out_qc_loess22F)

#reference data
nrow(ysiF_2022_b)

out_qc_labeled22F <- classify_data(refdata = ysiF_2022_b, 
                                 var_col = "code", 
                                outliers = ysiF_22_outlier_detect)

#Visualize labelled quality controlled dataset

ggenvironmentalspace(out_qc_labeled22F, 
                     type = '1D',
                     ggxangle = 45, 
                     scalecolor = 'viridis',
                     xhjust = 1,
                     legend_position = 'blank',
                     ylab = "Number of records",
                     xlab = "Outlier labels",
                     ncol = 5,
                     nrow = 1)

ggsave(filename = "plots/2022_plots/labelled_outlier_chart_MF.png", width = 8, height = 5, dpi = 300)


#labelled outlier or not 
labeled22F <- classify_data(refdata = ysiF_2022_b, 
                           var_col = "code",
                           outliers = ysiF_22_outlier_detect)



#plot clean dataset

out_qc_loess22F %>% 
mutate(code = as.factor(code)) %>% 
ggplot(aes(x = chl_rfu_F)) +
  geom_histogram(bins=5,binwidth = 1, boundary = 0, fill = "steelblue", color = "black") +
facet_wrap(~ code, scales = "free_y")+
  labs(x = "Chlorophyll (RFU)", y = "Frequency") +
  theme_bw() +
  theme(text = element_text(size = 20, colour = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))+
 scale_x_continuous(
  breaks = scales::pretty_breaks(n = 6)  #ggplot makes breaks based on data
)

ggsave(file="plots/2022_plots/ysi22_loessfilter_MF_freq_distribution.png", width = 10, height = 14, dpi = 300)



#No strong outlier so we will just average the values

ysiF_2022_c <- ysiF_2022_b %>% 
  rename(chl_rfu_F = chl_7,
         bga_rfu_F = bga_pc_9) %>% 
  
  filter(!is.na(code)) %>% 
  group_by(code) %>% 
  summarise(
    chl_ysi = mean(chl_rfu_F, na.rm = TRUE),   
    bga_ysi = mean(bga_rfu_F, na.rm = TRUE),
    temp_ysi = mean(temp, na.rm = TRUE),
    sp_cond_ysi = mean(sp_cond, na.rm = TRUE)
  ) %>% 
  ungroup() %>% 
  mutate(sample_type = "cruise-2022_MFysi")

```


+   Dunga chl sensor

```{r}
#DUNGA sonde

ysi_2022DG <-ysi_cruz22_files %>% 
  filter(file_name == "L_CMPDU") %>% 
   rename(chl_rfu_DG = chl_7,
         bga_rfu_DG = bga_pc_9) 
    
    
  
ysi_2022DG_b <- ysi_2022DG %>%
  mutate(date_time = as.POSIXct(date_time, format = "%m/%d/%Y %H:%M"),
         code = case_when(
           (date_time >= as.POSIXct("2022-07-12 8:20:00") & date_time <= as.POSIXct("2022-07-12 8:40:00")) ~ "BG10",
           (date_time >= as.POSIXct("2022-07-12 10:50:00") & date_time <= as.POSIXct("2022-07-12 11:00:00")) ~ "BG13",
           (date_time >= as.POSIXct("2022-07-12 12:50:00") & date_time <= as.POSIXct("2022-07-12 13:00:00")) ~ "BG16",
           (date_time >= as.POSIXct("2022-07-12 15:40:00") & date_time <= as.POSIXct("2022-07-12 15:50:00")) ~ "CNL1",
           (date_time >= as.POSIXct("2022-07-12 21:00:00") & date_time <= as.POSIXct("2022-07-12 21:20:00")) ~ "BG18",
           TRUE ~ NA_character_
         )) %>% 
  filter(!is.na(code))


#for outliers, same situation as MF sonde, so we will just do the average

ysi_2022DG_c <- ysi_2022DG_b %>% 
  filter(!is.na(code)) %>% 
group_by(code) %>% 
  summarise(
    chl_ysi = mean(chl_rfu_DG, na.rm = TRUE),
    bga_ysi = mean(bga_rfu_DG, na.rm= TRUE),
    temp_ysi = mean(temp, na.rm=TRUE),
    sp_cond_ysi = mean(sp_cond, na.rm = TRUE)
    ) %>% 
  ungroup() %>% 
  mutate(sample_type = "cruise-2022_DGysi")


```

+     combine MF and DG sonde data

```{r}
####bind DG and MF sonde cruise

cruise_sondes_2022 <- bind_rows(ysi_2022DG_c, ysiF_2022_c) 



```



+ import files for 2023 and 2024

+   MF chl sensor

```{r}


#YSI files

# Get a list of file paths in the "files" directory

file_paths.csv <- dir("ysi_cruise_files", pattern = "*.csv", full.names = TRUE)


# Read all csv files and combine them
 MF_cruz_22 <- map_df(file_paths.csv, read_and_combine_csv) %>% 
    rename(chl_rfu = chl_7,
         bga_rfu = bga_pc_9) 
 


```

+   attach time stamps to 2023 cruz

+   **2023**
```{r}



ysi2023 <- MF_cruz_22 %>%
  filter(file_name == "cruise-ysidata-2023_naphtali") %>% 
  
  mutate(date_time = as.POSIXct(date_time, format = "%m/%d/%Y %H:%M"),
         code = case_when(
           (date_time >= as.POSIXct("2023-06-27 11:26:00") & date_time <= as.POSIXct("2023-06-27 11:44:00")) ~ "CHL1",
           (date_time >= as.POSIXct("2023-06-27 11:46:00") & date_time <= as.POSIXct("2023-06-27 12:49:00")) ~ "CHL2",
           (date_time >= as.POSIXct("2023-06-27 12:50:00") & date_time <= as.POSIXct("2023-06-27 14:10:00")) ~ "BG7",
           (date_time >= as.POSIXct("2023-06-27 14:11:00") & date_time <= as.POSIXct("2023-06-27 14:40:00")) ~ "CHL3",
           (date_time >= as.POSIXct("2023-06-27 14:41:00") & date_time <= as.POSIXct("2023-06-27 15:01:00")) ~ "CHL4",
           (date_time >= as.POSIXct("2023-06-27 15:02:00") & date_time <= as.POSIXct("2023-06-27 15:20:00")) ~ "CHL5",
           (date_time >= as.POSIXct("2023-06-27 15:22:00") & date_time <= as.POSIXct("2023-06-27 15:58:00")) ~ "CHL6",
           (date_time >= as.POSIXct("2023-06-27 15:59:00") & date_time <= as.POSIXct("2023-06-27 16:28:00")) ~ "CHL7",
           (date_time >= as.POSIXct("2023-06-27 16:29:00") & date_time <= as.POSIXct("2023-06-27 18:08:00")) ~ "CHL8",
           (date_time >= as.POSIXct("2023-06-27 18:10:00") & date_time <= as.POSIXct("2023-06-27 19:16:00")) ~ "CHL9",
           (date_time >= as.POSIXct("2023-06-27 19:17:00") & date_time <= as.POSIXct("2023-06-27 20:08:00")) ~ "CHL10",
           (date_time >= as.POSIXct("2023-06-28 7:52:00")  & date_time <= as.POSIXct("2023-06-28 8:35:00"))  ~ "CHL11",
           (date_time >= as.POSIXct("2023-06-28 8:37:00")  & date_time <= as.POSIXct("2023-06-28 9:38:00")) ~ "CHL12",
           (date_time >= as.POSIXct("2023-06-28 9:40:00")  & date_time <= as.POSIXct("2023-06-28 10:05:00")) ~ "CHL13",
           (date_time >= as.POSIXct("2023-06-28 10:07:00") & date_time <= as.POSIXct("2023-06-28 11:28:00")) ~ "CHL14",
           (date_time >= as.POSIXct("2023-06-28 11:29:00") & date_time <= as.POSIXct("2023-06-28 12:26:00")) ~ "CHL15",
           (date_time >= as.POSIXct("2023-06-28 12:28:00") & date_time <= as.POSIXct("2023-06-28 13:46:00")) ~ "CHL16",
           (date_time >= as.POSIXct("2023-06-28 13:47:00") & date_time <= as.POSIXct("2023-06-28 15:01:00")) ~ "CHL17",
           (date_time >= as.POSIXct("2023-06-28 15:02:00") & date_time <= as.POSIXct("2023-06-28 15:34:00")) ~ "CHL18",
           (date_time >= as.POSIXct("2023-06-28 15:35:00") & date_time <= as.POSIXct("2023-06-28 16:16:00")) ~ "CHL19",
           (date_time >= as.POSIXct("2023-06-28 16:17:00") & date_time <= as.POSIXct("2023-06-28 17:04:00")) ~ "CHL20",
           (date_time >= as.POSIXct("2023-06-28 17:05:00") & date_time <= as.POSIXct("2023-06-28 17:19:00")) ~ "CHL21",
           (date_time >= as.POSIXct("2023-06-28 17:20:00") & date_time <= as.POSIXct("2023-06-28 18:26:00")) ~ "CHL22",
           (date_time >= as.POSIXct("2023-06-28 18:28:00") & date_time <= as.POSIXct("2023-06-28 20:05:00")) ~ "CHL23",
           (date_time >= as.POSIXct("2023-06-28 20:10:00") & date_time <= as.POSIXct("2023-06-28 20:50:00")) ~ "CHL24",
           (date_time >= as.POSIXct("2023-06-28 20:52:00") & date_time <= as.POSIXct("2023-06-28 21:22:00")) ~ "CHL25",
           (date_time >= as.POSIXct("2023-06-28 21:23:00") & date_time <= as.POSIXct("2023-06-28 21:55:00")) ~ "CHL26",
           
           TRUE ~ NA_character_
         )) %>% 
  filter(!is.na(code)) %>% 
  mutate(
    date_time = as.character(date_time),
    date_time_parsed = parse_date_time(
      date_time,
      orders = c(
  "Ymd HMS",
  "Ymd HM",
  "dmy HMS",
  "dmy HM",
  "dmY HMS",
  "mdy HMS",
  "mdy HM"
))) %>% 
  mutate(year= year(date_time_parsed),
         month = month(date_time_parsed),
         mmyy = format(date_time_parsed, "%m-%y"))

  


```

+   2023 chl frequency distribution

```{r}


#No manual filter

ysi2023 %>% 
  mutate(code = as.factor(code)) %>% 
ggplot(aes(x = chl_rfu)) +
  geom_histogram(bins = 30, binwidth = 2, boundary = 0, fill = "steelblue", color = "black") +
  #facet_wrap(~ code, scales = "free_y") +
  labs(x = "Chlorophyll (RFU)", y = "Frequency") +
  theme_bw() +
  theme(text = element_text(size = 30, colour = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))+
  scale_x_continuous(
    breaks = seq(0, max(ysi2023$chl_rfu, na.rm = TRUE), by = 2)
  ) 

ggsave(file="plots/2023_plots/ysi2023_gen_freq_distribution.png", width = 10, height = 12, dpi = 300)




#Make plots for separate sites

ysi2023 %>% 
  mutate(code = as.factor(code)) %>% 
ggplot(aes(x = chl_rfu)) +
  geom_histogram(bins = 30, binwidth=2, boundary=0, fill = "steelblue", color = "black") +
  facet_wrap(~ code) +
  labs(x = "Chlorophyll (RFU)", y = "Frequency") +
   theme(text = element_text(size = 30, colour = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))+
  theme_bw() +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 4))



ggsave(file="plots/2023_plots/ysi2023_site_plots.png",width = 16, height = 12, dpi = 300)




```

+   2023 filter for outliers

```{r}

ysicrjul_23_outlier_detect <- multidetect(data = ysi2023, 
                                var = "chl_rfu", 
                                var_col = "code",
                                multiple = TRUE,
                                methods = c("adjbox", "iqr", "hampel","jknife",
                                            "seqfences", "mixediqr",
                                            "distboxplot", "semiqr",
                                            "zscore", "logboxplot", "medianrule"),
                       silence_true_errors = TRUE, 
                       missingness = 0.1,
                       sdm = FALSE,
                       na.inform = TRUE)
  
  

#Visualize the number of outliers detected by each method

ggoutliers(ysicrjul_23_outlier_detect, select = unique(ysi2023$code),
           color = "lightblue",
           ncol = 5,
           nrow = 7)



#outlier_per_site24 <- ggoutliers(ysicrjul_23_outlier_detect,
                       # select = c("CHL26", "CHL25", "CHL24","CHL3","CHL4", "CHL5","CHL6", #"CHL7", "BG10","CHL9","CHL10","BG13", "CHL19","CHL18","CHL15","CNL1","CNL2",  "CHL12", "CHL13","BG16","CHL20","CHL21","CHL27","BG5"),
                       # color='purple',
                       # desc = TRUE,
                       # ncol = 5,
                       # nrow = 7)

ggsave(file = "plots/2023_plots/2023_outliersoutlier_per_site23.png", width = 10, height = 20, dpi = 300)


#Obtaining quality controlled dataset using loess method or data labeling


out_qc_loess23 <- extract_clean_data(refdata = ysi2023, 
            outliers = ysicrjul_23_outlier_detect, loess = TRUE,
            var_col = "code",
            threshold = 0.7)  #set threshold to 0.8 to remove the very strong outliers only

#clean dataset
nrow(out_qc_loess23)

#reference data
nrow(ysi2023)

out_qc_labeled23 <- classify_data(refdata = ysi2023, 
                                 var_col = "code", 
                                outliers = ysicrjul_23_outlier_detect)

#Visualize labelled quality controlled dataset

ggenvironmentalspace(out_qc_labeled23, 
                     type = '1D',
                     ggxangle = 45, 
                     scalecolor = 'viridis',
                     xhjust = 1,
                     legend_position = 'blank',
                     ylab = "Number of records",
                     xlab = "Outlier labels",
                     ncol = 5,
                     nrow = 7)

ggsave(filename = "plots/2023_plots/labelled_outlier_chart_site.png", width = 10, height = 26, dpi = 300)


#labelled outlier or not 
labeled23 <- classify_data(refdata = ysi2023, 
                           var_col = "code",
                           outliers = ysicrjul_23_outlier_detect)



#plot clean dataset

out_qc_loess23 %>% 
mutate(code = as.factor(code)) %>% 
ggplot(aes(x = chl_rfu)) +
  geom_histogram(bins = 30, binwidth = 2, boundary = 0, fill = "steelblue", color = "black") +
facet_wrap(~ code, scales = "free_y")+
  labs(x = "Chlorophyll (RFU)", y = "Frequency") +
  theme_bw() +
  theme(text = element_text(size = 10, colour = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))+
 scale_x_continuous(
  breaks = scales::pretty_breaks(n = 6)  #ggplot makes breaks based on data
)

ggsave(file="plots/2023_plots/ysi2023_loessfilter_sites_freq_distribution.png", width = 10, height = 14, dpi = 300)



```


+   final filtering of points for cruz 2023

```{r}

final23_df <- out_qc_labeled23 %>%
  group_by(code) %>%
  slice_max(order_by = chl_rfu, n = 3) %>%
  summarise(
    chl_ysi = mean(chl_rfu, na.rm = TRUE),
    bga_ysi = mean(bga_rfu, na.rm = TRUE),
    temp_ysi = mean(temp, na.rm = TRUE),
    sp_cond_ysi = mean(sp_cond, na.rm = TRUE),
    file_name  = unique(file_name)
  )
    
    

```


+   **2024**
```{r}


crjul_24 <- MF_cruz_22 %>%
  filter(file_name == "CR24JUL_ysi")
  
##cross check to only keep all depths > -0.7 i.e. -0.6 to -0.4 for lake samples

##adding timestamps to cruise (CR) and Dunga mapping data (they have a different date_time structure from the calibration data)

ysicrjul_24 <- crjul_24 %>%
  filter(file_name != c("CALIB24_ysi")) %>% 
  mutate(date_time = as.POSIXct(date_time, format = "%m/%d/%Y %H:%M"),
         code = case_when(
          
          (date_time >= as.POSIXct("2024-07-18 14:28:00") & date_time <= as.POSIXct("2024-07-18 14:48:00")) ~ "CHL26",
           (date_time >= as.POSIXct("2024-07-18 15:37:00") & date_time <= as.POSIXct("2024-07-18 15:42:00")) ~ "CHL25",
           (date_time >= as.POSIXct("2024-07-18 17:14:00") & date_time <= as.POSIXct("2024-07-18 17:31:00")) ~ "CHL24",
           (date_time >= as.POSIXct("2024-07-18 18:58:00") & date_time <= as.POSIXct("2024-07-18 19:13:00")) ~ "CHL3",
           (date_time >= as.POSIXct("2024-07-18 19:36:00") & date_time <= as.POSIXct("2024-07-18 19:50:00")) ~ "CHL4",
           (date_time >= as.POSIXct("2024-07-18 20:16:00") & date_time <= as.POSIXct("2024-07-18 20:27:00")) ~ "CHL5",
           (date_time >= as.POSIXct("2024-07-18 20:48:00") & date_time <= as.POSIXct("2024-07-18 21:04:00")) ~ "CHL6",
           (date_time >= as.POSIXct("2024-07-18 21:51:00") & date_time <= as.POSIXct("2024-07-18 22:12:00")) ~ "CHL7",
           (date_time >= as.POSIXct("2024-07-19 7:48:00") & date_time <= as.POSIXct("2024-07-19 8:05:00")) ~ "BG10",
           (date_time >= as.POSIXct("2024-07-19 9:23:00") & date_time <= as.POSIXct("2024-07-19 9:31:00")) ~ "CHL9",
           (date_time >= as.POSIXct("2024-07-19 10:46:00") & date_time <= as.POSIXct("2024-07-19 11:01:00")) ~ "CHL10",
           (date_time >= as.POSIXct("2024-07-19 11:31:00") & date_time <= as.POSIXct("2024-07-19 11:56:00"))  ~ "BG13",
           (date_time >= as.POSIXct("2024-07-19 14:23:00") & date_time <= as.POSIXct("2024-07-19 14:48:00")) ~ "CHL19",
           (date_time >= as.POSIXct("2024-07-19 16:02:00") & date_time <= as.POSIXct("2024-07-19 16:14:00")) ~ "CHL18",
          (date_time >= as.POSIXct("2024-07-19 16:53:00") & date_time <= as.POSIXct("2024-07-19 16:59:00")) ~ "CHL15",
           (date_time >= as.POSIXct("2024-07-19 17:52:00") & date_time <= as.POSIXct("2024-07-19 18:09:00")) ~ "CNL1",
          (date_time >= as.POSIXct("2024-07-20 8:31:00") & date_time <= as.POSIXct("2024-07-20 8:48:00")) ~ "CNL2",
          (date_time >= as.POSIXct("2024-07-20 9:52:00") & date_time <= as.POSIXct("2024-07-20 10:05:00")) ~ "CHL12",
           (date_time >= as.POSIXct("2024-07-20 10:46:00") & date_time <= as.POSIXct("2024-07-20 10:57:00")) ~ "CHL13",
           (date_time >= as.POSIXct("2024-07-20 12:01:00") & date_time <= as.POSIXct("2024-07-20 12:14:00")) ~ "BG16",
           (date_time >= as.POSIXct("2024-07-20 14:29:00") & date_time <= as.POSIXct("2024-07-20 14:39:00")) ~ "CHL20",
           (date_time >= as.POSIXct("2024-07-20 15:34:00") & date_time <= as.POSIXct("2024-07-20 15:43:00")) ~ "CHL21",
           (date_time >= as.POSIXct("2024-07-20 18:18:00") & date_time <= as.POSIXct("2024-07-20 18:23:00")) ~ "CHL27",
          (date_time >= as.POSIXct("2024-07-20 19:06:00") & date_time <= as.POSIXct("2024-07-20 19:13:00")) ~ "BG5",
          
              TRUE ~ NA_character_
         )) %>% 
  filter(!is.na(code)) %>% 
  mutate(
    date_time = as.character(date_time),
    date_time_parsed = parse_date_time(
      date_time,
      orders = c(
  "Ymd HMS",
  "Ymd HM",
  "dmy HMS",
  "dmy HM",
  "dmY HMS",
  "mdy HMS",
  "mdy HM"
))) %>% 
  mutate(year= year(date_time_parsed),
         month = month(date_time_parsed),
         mmyy = format(date_time_parsed, "%m-%y")) 



```


+     Frequency distribution for July 2024 cruz
```{r}



#No manual filter

ysicrjul_24 %>% 
  mutate(code = as.factor(code)) %>% 
ggplot(aes(x = chl_rfu)) +
  geom_histogram(bins = 30, binwidth = 2, boundary = 0, fill = "steelblue", color = "black") +
  #facet_wrap(~ code, scales = "free_y") +
  labs(x = "Chlorophyll (RFU)", y = "Frequency") +
  theme_bw() +
  theme(text = element_text(size = 30, colour = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))+
 scale_x_continuous(
  breaks = scales::pretty_breaks(n = 6)  #ggplot makes breaks based on data
)

ggsave(file="plots/2024_plots/ysi2024_gen_freq_distribution.png", width = 10, height = 12, dpi = 300)




#Make plots for separate sites

ysicrjul_24 %>% 
  mutate(code = as.factor(code)) %>% 
ggplot(aes(x = chl_rfu)) +
  geom_histogram(bins = 30, binwidth=2, boundary=0, fill = "steelblue", color = "black") +
  facet_wrap(~ code, scales = "free_y") +
  labs(x = "Chlorophyll (RFU)", y = "Frequency") +
   theme(text = element_text(size = 30, colour = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))+
  theme_bw() +
  scale_x_continuous(
  breaks = scales::pretty_breaks(n = 6)
)
  

ggsave(file="plots/2024_plots/ysi2024_site_freq_distribution.png",width = 16, height = 12, dpi = 300)


#this data might a manual filter to remove values chl >150






```

+  try outlier filters for July 2024 data

+   manual removal based on error readings

```{r}

ysicrjul_24_filter <- ysicrjul_24 %>% 
  filter(!(chl_rfu <= 0)) 

ysicrjul_24_filter %>% 
mutate(code = as.factor(code)) %>% 
ggplot(aes(x = chl_rfu)) +
  geom_histogram(bins = 30, binwidth = 2, boundary = 0, fill = "steelblue", color = "black") +
facet_wrap(~ code, scales = "free_y") +
  labs(x = "Chlorophyll (RFU)", y = "Frequency") +
  theme_bw() +
  theme(text = element_text(size = 10, colour = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))+
 scale_x_continuous(
  breaks = scales::pretty_breaks(n = 6)  #ggplot makes breaks based on data
)

ggsave(file="plots/2024_plots/ysi2024_manualfilter_sites_freq_distribution.png", width = 10, height = 12, dpi = 300)



```

+   outlier removal using specleanr

```{r}

ysicrjul_24_outlier_detect <- multidetect(data = ysicrjul_24_filter, 
                                var = "chl_rfu", 
                                var_col = "code",
                                multiple = TRUE,
                                methods = c("adjbox", "iqr", "hampel","jknife",
                                            "seqfences", "mixediqr",
                                            "distboxplot", "semiqr",
                                            "zscore", "logboxplot", "medianrule"),
                       silence_true_errors = TRUE, 
                       missingness = 0.1,
                       sdm = FALSE,
                       na.inform = TRUE)
  
  

#Visualize the number of outliers detected by each method

ggoutliers(ysicrjul_24_outlier_detect,
                        select = c("CHL26", "CHL25", "CHL24","CHL3","CHL4", "CHL5","CHL6", "CHL7", "BG10","CHL9","CHL10","BG13", "CHL19","CHL18","CHL15","CNL1","CNL2",  "CHL12", "CHL13","BG16","CHL20","CHL21","CHL27","BG5"),
                        color='purple',
                        desc = TRUE,
                        ncol = 10,
                        nrow = 10)

ggsave(file = "plots/2024_plots/2024_outliers_per_site24_no_chl18_.png", width = 10, height = 20, dpi = 300)


#Obtaining quality controlled dataset using loess method or data labeling


out_qc_loess24 <- extract_clean_data(refdata = ysicrjul_24_filter, 
            outliers = ysicrjul_24_outlier_detect, loess = TRUE,
            var_col = "code",
            threshold = 0.7)  #set threshold to 0.8 to remove the very strong outliers only

#clean dataset
nrow(out_qc_loess24)

#reference data
nrow(ysicrjul_24)

out_qc_labeled24 <- classify_data(refdata = ysicrjul_24_filter, 
                                  var_col = "code",
                                outliers = ysicrjul_24_outlier_detect)

#Visualize labelled quality controlled dataset

ggenvironmentalspace(out_qc_labeled24, 
                     type = '1D',
                     ggxangle = 45, 
                     ncol = 5,
                     nrow = 12,
                     scalecolor = 'viridis',
                     xhjust = 1,
                     legend_position = 'blank',
                     ylab = "Number of records",
                     xlab = "Outlier labels")


ggsave(filename = "plots/2024_plots/filtered_site_data.png", width = 14, height = 14, dpi = 300)



#labelled outlier or not 
labeled24 <- classify_data(refdata = ysicrjul_24_filter, 
                         var_col = "code",
                         outliers = ysicrjul_24_outlier_detect)



#plot clean dataset

out_qc_loess24 %>% 
mutate(code = as.factor(code)) %>% 
ggplot(aes(x = chl_rfu)) +
  geom_histogram(bins = 30, binwidth = 2, boundary = 0, fill = "steelblue", color = "black") +
facet_wrap(~ code, scales = "free_y")+
  labs(x = "Chlorophyll (RFU)", y = "Frequency") +
  theme_bw() +
  theme(text = element_text(size = 10, colour = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))+
 scale_x_continuous(
  breaks = scales::pretty_breaks(n = 6)  #ggplot makes breaks based on data
)

ggsave(file="plots/2024_plots/ysi2024_loessfilter_Sites_freq_distribution.png", width = 10, height = 12, dpi = 300)





```

+   final filtering of top points for ysi 
```{r}


final24_df <- out_qc_labeled24 %>%
  group_by(code, year) %>%
  slice_max(order_by = chl_rfu, n = 3) %>%
  summarise(
    chl_ysi = mean(chl_rfu, na.rm = TRUE),
    bga_ysi = mean(bga_rfu, na.rm = TRUE),
    temp_ysi = mean(temp, na.rm = TRUE),
    sp_cond_ysi = mean(sp_cond, na.rm = TRUE),
    file_name  = unique(file_name),
  ) 


```


+   Extracted chlorophyll files


+   2023


```{r}


file_paths.csv <- dir("ysi_extracted_chls", pattern = "*.csv", full.names = TRUE)


chl_df_23 <- read_csv(file_paths.csv[2]) %>% 
  clean_names() %>% 
  rename(code =site)

cr_jul23_merge <- full_join(chl_df_23, final23_df)

cr_jul23_merge_subset <- cr_jul23_merge %>% 
  dplyr::select(code, ext_chl, chl_ysi, bga_ysi, sp_cond_ysi, temp_ysi, file_name)

```

+   2024
```{r}


chl_dat_24 <- read_csv(file_paths.csv[1]) %>% 
  clean_names() %>% 
  mutate(dilution_ratio = as.character(dilution_ratio)) %>% 
  rename(index_code = chl_index)


# Define the function to compute the dilution factor (DF)
df_func <- function(total_vol, sample_vol) {
  DF = total_vol / sample_vol
  return(DF)
}

# Apply the function to the defined columns to get DF
chl_dat_24$DF <- df_func(chl_dat_24$total_vol_in_cuvette, chl_dat_24$sample_vol)


#convert final_FU to chl concentration using Pete's formula (for formula refer to WG_2022 analysis R file)

#Chl a (ug/l) = 0.0003*F*dilution factor*extract volume (mL) / volume filtered (L)
#where dilution factor is whether you needed to dilute in order to get a reading.

###Make a function to convert florescence units to chlorophyll concentration
convert_fluorescence_to_chlorophyll <- function(fluoresence_units, dilution_factor, extract_volume, volume_filtered) {
 ext_chl <- 0.0003 * fluoresence_units * dilution_factor * extract_volume / volume_filtered
 return(ext_chl)
}


fluoresence_units <- chl_dat_24$fu_after_dilution
dilution_factor  <- chl_dat_24$DF
extract_volume <- chl_dat_24$vol_extract
volume_filtered <- chl_dat_24$vol_filtered/1000


chl_dat_24$ext_chl <- convert_fluorescence_to_chlorophyll(fluoresence_units, dilution_factor, extract_volume, volume_filtered)


chl_dat_subset <- chl_dat_24[-36, ]

chl_df_24 <- chl_dat_subset %>% 
  select(index_code,site, ext_chl, sample_type) %>% 
  filter_out(sample_type == "expt-jul_24") %>% 
  rename(code = site)
 


```


+ 2022

```{r}


data_2022 <- read_csv(file_paths.csv[4])

#this file has 2022 cruise ext_chl


data_2022 <- data_2022 %>% 
  rename(date_time = date) 

data_2022 <- data_2022%>% 
  mutate(date = date(date_time),
         time = paste(hour(date_time), minute(date_time), second(date_time), 
                      sep = ":"))  %>% # Extract time components and combine)  
mutate(year = year(date))

cruise_sites_22 <- data_2022 %>% 
  filter( year == "2022" ) %>% 
  dplyr::select(station, fu_chl) %>%
  rename(code = station,
         ext_chl = fu_chl) %>% 
  filter(code %in% c("BG10", "BG13", "BG16", "CNL1", "BG18")) 
  
#cruise_sites_22 contains extracted chl for matching the 2022 cruise_compar sondes

ysidata_2022 <- full_join(cruise_sondes_2022, cruise_sites_22, by = "code") %>% 
  rename(file_name = sample_type)



```


+   Join 2022, 2023 and 2024 cruz datasets

```{r}


#join ext_chl with july 2024 ysi data

cr_jul24_merge <- full_join(final24_df, chl_df_24, by = "code")
     

#join ext_chl 2023 cruz with july 2023 ysi data


cr_jul24_merge_subset <- cr_jul24_merge %>% 
  dplyr::select(code, ext_chl, chl_ysi, bga_ysi, sp_cond_ysi, temp_ysi, file_name)




ysi_field_data <- bind_rows(cr_jul23_merge_subset, cr_jul24_merge_subset, ysidata_2022)  %>% 
  mutate(data_type = "field")



```


# Experiments data

+   2022 experiments

+   MF sensor
```{r}
#MF sonde

file_paths_csv <- dir("ysi_experiments", pattern = "*.csv", full.names = TRUE)


ysi_expt_files <- purrr::map_dfr(file_paths_csv, read_and_combine_csv) 



MF_EXPER <- ysi_expt_files %>% 
  filter(file_name == "MF_EXPER_2022") 

MF_EXPER_b <- MF_EXPER %>%
  mutate(date_time = as.POSIXct(date_time, format = "%m/%d/%Y %H:%M"),
         code = case_when(
           (date_time >= as.POSIXct("2022-07-10 12:01:00") & date_time <= as.POSIXct("2022-07-10 13:01:00")) ~ "X",
           (date_time >= as.POSIXct("2022-07-10 13:11:00") & date_time <= as.POSIXct("2022-07-10 14:21:00")) ~ "0.5X",
           (date_time >= as.POSIXct("2022-07-10 14:31:00") & date_time <= as.POSIXct("2022-07-10 15:21:00")) ~ "0.25X",
           (date_time >= as.POSIXct("2022-07-10 15:31:00") & date_time <= as.POSIXct("2022-07-10 16:21:00")) ~ "0.125X",
           (date_time >= as.POSIXct("2022-07-10 16:31:00") & date_time <= as.POSIXct("2022-07-10 17:21:00")) ~ "0.062X",
           (date_time >= as.POSIXct("2022-07-10 17:31:00") & date_time <= as.POSIXct("2022-07-10 18:21:00")) ~ "0.031X",
           (date_time >= as.POSIXct("2022-07-10 18:31:00") & date_time <= as.POSIXct("2022-07-11 19:21:00")) ~ "0.015X",
           (date_time >= as.POSIXct("2022-07-11 19:31:00") & date_time <= as.POSIXct("2022-07-11 20:21:00")) ~ "0.007X",
           TRUE ~ NA_character_
         )) %>% 
  filter(!is.na(code)) %>% 
  rename(chl_rfu_F = chl_7,
         bga_rfu_F = bga_pc_9)



```

+     remove outliers_MF sonde
```{r}

ysiF_expt_outlier_detect <- multidetect(data = MF_EXPER_b, 
                                var = "chl_rfu_F", 
                                var_col = "code",
                                multiple = TRUE,
                                methods = c("adjbox", "iqr", "hampel","jknife",
                                            "seqfences", "mixediqr",
                                            "distboxplot", "semiqr",
                                            "zscore", "logboxplot", "medianrule"),
                       silence_true_errors = TRUE, 
                       missingness = 0.1,
                       sdm = FALSE,
                       na.inform = TRUE)
  
  

#Visualize the number of outliers detected by each method

ggoutliers(ysiF_expt_outlier_detect, select = unique(MF_EXPER_b$code),
           color = "lightblue",
           ncol = 5,
           nrow = 10)



ggsave(file = "plots/2022_plots/expts/2022MFsonde_Expt_outlier_graph.png", width = 10, height = 5, dpi = 300)


#Obtaining quality controlled dataset using loess method or data labeling


out_qc_loess22F <- extract_clean_data(refdata = MF_EXPER_b, 
            outliers = ysiF_expt_outlier_detect, loess = TRUE,
            var_col = "code",
            threshold = 0.7)  #set threshold to 0.8 to remove the very strong outliers only

#clean dataset
nrow(MF_EXPER_b)

#reference data
nrow(ysiF_2022_b)

out_qc_labeled22F <- classify_data(refdata = MF_EXPER_b, 
                                 var_col = "code", 
                                outliers = ysiF_expt_outlier_detect)

#Visualize labelled quality controlled dataset

ggenvironmentalspace(out_qc_labeled22F, 
                     type = '1D',
                     ggxangle = 45, 
                     scalecolor = 'viridis',
                     xhjust = 1,
                     legend_position = 'blank',
                     ylab = "Number of records",
                     xlab = "Outlier labels",
                     ncol = 5,
                     nrow = 2)  #investigate why it's not working#############################

ggsave(filename = "plots/2022_plots/labelled_outlier_chart_MFexpt22.png", width = 8, height = 5, dpi = 300)


#labelled outlier or not 
labeled22F <- classify_data(refdata = MF_EXPER_b, 
                           var_col = "code",
                           outliers = ysiF_expt_outlier_detect)



#plot clean dataset

out_qc_loess22F %>% 
mutate(code = as.factor(code)) %>% 
ggplot(aes(x = chl_rfu_F)) +
  geom_histogram(bins=5,binwidth = 1, boundary = 0, fill = "steelblue", color = "black") +
#facet_wrap(~ code, scales = "free_y")+
  labs(x = "Chlorophyll (RFU)", y = "Frequency") +
  theme_bw() +
  theme(text = element_text(size = 20, colour = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))+
 scale_x_continuous(
  breaks = scales::pretty_breaks(n = 6)  #ggplot makes breaks based on data
)

ggsave(file="plots/2022_plots/expts/MF_expt_loessfilter_general_freq_distribution.png", width = 10, height = 14, dpi = 300)


#final dataset

MF_EXPER_c <- out_qc_loess22F %>% filter(!is.na(code)) %>% 
  group_by(code) %>%
  slice_max(order_by = chl_rfu_F, n = 3) %>%
  summarise(
    chl_ysi = mean(chl_rfu_F, na.rm = TRUE),
    bga_ysi = mean(bga_rfu_F, na.rm= TRUE),
    temp_ysi = mean(temp, na.rm=TRUE),
    sp_cond_ysi = mean(sp_cond, na.rm = TRUE)
    ) %>% 
  ungroup() %>% 
  mutate(file_name = "MFysi-2022_expt")

```


+     DG sensor
```{r}


DU_EXPER <- ysi_expt_files %>% 
  filter(file_name == "DU_EXPER_2022") 


DU_EXPER_b <- DU_EXPER %>%
  mutate(date_time = as.POSIXct(date_time, format = "%m/%d/%Y %H:%M"),
         code = case_when(
           (date_time >= as.POSIXct("2022-07-10 12:00:00") & date_time <= as.POSIXct("2022-07-10 13:00:00")) ~ "X",
           (date_time >= as.POSIXct("2022-07-10 13:10:00") & date_time <= as.POSIXct("2022-07-10 14:20:00")) ~ "0.5X",
           (date_time >= as.POSIXct("2022-07-10 14:30:00") & date_time <= as.POSIXct("2022-07-10 15:20:00")) ~ "0.25X",
           (date_time >= as.POSIXct("2022-07-10 15:30:00") & date_time <= as.POSIXct("2022-07-10 16:20:00")) ~ "0.125X",
           (date_time >= as.POSIXct("2022-07-10 16:30:00") & date_time <= as.POSIXct("2022-07-10 17:21:00")) ~ "0.062X",
           (date_time >= as.POSIXct("2022-07-10 17:31:00") & date_time <= as.POSIXct("2022-07-10 18:21:00")) ~ "0.031X",
           (date_time >= as.POSIXct("2022-07-10 18:31:00") & date_time <= as.POSIXct("2022-07-11 19:21:00")) ~ "0.015X",
           (date_time >= as.POSIXct("2022-07-11 19:30:00") & date_time <= as.POSIXct("2022-07-11 20:40:00")) ~ "0.007X",
           TRUE ~ NA_character_
         )) %>% 
  rename(chl_rfu_DG = chl_7,
         bga_rfu_DG = bga_pc_9)





```


+ remove outliers from the DG sonde and make final 2022 expt data

```{r}


DG22_expt_outlier_detect <- multidetect(data = DU_EXPER_b, 
                                var = "chl_rfu_DG", 
                                var_col = "code",
                                multiple = TRUE,
                                methods = c("adjbox", "iqr", "hampel","jknife",
                                            "seqfences", "mixediqr",
                                            "distboxplot", "semiqr",
                                            "zscore", "logboxplot", "medianrule"),
                       silence_true_errors = TRUE, 
                       missingness = 0.1,
                       sdm = FALSE,
                       na.inform = TRUE)
  
  

#Visualize the number of outliers detected by each method

ggoutliers(DG22_expt_outlier_detect, select = unique(DU_EXPER_b$code),
           color = "lightblue",
           ncol = 5,
           nrow = 10)



ggsave(file = "plots/2022_plots/expts/2022MFsonde_Expt_outlier_graph.png", width = 10, height = 5, dpi = 300)


#Obtaining quality controlled dataset using loess method or data labeling


out_qc_loess22DG <- extract_clean_data(refdata = DU_EXPER_b, 
            outliers = DG22_expt_outlier_detect, loess = TRUE,
            var_col = "code",
            threshold = 0.8)  #set threshold to 0.8 to remove the very strong outliers only

#clean dataset
nrow(out_qc_loess22DG)

#reference data
nrow(DU_EXPER_b)

out_qc_labeled22DU <- classify_data(refdata = DU_EXPER_b, 
                                 var_col = "code", 
                                outliers = DG22_expt_outlier_detect)

#Visualize labelled quality controlled dataset

ggenvironmentalspace(out_qc_labeled22DU, 
                     type = '1D',
                     ggxangle = 45, 
                     scalecolor = 'viridis',
                     xhjust = 1,
                     legend_position = 'blank',
                     ylab = "Number of records",
                     xlab = "Outlier labels",
                     ncol = 5,
                     nrow = 3)  #investigate why it's not working#############################

#ggsave(filename = "plots/2022_plots/labelled_outlier_chart_MF.png", width = 8, height = 5, dpi = 300)


#labelled outlier or not 
labeled22DG <- classify_data(refdata = DU_EXPER_b, 
                           var_col = "code",
                           outliers = DG22_expt_outlier_detect)



#plot clean dataset

out_qc_loess22DG %>% 
mutate(code = as.factor(code)) %>% 
ggplot(aes(x = chl_rfu_DG)) +
  geom_histogram(bins=5,binwidth = 1, boundary = 0, fill = "steelblue", color = "black") +
#facet_wrap(~ code, scales = "free_y")+
  labs(x = "Chlorophyll (RFU)", y = "Frequency") +
  theme_bw() +
  theme(text = element_text(size = 20, colour = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))+
 scale_x_continuous(
  breaks = scales::pretty_breaks(n = 6)  #ggplot makes breaks based on data
)

ggsave(file="plots/2022_plots/expts/DG_expt_loessfilter_general_freq_distribution.png", width = 10, height = 14, dpi = 300)


##################################################333
#Final dataset

DU_EXPER_c <- out_qc_loess22DG %>% 
  group_by(code) %>%
  slice_max(order_by = chl_rfu_DG, n = 3) %>%
  summarise(
    chl_ysi = mean(chl_rfu_DG, na.rm = TRUE),
    bga_ysi = mean(bga_rfu_DG, na.rm= TRUE),
    temp_ysi = mean(temp, na.rm=TRUE),
    sp_cond_ysi = mean(sp_cond, na.rm = TRUE)
    ) %>% 
  ungroup() %>% 
  mutate(file_name = "DGysi-2022_expt")


###joining the expt readings for ysi_DG and ysi_MF
expt22_ysi <- full_join(DU_EXPER_c, MF_EXPER_c)




```




+   extracted chl for the dilution expt

```{r}

ext_chl_expt22 <- read_csv("sonde_expt_22.csv") 



# Apply the function to the defined columns to get DF

ext_chl_expt22$DF <- df_func(ext_chl_expt22$total_vol, ext_chl_expt22$sample_vol)


# Apply the function to the defined columns to get the FU conversion to chl ug/l

dilution_factor  <- ext_chl_expt22$DF

extract_volume <- ext_chl_expt22$vol_ext_ml

volume_filtered <- ext_chl_expt22$vol_fil_ml/1000
  
fluoresence_units <- ext_chl_expt22$FU

ext_chl_expt22$ext_chl <- convert_fluorescence_to_chlorophyll(fluoresence_units, dilution_factor, extract_volume, volume_filtered)




ext_chl_expt22_b <- ext_chl_expt22 %>% 
      dplyr::select(code, ext_chl)


#join the ext_chl with the ysi readings

ysi_dilution_expt <- full_join(expt22_ysi, ext_chl_expt22_b, by = "code") %>% 
  mutate(data_type = "experiment") 








```

+   Are the 2 ysi sondes seeing the same thing?


```{r}

dat1 <- ysi_field_data
dat2 <- ysi_dilution_expt

compar_data <- bind_rows(dat1, dat2) %>% 
  filter(!file_name %in% c("CR24JUL_ysi","cruise-ysidata-2023_naphtali"),
         !is.na(ext_chl) , !is.na(chl_ysi)) 
  
  

#reshape ysi data

wide_data <- compar_data %>%
  mutate(
    sonde = case_when(
      grepl("DG", file_name) ~ "DG",     #does file_name contain DG or MF?
      grepl("MF", file_name) ~ "MF"
    )
  ) %>%
  dplyr::select(code, sonde, chl_ysi) %>%
  pivot_wider(
    names_from = sonde,
    values_from = chl_ysi
  )


#plot

ggplot(wide_data, aes(x = DG, y = MF)) +
  geom_point(size = 3, alpha = 0.7) +
  
  # 1:1 line (perfect agreement)
  geom_abline(slope = 1, intercept = 0,
              color = "red", linetype = "dashed", linewidth = 1.2) +
  
  # regression line (actual relationship)
  geom_smooth(method = "lm", se = FALSE,
              color = "blue", linewidth = 1.5) +
  
  labs(
    x = "DG Sonde Chlorophyll",
    y = "MF Sonde Chlorophyll",
    title = "DG vs MF Sonde Comparison"
  ) +
  
  coord_equal() +
  theme_bw(base_size = 16) +
  stat_regline_equation(
    aes(label =  paste(after_stat(eq.label), ..rr.label.., 
                       sep = "~~~~")), show.legend = FALSE)

  

ggsave(filename = "plots/DGvsMFsensor.png", height = 10, width = 14, dpi = 300)

```

The DG and MF chl sensors were 99% in agreement.


+ Check the extracted and ysi data (lake and experiment) ??????????


```{r}



#plot
compar_data %>% 
  filter(file_name %in% c("cruise-2022_DGysi", "DGysi-2022_expt")) %>% 
ggplot(aes(x = log10(chl_ysi), y = log10(ext_chl))) +
  geom_point(size = 3, alpha = 0.7) +
  
  
  # regression line (actual relationship)
  geom_smooth(method = "lm", se = FALSE,
              color = "blue", linewidth = 1.5) +
  
  #facet_wrap(~ file_name, scales = "free_y") +
  
  labs(
    x = "Log_DG Sensor Chlorophyll",
    y = "Log_Extracted Chlorophyll",
    #title = "DG vs MF Sonde Comparison"
  ) +
  
  theme_bw(base_size = 14) +
  stat_regline_equation(
    aes(label =  paste(after_stat(eq.label), ..rr.label.., 
                       sep = "~~~~")), show.legend = FALSE)


compar_data %>% 
  filter(file_name %in% c("cruise-2022_DGysi", "DGysi-2022_expt")) %>% 
ggplot(aes(ext_chl, chl_ysi, color = file_name)) +
  geom_point(size = 3) +
  scale_x_continuous(trans = "log10") +
  geom_smooth(method = "lm", se = FALSE)+
  facet_wrap(~ file_name, scales = "free_x")+
  theme_bw(base_size = 10) +
  stat_regline_equation(
    aes(label =  paste(after_stat(eq.label), ..rr.label.., 
                       sep = "~~~~")), show.legend = FALSE)


```



+ 2024 spiking MF sonde sensor and c3 experiments with ext_chl

This file was generated from another RMD file.

```{r}


exptsjul_24 <- read_csv("2024_spiking_expts_combined_data.csv") 


MF_EXPERjul_24 <- exptsjul_24 %>% 
  filter(sos == "ysi") %>% 
  mutate(code = site,
         chl_ysi = top_chl,
         bga_ysi = top_bga,
         file_name = "2024_MFsonde_experiment",
         data_type = "experiment") %>% 
  dplyr::select(code, chl_ysi, bga_ysi, ext_chl, file_name, data_type)

MF_spiking_julexpt_24 <- MF_EXPERjul_24





```




+ join ysi dilution expt and spiking 2024 expt data with field data

```{r}


ysi_calib_data <- bind_rows(ysi_field_data, ysi_dilution_expt) %>% 
  bind_rows(MF_spiking_julexpt_24)
  
  


```





+ Linear regression for chl_ysi ~ ext_chl

+   check slopes across years

+ MF sensor

```{r}

#Field only
MF_calib <- ysi_calib_data  %>% 
  filter(data_type %in% "field") %>% 
  filter(file_name %in% c("cruise-2022_MFysi", "cruise-ysidata-2023_naphtali", "CR24JUL_ysi")) %>%
  filter(!chl_ysi > 100, !is.na(ext_chl))



MF_calib %>% 
  arrange(file_name, chl_ysi) %>% 
  ggplot(aes(x = chl_ysi, y = ext_chl)) +
  geom_point(color = "steelblue") +
  facet_wrap(~ file_name, scales= "free_y") +
  theme_bw(base_size = 14) +
  theme(
    panel.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_line(colour = "gray90", linetype = "dotted", linewidth = 0.5),
    legend.position = "right",
      legend.text = element_text(size = 12),   # 👈 increases legend labels
  legend.title = element_text(size = 14),
  
    axis.text = element_text(size = 16, margin = margin(t = 15)),
    axis.title = element_text(size = 14),
    axis.text.x = element_text(vjust = 1, hjust = 1),
    #plot.title = element_text(hjust = 0.5, size = 30, face = "bold"),
  ) 

  
ggsave(filename = "plots/CALIB_MF_sensor/rship_ext~ysi_field.png", width = 14, height = 10, dpi = 300)

```

Run lm model for MF sensor

```{r}


#MF_calib <- ysi_calib_data %>% 
  #filter(if_all(c(ext_chl, chl_ysi), ~ !is.na(.)))

model_MF <- lm(ext_chl ~ chl_ysi, data = MF_calib)
summary(model_MF)

par(mfrow = c(2,2))
plot(model_MF)



```


Transformations

+ Log data 
```{r}



MF_model_data <- MF_calib %>%
    mutate(
    log_ext_chl = log10(ext_chl),
    log_chl = log10(chl_ysi),
    log_bga = log10(bga_ysi),
    log_spcond = log10(sp_cond_ysi)
  ) %>%
  mutate(
    log_chl_c = scale(log_chl, scale = FALSE),
    log_bga_c = scale(log_bga,  scale = FALSE),
    log_spcond_c = scale(log_spcond, scale = FALSE),
    temp_c    = scale(temp_ysi,    scale = FALSE)
  )


```



log transformed
```{r}

model_MF_log <- lm(log_ext_chl ~ log_chl, data = MF_model_data)

summary(model_MF_log)

par(mfrow = c(2,2))
plot(model_MF_log)




MF_model_data %>% 
  ggplot(aes(log_chl, log_ext_chl, color = file_name)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  facet_wrap(~ file_name, scales = "free_y") +
  theme_bw(base_size = 14) +
  theme(
    panel.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_line(colour = "gray90", linetype = "dotted", linewidth = 0.5),
    legend.position = "right",
      legend.text = element_text(size = 12),   # 👈 increases legend labels
  legend.title = element_text(size = 14),
  
    axis.text = element_text(size = 16, margin = margin(t = 15)),
    axis.title = element_text(size = 14),
    axis.text.x = element_text(vjust = 1, hjust = 1),
    #plot.title = element_text(hjust = 0.5, size = 30, face = "bold"),
  ) +
  stat_regline_equation(
    aes(label =  paste(after_stat(eq.label), ..rr.label.., 
                       sep = "~~~~")), show.legend = FALSE)


ggsave(filename = "plots/CALIB_MF_sensor/log_ext~ysi.png", width = 14, height = 10, dpi = 300)




```




+   square root square root

```{r}


model_mf_sq <- lm(sqrt(ext_chl) ~ sqrt(chl_ysi), data = MF_calib)

summary(model_mf_sq)

par(mfrow = c(2,2))
plot(model_mf_sq)





```

The sqrt model does not improve the model and points still show heteroscedasticity.

+ lets try models with interaction

```{r}


model_1 <- lm(log_ext_chl ~ log_chl  * file_name,
  data = MF_model_data)


summary(model_1)


```

+   let's add some environmental variables
```{r}

#with DG

model_1 <- lm(
  log10(ext_chl + 1) ~ 
    log10(chl_ysi + 1) * file_name +
    temp_ysi + sp_cond_ysi,
  data = calib_data_field_withdg
)

summary(model_1)

#no DG

model_1 <- lm(
  log10(ext_chl + 1) ~ 
    log10(chl_ysi + 1) * file_name +
    temp_ysi + sp_cond_ysi,
  data = calib_data_field_nodg
)

summary(model_1)



```




+ Mixed effect model

```{r}

MF_model_data$file_name <- relevel(
  factor(MF_model_data$file_name), 
  ref = "cruise-ysidata-2023_naphtali" 
)


#using uncentered log data


# 1. Run the Fixed Effects Model (lm)
# This treats the 'Standard' vs 'Outlier' as hard categories
model_fixed_MF <- lm(log_ext_chl ~ log_chl + file_name, data = MF_model_data)
summary(model_fixed_MF)

# 2. Run the Mixed Effects Model (lmer)
# This treats each cruise as a random sample from a larger population
library(lme4)
model_mixed_MF <- lmer(log_ext_chl ~ log_chl + (1 | file_name), data = MF_model_data)
summary(model_mixed_MF)

par(mfrow = c(2,2))
plot(model_mixed_MF)



```

+ Predict values for global (flat average) and lmer model (accounts for cruise as random effect)

```{r}


#Global model predicted values

MF_model_data$pred_log_global <- predict(model_MF_log, newdata = MF_model_data)

sigma_log <- summary(model_MF_log)$sigma^2

bcf_log <- exp((sigma_log^2 / 2) * (log(10)^2)) #corrects log bias

MF_model_data$pred_ext_chl_global <- 10^(MF_model_data$pred_log_global) * bcf_log


# 1. Extract Fixed Effects and Sigma
coeffs <- fixef(model_mixed_MF)
b0 <- coeffs["(Intercept)"]
b1 <- coeffs["log_chl"]
sig <- sigma(model_mixed_MF)



# 2. Calculate Bias Correction Factor (BCF)
bcf_mm <- exp((sig^2 / 2) * (log(10)^2))

# 3. Predict for your dataset
MF_model_data <- MF_model_data %>%
  mutate(
    log_pred_mm   = b0 + (b1 * log_chl),  #predict data using the model
    pred_ext_mm = (10^log_pred_mm) * bcf_mm   #back transform
  )


```

+ Evaluate both global and mixed effects model _ plot both models and compare with observed

```{r}

MF_model_data_long <- MF_model_data %>% 
  pivot_longer(
    cols = c(ext_chl, pred_ext_chl_global, pred_ext_mm),
    names_to = "model",
               values_to = "chlorophyll") %>%
mutate(type = ifelse(model == "ext_chl", "Observed", "Predicted"))



ggplot(
  MF_model_data,
  aes(x = ext_chl)
) +
  geom_point(aes(y = pred_ext_chl_global, color = "Global model"), size = 2) +
   geom_point(aes(y = pred_ext_mm, color = "Mixed model"), size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  

  labs(
    x = "Observed Extracted Chlorophyll",
    y = "Predicted Extracted Chlorophyll",
    color = "Model"
  ) +
  theme_bw(base_size = 10) +
  theme(
    panel.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_line(colour = "gray90", linetype = "dotted", linewidth = 0.5),
    legend.position = "right",
      legend.text = element_text(size = 12),   # 👈 increases legend labels
  legend.title = element_text(size = 14),
  
    axis.text = element_text(size = 16, margin = margin(t = 15)),
    axis.title = element_text(size = 14),
    axis.text.x = element_text(vjust = 1, hjust = 1),
    #plot.title = element_text(hjust = 0.5, size = 30, face = "bold"),
  ) 

#+  guides(color = guide_legend(override.aes = list(size = 8))) + 
  


ggsave(file="plots/CALIB_MF_sensor/plot_models_predictedvsobs_extchl.png", width = 16, height = 10, dpi = 300)




# Calculate metrics

library(Metrics)

rmse_global <- rmse(MF_model_data$ext_chl, MF_model_data$pred_ext_chl_global) #residual mean of squared erros
mae_global <- mae(MF_model_data$ext_chl, MF_model_data$pred_ext_chl_global) #mean abs error

rmse_mm <- rmse(MF_model_data$ext_chl, MF_model_data$pred_ext_mm) #residual mean of 
mae_mm <- mae(MF_model_data$ext_chl, MF_model_data$pred_ext_mm) #mean abs error


MF_rmse_table <- tibble(rmse_global, mae_global, rmse_mm, mae_mm)

print(MF_rmse_table)



```

# MF sensor model validation (Mixed effects model)

+   Test/Train 

```{r}

#library(boot)

#cv.glm(c3_calib_data_clean, model_log, K = 10)$delta


library(dplyr)

set.seed(123)

K <- 10
folds <- sample(rep(1:K, length.out = nrow(MF_model_data)))

cv_results <- data.frame()

for (k in 1:K) {
  
  train <- MF_model_data[folds != k, ]
  test  <- MF_model_data[folds == k, ]
  
# 1. Extract Fixed Effects and Sigma

model <- model_mixed_MF
coeffs <- fixef(model_mixed_MF)
b0 <- coeffs["(Intercept)"]
b1 <- coeffs["log_chl"]
sig <- sigma(model_mixed_MF)


# 2. Calculate Bias Correction Factor (BCF)
bcf_mm <- exp((sig^2 / 2) * (log(10)^2))

# 3. Predict for your dataset
test <- test %>%
  mutate(
    log_pred_mm   = b0 + (b1 * log_chl),  #predict data using the model
    pred_ext_mm = (10^log_pred_mm) * bcf_mm   #back transform
  )

  
  # Store results
  cv_results <- rbind(cv_results, test)
}

# Final RMSE: the square root of the average of squared differences between observed and predicted values
rmse_mm_MF <- rmse(cv_results$ext_chl, cv_results$pred_ext_mm)



```

The difference between the observed ext_chl and the predicted using the mixed effects model is `r rmse_mm_MF`.


# Loading ysi monthly files

+ Apply mixed model to the monthly ysi MF sensor data


```{r}
file_paths_ysi_csv <- dir("ysi_timeseries_files", pattern = "*.csv", full.names = TRUE)


ysi_month <- purrr::map_dfr(file_paths_ysi_csv, read_and_combine_csv)  

ysi_month_b <- ysi_month %>% 
  mutate(
    # Extract "Dunga" or "Mfangano" from the file_name
    location = str_extract(file_name, "(?i)DUNGA|MFANGANO")
  ) %>% 
  
  mutate(
    date_time = as.character(date_time),
    # Remove invalid characters
    date_time = iconv(date_time, "UTF-8", "UTF-8", sub = ""), 
    date_time_parsed = parse_date_time(
      date_time,
      orders = c(
  "Ymd HMS",
  "Ymd HM",
  "dmy HMS",
  "dmy HM",
  "dmY HMS",
  "mdy HMS",
  "mdy HM"
))) %>% 
  mutate(year= year(date_time_parsed),
         month = month(date_time_parsed),
         mmyy = format(date_time_parsed, "%m-%y"))
```

+ MFANGANO DATA

```{r}

MF_month <- ysi_month_b %>% 
  filter(location %in% "MFANGANO") %>% 
  rename(chl_rfu = chl_7,
         bga_rfu = bga_pc_9)


#predict data


#transform data

MF_month$log_chl = log10(MF_month$chl_rfu)


model <- model_mixed_MF
coeffs <- fixef(model_mixed_MF)
b0 <- coeffs["(Intercept)"]
b1 <- coeffs["log_chl"]
sig <- sigma(model_mixed_MF)


# 2. Calculate Bias Correction Factor (BCF)
bcf_mm <- exp((sig^2 / 2) * (log(10)^2))

# 3. Predict for your dataset
MF_month <- MF_month %>%
  mutate(
    log_pred_mm   = b0 + (b1 * log_chl),  #predict data using the model
    pred_ext_mm = (10^log_pred_mm) * bcf_mm   #back transform and add correction term
  )






```








    + DG sonde sensor

```{r}

cruise_sondes_2022

ysi_calib_data_field <- ysi_calib_data  %>% 
  filter(data_type %in% "field") 


DG_sensor <- ysi_calib_data_field %>% 
  filter(file_name == "cruise-2022_DGysi")


model_dg <- lm(ext_chl ~ chl_ysi, data = DG_sensor)

summary(model_dg)

par(mfrow = c(2,2))
plot(model_dg)

plot(model_dg, which = 4)  #Cook's distance


ggplot(DG_sensor, aes(chl_ysi, ext_chl)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)+
  stat_regline_equation(
    aes(label =  paste(after_stat(eq.label), ..rr.label.., 
                       sep = "~~~~")), show.legend = FALSE)



# Cook’s distance

plot(model_dg, which = 4)  

cooks_d <- cooks.distance(model_dg)


DG_sensor_sub <- DG_sensor[c(1,3,4,5), ]

ggplot(DG_sensor_sub, aes(chl_ysi, ext_chl)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)+
  stat_regline_equation(
    aes(label =  paste(after_stat(eq.label), ..rr.label.., 
                       sep = "~~~~")), show.legend = FALSE)




```

 



#   **c3 files**

+   2023 cruise
```{r}

file_paths_csv <- dir("c3_cruise_files", pattern = "*.csv", full.names = TRUE)


#function to read files

read_c3_file <- function(file) {
  
  readr::read_csv(
    file,
    skip = 5,
    col_names = c("date_time", "PC", "chl", "depth", "temp")
  ) %>%
    dplyr::slice(4:n()) %>%
    dplyr::select(date_time, PC, chl, depth, temp) %>%
    
    # Force numeric conversion 
    dplyr::mutate(
      dplyr::across(
        c(PC, chl, depth, temp),
        ~ as.numeric(.)    #readr::parse_number alternaitve code
      )
    ) %>%
    
    dplyr::mutate(source_file = basename(file))
}


#read files

c3_cruise_files <- purrr::map_dfr(file_paths_csv, read_c3_file) 

c3_cruise_files_b <- c3_cruise_files  %>% 
  mutate(
    date_time = as.character(date_time),
    date_time_parsed = parse_date_time(
      date_time,
      orders = c(
  "Ymd HMS",
  "Ymd HM",
  "dmy HMS",
  "dmy HM",
  "dmY HMS",
  "mdy HMS",
  "mdY HMS",
  "mdy HM"
))) %>%
  mutate(year= year(date_time_parsed),
         month = month(date_time_parsed),
         mmyy = format(date_time_parsed, "%m-%y")) %>% 
  arrange(date_time) %>% 
  clean_names()


#check files
table(c3_cruise_files_b$source_file)

```


+   attach timestamps to different files

+   27 June 2023 cruz
```{r}

c327jun_2023 <- c3_cruise_files_b %>%
  filter(source_file %in% "c3_CR_27_June_2023.csv") %>% 
  mutate(date_time = as.POSIXct(date_time, format = "%m/%d/%Y %H:%M"),
         code = case_when(
           (date_time >= as.POSIXct("2023-06-27 11:26:00") & date_time <= as.POSIXct("2023-06-27 11:44:00")) ~ "CHL1",
          
           (date_time >= as.POSIXct("2023-06-27 11:53:00") & date_time <= as.POSIXct("2023-06-27 12:49:00")) ~ "CHL2",
           (date_time >= as.POSIXct("2023-06-27 12:51:00") & date_time <= as.POSIXct("2023-06-27 14:08:00")) ~ "BG7",
           (date_time >= as.POSIXct("2023-06-27 14:26:00") & date_time <= as.POSIXct("2023-06-27 14:40:00")) ~ "CHL3",
           (date_time >= as.POSIXct("2023-06-27 14:42:00") & date_time <= as.POSIXct("2023-06-27 14:59:00")) ~ "CHL4",
           (date_time >= as.POSIXct("2023-06-27 15:02:00") & date_time <= as.POSIXct("2023-06-27 15:18:00")) ~ "CHL5",
           (date_time >= as.POSIXct("2023-06-27 15:22:00") & date_time <= as.POSIXct("2023-06-27 15:58:00")) ~ "CHL6",
           (date_time >= as.POSIXct("2023-06-27 15:59:00") & date_time <= as.POSIXct("2023-06-27 16:28:00")) ~ "CHL7",
           (date_time >= as.POSIXct("2023-06-27 16:29:00") & date_time <= as.POSIXct("2023-06-27 18:08:00")) ~ "CHL8",
           (date_time >= as.POSIXct("2023-06-27 18:10:00") & date_time <= as.POSIXct("2023-06-27 19:14:00")) ~ "CHL9",
           (date_time >= as.POSIXct("2023-06-27 19:17:00") & date_time <= as.POSIXct("2023-06-27 20:08:00")) ~ "CHL10",
  
           TRUE ~ NA_character_
         ))





c327jun_2023_clean <- c327jun_2023 %>% 
  filter(!is.na(code)) %>% 
  arrange(date_time)


```

+ 28 June 2023 data
```{r}



c328jun_2023 <- c3_cruise_files_b %>%
  filter(source_file %in% "Turner_C3_28_June2023_cruise.csv") %>% 
  arrange(date_time) %>% 
  mutate(date_time = as.POSIXct(date_time, format = "%m/%d/%y %H:%M:%S"),
         code = case_when(
          (date_time >= as.POSIXct("2023-06-28 7:52:00")  & date_time <= as.POSIXct("2023-06-28 8:35:00"))  ~ "CHL11",
          (date_time >= as.POSIXct("2023-06-28 8:37:00")  & date_time <= as.POSIXct("2023-06-28 9:36:00")) ~ "CHL12",
          (date_time >= as.POSIXct("2023-06-28 9:40:00")  & date_time <= as.POSIXct("2023-06-28 10:05:00")) ~ "CHL13",
         (date_time >= as.POSIXct("2023-06-28 10:09:00") & date_time <= as.POSIXct("2023-06-28 11:12:00")) ~ "CHL14",
           (date_time >= as.POSIXct("2023-06-28 11:29:00") & date_time <= as.POSIXct("2023-06-28 12:24:00")) ~ "CHL15",
        (date_time >= as.POSIXct("2023-06-28 12:28:00") & date_time <= as.POSIXct("2023-06-28 13:46:00")) ~ "CHL16",
           (date_time >= as.POSIXct("2023-06-28 13:47:00") & date_time <= as.POSIXct("2023-06-28 14:58:00")) ~ "CHL17",
           (date_time >= as.POSIXct("2023-06-28 15:02:00") & date_time <= as.POSIXct("2023-06-28 15:34:00")) ~ "CHL18",
           (date_time >= as.POSIXct("2023-06-28 15:36:00") & date_time <= as.POSIXct("2023-06-28 16:16:00")) ~ "CHL19",
           (date_time >= as.POSIXct("2023-06-28 16:19:00") & date_time <= as.POSIXct("2023-06-28 17:04:00")) ~ "CHL20",
           (date_time >= as.POSIXct("2023-06-28 17:05:00") & date_time <= as.POSIXct("2023-06-28 17:19:00")) ~ "CHL21",
           (date_time >= as.POSIXct("2023-06-28 17:22:00") & date_time <= as.POSIXct("2023-06-28 18:26:00")) ~ "CHL22",   
           (date_time >= as.POSIXct("2023-06-28 18:28:00") & date_time <= as.POSIXct("2023-06-28 20:05:00")) ~ "CHL23",
            (date_time >= as.POSIXct("2023-06-28 20:10:00") & date_time <= as.POSIXct("2023-06-28 20:30:00")) ~ "CHL24",
           (date_time >= as.POSIXct("2023-06-28 20:52:00") & date_time <= as.POSIXct("2023-06-28 21:15:00")) ~ "CHL25",
           (date_time >= as.POSIXct("2023-06-28 21:23:00") & date_time <= as.POSIXct("2023-06-28 21:55:00")) ~ "CHL26",
           TRUE ~ NA_character_
         ))


 

```



+   Join 27 and 28 June cruise data
```{r}


 c3_2023_clean <- bind_rows(c327jun_2023,c328jun_2023)  %>% 
    filter(!is.na(code)) %>% 
  mutate(code_clean = str_remove_all(code, " ")) 

#write.csv(c3_2023_clean, file="c3_2023_clean.csv") 



```

+ Frequency distribution for c3 cuise dataset


```{r}


c3_2023_clean %>% 
  filter(chl > 0) %>% 
  mutate(code = as.factor(code_clean)) %>% 
ggplot(aes(x = chl)) +
  geom_histogram(bins = 30, binwidth = 2, boundary = 0, fill = "steelblue", color = "black") +
  #facet_wrap(~ code) +
  labs(x = "Chlorophyll (RFU)", y = "Frequency") +
  theme_bw() +
  theme(text = element_text(size = 15, colour = "black"),
        axis.text.x = element_text(size=20, angle = 45, hjust = 1, vjust = 1))+
 scale_x_continuous(
  breaks = scales::pretty_breaks(n = 6)  #ggplot makes breaks based on data
)


ggsave(file="plots/c3/2023_plots/c3_2023cruz_gen_freq_distribution.png", width = 10, height = 12, dpi = 300)







```




+   Remove outliers from 2023 cruise c3 data

```{r}

c3_23_outlier_detect <- multidetect(data = c3_2023_clean, 
                                var = "chl", 
                                var_col = "code_clean",
                                multiple = TRUE,
                                methods = c("adjbox", "iqr", "hampel","jknife",
                                            "seqfences", "mixediqr",
                                            "distboxplot", "semiqr",
                                            "zscore", "logboxplot", "medianrule"),
                       silence_true_errors = TRUE, 
                       missingness = 0.1,
                       sdm = FALSE,
                       na.inform = TRUE)
  
  

#Visualize the number of outliers detected by each method

ggoutliers(c3_23_outlier_detect,
                        select = unique(c3_2023_clean$code_clean),
                        color='purple',
                        desc = TRUE,
                        ncol = 10,
                        nrow = 10)

ggsave(file = "plots/c3/2023_plots/outliers_per_site.png", width = 10, height = 10, dpi = 300)


#Obtaining quality controlled dataset using loess method or data labeling


out_qc_loess23 <- extract_clean_data(refdata = c3_2023_clean, 
            outliers = c3_23_outlier_detect, loess = TRUE,
            var_col = "code_clean"
            )  #set threshold to 0.8 to remove the very strong outliers only

#clean dataset
nrow(out_qc_loess23)

#reference data
nrow(c3_2023_clean)

out_qc_labeled23 <- classify_data(refdata = c3_2023_clean, 
                                  var_col = "code_clean",
                                outliers = c3_23_outlier_detect)


#Visualize labelled quality controlled dataset

ggenvironmentalspace(out_qc_labeled23, 
                     type = '1D',
                     ggxangle = 45, 
                     ncol = 10,
                     nrow = 15,
                     scalecolor = 'viridis',
                     xhjust = 1,
                     legend_position = 'blank',
                     ylab = "Number of records",
                     xlab = "Outlier labels")


ggsave(filename = "plots/c3/2023_plots/filtered_site_data.png", width = 14, height = 10, dpi = 300)



#labelled outlier or not 
labeled23 <- classify_data(refdata = c3_2023_clean, 
                         var_col = "code_clean",
                         outliers = c3_23_outlier_detect)



#plot clean dataset

out_qc_loess23 %>% 
mutate(code = as.factor(code_clean)) %>% 
ggplot(aes(x = chl)) +
  geom_histogram(bins = 30, binwidth = 2, boundary = 0, fill = "steelblue", color = "black") +
facet_wrap(~ code, scales = "free_y")+
  labs(x = "Chlorophyll (ug/l)", y = "Frequency") +
  theme_bw() +
  theme(text = element_text(size = 10, colour = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))+
 scale_x_continuous(
  breaks = scales::pretty_breaks(n = 6)  #ggplot makes breaks based on data
)

ggsave(file="plots/2023_plots/JUN2023_loessfilter_Sites_freq_distribution.png", width = 12, height = 10, dpi = 300)








```

The low value of chl at CHL22 is 97ug/l. It is not removed but still is high enough compared to many other sites... The threshold for this cruise is flexible

Chl at chl4 is also somewhat variable but the values are over all high 

+ Final filtering of points for Jun 2023 cruise

```{r}

final_crjun_2023_df <- out_qc_loess23 %>%
  group_by(code_clean, year) %>%
  slice_max(order_by = chl, n = 3) %>%
  summarise(
    chl = mean(chl, na.rm = TRUE),
    pc = mean(pc, na.rm = TRUE),
    temp = mean(temp, na.rm = TRUE),
    depth = mean(depth, na.rm = TRUE),
    source_file  = unique(source_file),
  ) 




```







+     July 2024

```{r}

c3_jul2024 <- c3_cruise_files_b %>%
  filter(source_file %in% "c3_CR_24JUL2024.csv") %>% 
  mutate(date_time = as.POSIXct(date_time, format = "%m/%d/%y %H:%M:%S"),
         code = case_when(
        (date_time >= as.POSIXct("2024-07-18 14:29:00") & date_time <= as.POSIXct("2024-07-18 15:00:00")) ~ "CHL26",
        (date_time >= as.POSIXct("2024-07-18 15:37:00") & date_time <= as.POSIXct("2024-07-18 15:46:00")) ~ "CHL25",
        (date_time >= as.POSIXct("2024-07-18 17:12:00") & date_time <= as.POSIXct("2024-07-18 17:31:00")) ~ "CHL24",
        (date_time >= as.POSIXct("2024-07-18 18:57:00") & date_time <= as.POSIXct("2024-07-18 19:15:00")) ~ "CHL3",
           (date_time >= as.POSIXct("2024-07-18 19:37:00") & date_time <= as.POSIXct("2024-07-18 19:50:00")) ~ "CHL4",
           (date_time >= as.POSIXct("2024-07-18 20:15:00") & date_time <= as.POSIXct("2024-07-18 20:30:00")) ~ "CHL5",
           (date_time >= as.POSIXct("2024-07-18 20:48:00") & date_time <= as.POSIXct("2024-07-18 21:05:00")) ~ "CHL6",
           (date_time >= as.POSIXct("2024-07-18 21:51:00") & date_time <= as.POSIXct("2024-07-18 22:15:00")) ~ "CHL7",
        
           (date_time >= as.POSIXct("2024-07-19 07:48:00") & date_time <= as.POSIXct("2024-07-19 08:09:00")) ~ "BG10",
           (date_time >= as.POSIXct("2024-07-19 09:27:00") & date_time <= as.POSIXct("2024-07-19 09:50:00")) ~ "CHL9",
          (date_time >= as.POSIXct("2024-07-19 10:45:00") & date_time <= as.POSIXct("2024-07-19 11:02:00")) ~ "CHL10",
        (date_time >= as.POSIXct("2024-07-19 11:45:00")  & date_time <= as.POSIXct("2024-07-19 11:57:00"))  ~ "BG13",
        (date_time >= as.POSIXct("2024-07-19 14:20:00")  & date_time <= as.POSIXct("2024-07-19 14:40:00")) ~ "CHL19",
        (date_time >= as.POSIXct("2024-07-19 16:02:00")  & date_time <= as.POSIXct("2024-07-19 16:13:00")) ~ "CHL18",
        (date_time >= as.POSIXct("2024-07-19 16:52:00") & date_time <= as.POSIXct("2024-07-19 17:13:00")) ~ "CHL15",
        (date_time >= as.POSIXct("2024-07-19 17:50:00") & date_time <= as.POSIXct("2024-07-19 18:10:00")) ~ "CNL1",
        
        (date_time >= as.POSIXct("2024-07-20 08:31:00") & date_time <= as.POSIXct("2024-07-20 08:51:00")) ~ "CNL2",
        (date_time >= as.POSIXct("2024-07-20 09:52:00") & date_time <= as.POSIXct("2024-07-20 10:06:00")) ~ "CHL12",
        (date_time >= as.POSIXct("2024-07-20 10:46:00") & date_time <= as.POSIXct("2024-07-20 11:00:00")) ~ "CHL13",
        (date_time >= as.POSIXct("2024-07-20 12:01:00") & date_time <= as.POSIXct("2024-07-20 12:15:00")) ~ "BG16",
        (date_time >= as.POSIXct("2024-07-20 14:29:00") & date_time <= as.POSIXct("2024-07-20 14:39:00")) ~ "CHL20",
        (date_time >= as.POSIXct("2024-07-20 15:35:00") & date_time <= as.POSIXct("2024-07-20 15:41:00")) ~ "CHL21",
        (date_time >= as.POSIXct("2024-07-20 18:16:00") & date_time <= as.POSIXct("2024-07-20 18:25:00")) ~ "CHL27",
        (date_time >= as.POSIXct("2024-07-20 19:00:00") & date_time <= as.POSIXct("2024-07-20 19:15:00")) ~ "BG5",
           
           
           TRUE ~ NA_character_
         ))

c3_jul2024_clean <- c3_jul2024 %>% 
  filter(!is.na(code))





```





+   Frequency distribution for c3 2024 July cruise dataset



```{r}

#Keeping chl values > 0

c3_jul2024_clean %>% 
  filter(chl > 0) %>% 
  mutate(code = as.factor(code)) %>% 
ggplot(aes(x = chl)) +
  geom_histogram(bins = 30, binwidth = 2, boundary = 0, fill = "steelblue", color = "black") +
  #facet_wrap(~ code) +
  labs(x = "Chlorophyll (RFU)", y = "Frequency") +
  theme_bw() +
  theme(text = element_text(size = 15, colour = "black"),
        axis.text.x = element_text(size=14, angle = 45, hjust = 1, vjust = 1))+
 scale_x_continuous(
  breaks = scales::pretty_breaks(n = 6)  #ggplot makes breaks based on data
)


ggsave(file="plots/c3/2024_plots/c3_2024JULcruz_site_freq_distribution.png", width = 10, height = 12, dpi = 300)








```

+   Remove outliers from 2024 july cruise c3 data

```{r}


c3_j24_outlier_detect <- multidetect(data = c3_jul2024_clean, 
                                var = "chl", 
                                var_col = "code",
                                multiple = TRUE,
                                methods = c("adjbox", "iqr", "hampel","jknife",
                                            "seqfences", "mixediqr",
                                            "distboxplot", "semiqr",
                                            "zscore", "logboxplot", "medianrule"),
                       silence_true_errors = TRUE, 
                       missingness = 0.1,
                       sdm = FALSE,
                       na.inform = TRUE)
  
  

#Visualize the number of outliers detected by each method

ggoutliers(c3_j24_outlier_detect,
                        select = unique(c3_jul2024_clean$code),
                        color='purple',
                        desc = TRUE,
                        ncol = 5,
                        nrow = 10)

ggsave(file = "plots/c3/2024_plots/outliers_per_site.png", width = 10, height = 20, dpi = 300)


#Obtaining quality controlled dataset using loess method or data labeling


out_qc_loessj24 <- extract_clean_data(refdata = c3_jul2024_clean, 
            outliers = c3_j24_outlier_detect, loess = TRUE,
            var_col = "code"
            )  #set threshold to 0.8 to remove the very strong outliers only

#clean dataset
nrow(out_qc_loessj24)

#reference data
nrow(c3_jul2024_clean)

out_qc_labeled24 <- classify_data(refdata = c3_jul2024_clean, 
                                  var_col = "code",
                                outliers = c3_j24_outlier_detect)


#Visualize labelled quality controlled dataset

ggenvironmentalspace(out_qc_labeled24, 
                     type = '1D',
                     ggxangle = 45, 
                     ncol = 5,
                     nrow = 15,
                     scalecolor = 'viridis',
                     xhjust = 1,
                     legend_position = 'blank',
                     ylab = "Number of records",
                     xlab = "Outlier labels")


ggsave(filename = "plots/c3/2024_plots/filtered_site_data.png", width = 14, height = 14, dpi = 300)



#labelled outlier or not 
labeled24 <- classify_data(refdata = c3_jul2024_clean, 
                         var_col = "code",
                         outliers = c3_j24_outlier_detect)



#plot clean dataset

out_qc_loessj24 %>% 
mutate(code = as.factor(code)) %>% 
ggplot(aes(x = chl)) +
  geom_histogram(bins = 30, binwidth = 2, boundary = 0, fill = "steelblue", color = "black") +
facet_wrap(~ code)+
  labs(x = "Chlorophyll (ug/l)", y = "Frequency") +
  theme_bw() +
  theme(text = element_text(size = 10, colour = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))+
 scale_x_continuous(
  breaks = scales::pretty_breaks(n = 6)  #ggplot makes breaks based on data
)

ggsave(file="plots/2024_plots/c3_2024_loessfilter_Sites_freq_distribution.png", width = 10, height = 12, dpi = 300)






```

Notes: CHL21 is problematic... high variation



+ Final July cruz 2024 dataset

```{r}


final_crjul_2024_df <- out_qc_loessj24 %>%
  rename(code_clean = code) %>% 
  group_by(code_clean, year) %>%
  summarise(
    chl = mean(chl, na.rm = TRUE),
    pc = mean(pc, na.rm = TRUE),
    temp = mean(temp, na.rm = TRUE),
    depth = mean(depth, na.rm = TRUE),
    source_file  = unique(source_file),
  ) 



final_crjul_2024_df2 <- out_qc_loessj24 %>%
  group_by(code, year) %>%
  slice_max(order_by = chl, n = 3) %>%
  summarise(
    chl = mean(chl, na.rm = TRUE),
    pc = mean(pc, na.rm = TRUE),
    temp = mean(temp, na.rm = TRUE),
    depth = mean(depth, na.rm = TRUE),
    source_file  = unique(source_file),
  ) 





```

For final dataset, I tried the top average vs averaging all cleaned values. The former over estimates chl in gulf areas that are more turbid while the latter averages better to provide a better comparison across neigbouring sites. So, I will just do the average of the cleaned values after removing outliers.




+     Nov 2024

```{r}

c3_nov2024 <- c3_cruise_files_b %>%
  filter(source_file %in% "c3_CR_NOV2024.csv") %>% 
  mutate(date_time = as.POSIXct(date_time, format = "%m/%d/%Y %H:%M"),
         code = case_when(
        (date_time >= as.POSIXct("2024-11-07 09:35:00") & date_time <= as.POSIXct("2024-11-07 09:43:00")) ~ "CHL26",
        (date_time >= as.POSIXct("2024-11-07 11:06:00") & date_time <= as.POSIXct("2024-11-07 11:35:00")) ~ "CHL25",
        (date_time >= as.POSIXct("2024-11-07 12:27:00") & date_time <= as.POSIXct("2024-11-07 12:55:00")) ~ "CHL24",
        (date_time >= as.POSIXct("2024-11-07 14:46:00") & date_time <= as.POSIXct("2024-11-07 14:56:00")) ~ "BG5",
        (date_time >= as.POSIXct("2024-11-07 16:28:00") & date_time <= as.POSIXct("2024-11-07 16:48:00")) ~ "CHL27",
        (date_time >= as.POSIXct("2024-11-07 17:40:00") & date_time <= as.POSIXct("2024-11-07 18:04:00")) ~ "CHL3",
        (date_time >= as.POSIXct("2024-11-07 18:30:00") & date_time <= as.POSIXct("2024-11-07 18:56:00")) ~ "CHL4",
        (date_time >= as.POSIXct("2024-11-07 19:24:00") & date_time <= as.POSIXct("2024-11-07 19:40:00")) ~ "CHL5",
        (date_time >= as.POSIXct("2024-11-07 20:05:00") & date_time <= as.POSIXct("2024-11-07 20:30:00")) ~ "CHL6",
        
        (date_time >= as.POSIXct("2024-11-08 08:06:00") & date_time <= as.POSIXct("2024-11-08 08:24:00")) ~ "BG10",
           
        (date_time >= as.POSIXct("2024-11-08 09:29:00")  & date_time <= as.POSIXct("2024-11-08 09:50:00")) ~ "CHL9",
        (date_time >= as.POSIXct("2024-11-08 10:07:00")  & date_time <= as.POSIXct("2024-11-08 10:26:00"))  ~ "CHL7",
        (date_time >= as.POSIXct("2024-11-08 11:22:00")  & date_time <= as.POSIXct("2024-11-08 11:42:00")) ~ "CHL10",
        (date_time >= as.POSIXct("2024-11-08 12:20:00")  & date_time <= as.POSIXct("2024-11-08 12:34:00")) ~ "BG13",
        (date_time >= as.POSIXct("2024-11-08 14:00:00") & date_time <= as.POSIXct("2024-11-08 14:23:00")) ~ "CHL19",
        (date_time >= as.POSIXct("2024-11-08 14:59:00") & date_time <= as.POSIXct("2024-11-08 15:20:00")) ~ "CHL18",
        (date_time >= as.POSIXct("2024-11-08 15:51:00") & date_time <= as.POSIXct("2024-11-08 16:05:00")) ~ "BG16",
        (date_time >= as.POSIXct("2024-11-08 17:13:00") & date_time <= as.POSIXct("2024-11-08 17:29:00")) ~ "CHL15",
        (date_time >= as.POSIXct("2024-11-08 17:55:00") & date_time <= as.POSIXct("2024-11-08 18:14:00")) ~ "CNL1",
        
        
        (date_time >= as.POSIXct("2024-11-09 07:40:00") & date_time <= as.POSIXct("2024-11-09 08:14:00")) ~ "CNL2",
        (date_time >= as.POSIXct("2024-11-09 09:20:00") & date_time <= as.POSIXct("2024-11-09 09:47:00")) ~ "CHL12",
        (date_time >= as.POSIXct("2024-11-09 10:56:00") & date_time <= as.POSIXct("2024-11-09 11:20:00")) ~ "CHL13",
        (date_time >= as.POSIXct("2024-11-09 15:52:00") & date_time <= as.POSIXct("2024-11-09 16:09:00")) ~ "CHL20",
        (date_time >= as.POSIXct("2024-11-09 17:03:00") & date_time <= as.POSIXct("2024-11-09 17:14:00")) ~ "CHL21",
        (date_time >= as.POSIXct("2024-11-09 17:35:00") & date_time <= as.POSIXct("2024-11-09 17:52:00")) ~ "BG18",
        (date_time >= as.POSIXct("2024-11-09 19:33:00") & date_time <= as.POSIXct("2024-11-09 19:50:00")) ~ "BG7",
           
           
           TRUE ~ NA_character_
         ))

c3_nov2024_clean <- c3_nov2024 %>% 
  filter(!is.na(code))




```

CHL21 data is still quite funky.... 


+   Frequency distribution for Nov 2024 cruz C3 data

```{r}

c3_nov2024_clean %>% 
  filter(chl > 0) %>% 
  mutate(code = as.factor(code)) %>% 
ggplot(aes(x = chl)) +
  geom_histogram(bins = 30, binwidth = 2, boundary = 0, fill = "steelblue", color = "black") +
  facet_wrap(~ code) +
  labs(x = "Chlorophyll (RFU)", y = "Frequency") +
  theme_bw() +
  theme(text = element_text(size = 15, colour = "black"),
        axis.text.x = element_text(size=14, angle = 45, hjust = 1, vjust = 1))+
 scale_x_continuous(
  breaks = scales::pretty_breaks(n = 6)  #ggplot makes breaks based on data
)


ggsave(file="plots/c3/2024_plots/c3_2024NOVcruz_site_freq_distribution.png", width = 10, height = 12, dpi = 300)

```


+ Removing outliers from Nov cruise data

```{r}
c3_n24_outlier_detect <- multidetect(data = c3_nov2024_clean, 
                                var = "chl", 
                                var_col = "code",
                                multiple = TRUE,
                                methods = c("adjbox", "iqr", "hampel","jknife",
                                            "seqfences", "mixediqr",
                                            "distboxplot", "semiqr",
                                            "zscore", "logboxplot", "medianrule"),
                       silence_true_errors = TRUE, 
                       missingness = 0.1,
                       sdm = FALSE,
                       na.inform = TRUE)
  
  

#Visualize the number of outliers detected by each method

ggoutliers(c3_n24_outlier_detect,
                        select = unique(c3_nov2024_clean$code),
                        color='purple',
                        desc = TRUE,
                        ncol = 10,
                        nrow = 10)

ggsave(file = "plots/c3/2024_plots/outliersNOV_per_site.png", width = 10, height = 20, dpi = 300)


#Obtaining quality controlled dataset using loess method or data labeling


out_qc_loessN24 <- extract_clean_data(refdata = c3_nov2024_clean, 
            outliers = c3_n24_outlier_detect, loess = TRUE,
            var_col = "code"
            )  #set threshold to 0.8 to remove the very strong outliers only

#clean dataset
nrow(out_qc_loessN24)

#reference data
nrow(c3_nov2024_clean)

out_qc_labeledN24 <- classify_data(refdata = c3_nov2024_clean, 
                                  var_col = "code",
                                outliers = c3_n24_outlier_detect)


#Visualize labelled quality controlled dataset

ggenvironmentalspace(out_qc_labeledN24, 
                     type = '1D',
                     ggxangle = 45, 
                     ncol = 10,
                     nrow = 15,
                     scalecolor = 'viridis',
                     xhjust = 1,
                     legend_position = 'blank',
                     ylab = "Number of records",
                     xlab = "Outlier labels")


ggsave(filename = "plots/c3/2024_plots/filtered_siteNOV_data.png", width = 14, height = 14, dpi = 300)



#labelled outlier or not 
labeledN24 <- classify_data(refdata = c3_nov2024_clean, 
                         var_col = "code",
                         outliers = c3_n24_outlier_detect)



#plot clean dataset

out_qc_loessN24 %>% 
mutate(code = as.factor(code)) %>% 
ggplot(aes(x = chl)) +
  geom_histogram(bins = 30, binwidth = 2, boundary = 0, fill = "steelblue", color = "black") +
facet_wrap(~ code, scales = "free_y")+
  labs(x = "Chlorophyll (ug/l)", y = "Frequency") +
  theme_bw() +
  theme(text = element_text(size = 10, colour = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))+
 scale_x_continuous(
  breaks = scales::pretty_breaks(n = 6)  #ggplot makes breaks based on data
)

ggsave(file="plots/2024_plots/c3_2024_loessfilter_SitesNOV_freq_distribution.png", width = 10, height = 12, dpi = 300)



```

CHL21 is also very problematic here



+ Get final dataset

```{r}


final_crnov_2024_df <- out_qc_loessN24 %>%
  rename(code_clean = code) %>% 
  group_by(code_clean, year) %>%
  summarise(
    chl = mean(chl, na.rm = TRUE),
    pc = mean(pc, na.rm = TRUE),
    temp = mean(temp, na.rm = TRUE),
    depth = mean(depth, na.rm = TRUE),
    source_file  = unique(source_file),
  ) 


```


+ April 2025 time stamps

```{r}

c3_apr2025 <- c3_cruise_files_b %>%
  filter(source_file %in% "c3_CR_APR2025.csv") %>% 
  mutate(date_time = as.POSIXct(date_time, format = "%m/%d/%y %H:%M:%S"),
         code = case_when(
        (date_time >= as.POSIXct("2025-04-05 08:37:00") & date_time <= as.POSIXct("2025-04-05 08:45:00")) ~ "BG5",
        (date_time >= as.POSIXct("2025-04-05 10:40:00") & date_time <= as.POSIXct("2025-04-05 11:00:00")) ~ "CHL27",
        
        (date_time >= as.POSIXct("2025-04-05 11:33:00") & date_time <= as.POSIXct("2025-04-05 11:36:00")) ~ "CHL1",
        (date_time >= as.POSIXct("2025-04-05 11:38:00") & date_time <= as.POSIXct("2025-04-05 11:42:00")) ~ "CHL2",
        (date_time >= as.POSIXct("2025-04-05 13:08:00") & date_time <= as.POSIXct("2025-04-05 13:40:00")) ~ "BG7",
        (date_time >= as.POSIXct("2025-04-05 14:54:00") & date_time <= as.POSIXct("2025-04-05 15:02:00")) ~ "CHL3",
        (date_time >= as.POSIXct("2025-04-05 15:52:00") & date_time <= as.POSIXct("2025-04-05 16:00:00")) ~ "CHL4",
        (date_time >= as.POSIXct("2025-04-05 16:40:00") & date_time <= as.POSIXct("2025-04-05 17:00:00")) ~ "CHL5",
        (date_time >= as.POSIXct("2025-04-05 17:24:00")  & date_time <= as.POSIXct("2025-04-05 17:44:00")) ~ "CHL6",
        (date_time >= as.POSIXct("2025-04-05 18:19:00")  & date_time <= as.POSIXct("2025-04-05 18:37:00"))  ~ "CHL7",
        
        (date_time >= as.POSIXct("2025-04-06 17:55:00")  & date_time <= as.POSIXct("2025-04-06 18:15:00")) ~ "BG10",
        (date_time >= as.POSIXct("2025-04-06 19:03:00")  & date_time <= as.POSIXct("2025-04-06 19:30:00")) ~ "CHL9",
        
        (date_time >= as.POSIXct("2025-04-07 10:12:00") & date_time <= as.POSIXct("2025-04-07 10:24:00")) ~ "CHL18",
        (date_time >= as.POSIXct("2025-04-07 11:14:00") & date_time <= as.POSIXct("2025-04-07 11:26:00")) ~ "CHL15",
        (date_time >= as.POSIXct("2025-04-07 12:06:00") & date_time <= as.POSIXct("2025-04-07 12:30:00")) ~ "CNL1",
        (date_time >= as.POSIXct("2025-04-07 15:18:00") & date_time <= as.POSIXct("2025-04-07 15:38:00")) ~ "CNL2",
        (date_time >= as.POSIXct("2025-04-07 17:00:00") & date_time <= as.POSIXct("2025-04-07 17:30:00")) ~ "CHL12",
        (date_time >= as.POSIXct("2025-04-07 07:09:00") & date_time <= as.POSIXct("2025-04-07 07:17:00")) ~ "BG13",
         (date_time >= as.POSIXct("2025-04-07 09:16:00") & date_time <= as.POSIXct("2025-04-07 09:28:00")) ~ "CHL19",
        (date_time >= as.POSIXct("2025-04-07 17:47:00") & date_time <= as.POSIXct("2025-04-07 18:05:00")) ~ "CHL13",
        
        
        (date_time >= as.POSIXct("2025-04-08 07:20:00") & date_time <= as.POSIXct("2025-04-08 07:43:00")) ~ "BG16",
        (date_time >= as.POSIXct("2025-04-08 10:37:00") & date_time <= as.POSIXct("2025-04-08 10:45:00")) ~ "CHL20",
        (date_time >= as.POSIXct("2025-04-08 11:53:00") & date_time <= as.POSIXct("2025-04-08 12:03:00")) ~ "CHL21",
        (date_time >= as.POSIXct("2025-04-08 12:58:00") & date_time <= as.POSIXct("2025-04-08 13:10:00")) ~ "CHL22",
        (date_time >= as.POSIXct("2025-04-08 13:28:00") & date_time <= as.POSIXct("2025-04-08 13:40:00")) ~ "BG18",
        (date_time >= as.POSIXct("2025-04-08 14:25:00") & date_time <= as.POSIXct("2025-04-08 14:35:00")) ~ "CHL23B",
        (date_time >= as.POSIXct("2025-04-08 16:12:00") & date_time <= as.POSIXct("2025-04-08 16:29:00")) ~ "CHL24",
           
           
           TRUE ~ NA_character_
         ))

c3_apr2025_clean <- c3_apr2025 %>% 
  filter(!is.na(code))







```



+   Frequency distribution for April 2025 cruz C3 data

```{r}

c3_apr2025_clean %>% 
  filter(chl > 0) %>% 
  mutate(code = as.factor(code)) %>% 
ggplot(aes(x = chl)) +
  geom_histogram(bins = 30, binwidth = 2, boundary = 0, fill = "steelblue", color = "black") +
  facet_wrap(~ code) +
  labs(x = "Chlorophyll (RFU)", y = "Frequency") +
  theme_bw() +
  theme(text = element_text(size = 15, colour = "black"),
        axis.text.x = element_text(size=14, angle = 45, hjust = 1, vjust = 1))+
 scale_x_continuous(
  breaks = scales::pretty_breaks(n = 6)  #ggplot makes breaks based on data
)


ggsave(file="plots/c3/2025_plots/c3_2025APRcruz_site_freq_distribution.png", width = 10, height = 10, dpi = 300)

```



+ Removing outliers from April cruise data

```{r}

c3_Apr25_outlier_detect <- multidetect(data = c3_apr2025_clean, 
                                var = "chl", 
                                var_col = "code",
                                multiple = TRUE,
                                methods = c("adjbox", "iqr", "hampel","jknife",
                                            "seqfences", "mixediqr",
                                            "distboxplot", "semiqr",
                                            "zscore", "logboxplot", "medianrule"),
                       silence_true_errors = TRUE, 
                       missingness = 0.1,
                       sdm = FALSE,
                       na.inform = TRUE)
  
  

#Visualize the number of outliers detected by each method

ggoutliers(c3_Apr25_outlier_detect,
                        select = unique(c3_apr2025_clean$code),
                        color='purple',
                        desc = TRUE,
                        ncol = 10,
                        nrow = 10)

ggsave(file = "plots/c3/2025_plots/outliersAPR25_per_site.png", width = 10, height = 10, dpi = 300)


#Obtaining quality controlled dataset using loess method or data labeling


out_qc_loessAPR25 <- extract_clean_data(refdata = c3_apr2025_clean, 
            outliers = c3_Apr25_outlier_detect, loess = TRUE,
            var_col = "code"
            )  #set threshold to 0.8 to remove the very strong outliers only

#clean dataset
nrow(out_qc_loessAPR25)

#reference data
nrow(c3_apr2025_clean)

out_qc_labeledAPR25 <- classify_data(refdata = c3_apr2025_clean, 
                                  var_col = "code",
                                outliers = c3_Apr25_outlier_detect)


#Visualize labelled quality controlled dataset

ggenvironmentalspace(out_qc_labeledAPR25, 
                     type = '1D',
                     ggxangle = 45, 
                     ncol = 10,
                     nrow = 10,
                     scalecolor = 'viridis',
                     xhjust = 1,
                     legend_position = 'blank',
                     ylab = "Number of records",
                     xlab = "Outlier labels")


ggsave(filename = "plots/c3/2025_plots/filtered_siteAPR25_data.png", width = 14, height = 10, dpi = 300)



#labelled outlier or not 
labeledAPR25 <- classify_data(refdata = c3_apr2025_clean, 
                         var_col = "code",
                         outliers = c3_Apr25_outlier_detect)



#plot clean dataset

out_qc_loessAPR25 %>% 
mutate(code = as.factor(code)) %>% 
ggplot(aes(x = chl)) +
  geom_histogram(bins = 30, binwidth = 2, boundary = 0, fill = "steelblue", color = "black") +
facet_wrap(~ code, scales = "free_y")+
  labs(x = "Chlorophyll (ug/l)", y = "Frequency") +
  theme_bw() +
  theme(text = element_text(size = 10, colour = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))+
 scale_x_continuous(
  breaks = scales::pretty_breaks(n = 6)  #ggplot makes breaks based on data
)

ggsave(file="plots/2025_plots/loessfilter_siteAPR_freq_distribution.png", width = 10, height = 10, dpi = 300)



```





+ Get final dataset

```{r}


final_crapr_2025_df <- out_qc_loessAPR25 %>% 
  rename(code_clean = code) %>% 
  group_by(code_clean, year) %>%
  summarise(
    chl = mean(chl, na.rm = TRUE),
    pc = mean(pc, na.rm = TRUE),
    temp = mean(temp, na.rm = TRUE),
    depth = mean(depth, na.rm = TRUE),
    source_file  = unique(source_file),
  ) 


```


+ Jun 2025


```{r}


c3_jun2025 <- c3_cruise_files_b %>%
  filter(source_file %in% "c3_CR_JUN2025.csv") %>% 
  mutate(date_time = as.POSIXct(date_time, format = "%m/%d/%Y %H:%M"),
         code = case_when(
       
        (date_time >= as.POSIXct("2025-06-23 07:34:00") & date_time <= as.POSIXct("2025-06-23 08:00:00")) ~ "CHL26",
        (date_time >= as.POSIXct("2025-06-23 09:06:00") & date_time <= as.POSIXct("2025-06-23 09:17:00")) ~ "CHL25",
        (date_time >= as.POSIXct("2025-06-23 10:31:00") & date_time <= as.POSIXct("2025-06-23 10:46:00")) ~ "BG5",
        (date_time >= as.POSIXct("2025-06-23 12:30:00") & date_time <= as.POSIXct("2025-06-23 12:42:00")) ~ "CHL27",
        
        (date_time >= as.POSIXct("2025-06-23 13:42:00") & date_time <= as.POSIXct("2025-06-23 13:55:00")) ~ "CHL3",
        (date_time >= as.POSIXct("2025-06-23 14:18:00") & date_time <= as.POSIXct("2025-06-23 14:23:00")) ~ "BG7",
        (date_time >= as.POSIXct("2025-06-23 15:12:00") & date_time <= as.POSIXct("2025-06-23 15:17:00")) ~ "CHL2",
        (date_time >= as.POSIXct("2025-06-23 15:40:00") & date_time <= as.POSIXct("2025-06-23 16:00:00")) ~ "CHL1",
        (date_time >= as.POSIXct("2025-06-23 17:00:00") & date_time <= as.POSIXct("2025-06-23 17:15:00")) ~ "CHL4",
        (date_time >= as.POSIXct("2025-06-23 17:29:00") & date_time <= as.POSIXct("2025-06-23 17:38:00")) ~ "CHL5",
        
        (date_time >= as.POSIXct("2025-06-23 18:02:00") & date_time <= as.POSIXct("2025-06-23 18:15:00")) ~ "CHL6",
        (date_time >= as.POSIXct("2025-06-23 18:46:00") & date_time <= as.POSIXct("2025-06-23 18:55:00")) ~ "CHL7",
        
        (date_time >= as.POSIXct("2025-06-24 07:30:00") & date_time <= as.POSIXct("2025-06-24 07:48:00")) ~ "BG10",
        
        (date_time >= as.POSIXct("2025-06-24 08:52:00")  & date_time <= as.POSIXct("2025-06-24 09:02:00")) ~ "CHL9",
        (date_time >= as.POSIXct("2025-06-24 10:07:00")  & date_time <= as.POSIXct("2025-06-24 10:20:00"))  ~ "CHL10",
        (date_time >= as.POSIXct("2025-06-24 10:50:00")  & date_time <= as.POSIXct("2025-06-24 10:58:00")) ~ "BG13",
        
        (date_time >= as.POSIXct("2025-06-24 10:50:00") & date_time <= as.POSIXct("2025-06-24 10:58:00")) ~ "BG13",
        (date_time >= as.POSIXct("2025-06-24 12:23:00") & date_time <= as.POSIXct("2025-06-24 12:33:00")) ~ "CHL19",
        (date_time >= as.POSIXct("2025-06-24 13:18:00") & date_time <= as.POSIXct("2025-06-24 13:25:00")) ~ "CHL18",
        (date_time >= as.POSIXct("2025-06-24 14:18:00") & date_time <= as.POSIXct("2025-06-24 14:29:00")) ~ "CHL15",
        
     
        (date_time >= as.POSIXct("2025-06-24 15:09:00") & date_time <= as.POSIXct("2025-06-24 15:28:00")) ~ "CNL1",
        (date_time >= as.POSIXct("2025-06-24 17:13:00") & date_time <= as.POSIXct("2025-06-24 17:28:00")) ~ "CNL2",
        (date_time >= as.POSIXct("2025-06-24 18:22:00") & date_time <= as.POSIXct("2025-06-24 18:32:00")) ~ "CHL12",
        (date_time >= as.POSIXct("2025-06-24 19:53:00") & date_time <= as.POSIXct("2025-06-24 19:59:00")) ~ "CHL13",
        
        (date_time >= as.POSIXct("2025-06-25 09:33:00") & date_time <= as.POSIXct("2025-06-25 09:53:00")) ~ "BG16",
        (date_time >= as.POSIXct("2025-06-25 15:05:00") & date_time <= as.POSIXct("2025-06-25 15:17:00")) ~ "CHL20",
        (date_time >= as.POSIXct("2025-06-25 16:02:00") & date_time <= as.POSIXct("2025-06-25 16:15:00")) ~ "CHL21",
        (date_time >= as.POSIXct("2025-06-25 16:26:00") & date_time <= as.POSIXct("2025-06-25 16:34:00")) ~ "CHL22",
        (date_time >= as.POSIXct("2025-06-25 16:49:00") & date_time <= as.POSIXct("2025-06-25 16:54:00")) ~ "BG18",
        (date_time >= as.POSIXct("2025-06-25 18:29:00") & date_time <= as.POSIXct("2025-06-25 18:39:00")) ~ "CHL23B",
        (date_time >= as.POSIXct("2025-06-25 19:15:00") & date_time <= as.POSIXct("2025-06-25 19:30:00")) ~ "CHL24",
           
           
           TRUE ~ NA_character_
         ))

c3_jun2025_clean <- c3_jun2025 %>% 
  filter(!is.na(code))




```




+   Frequency distribution for June 2025 cruz C3 data

```{r}

c3_jun2025_clean %>% 
  filter(chl > 0) %>% 
  mutate(code = as.factor(code)) %>% 
ggplot(aes(x = chl)) +
  geom_histogram(bins = 30, binwidth = 2, boundary = 0, fill = "steelblue", color = "black") +
  #facet_wrap(~ code, scales="free_y") +
  labs(x = "Chlorophyll (RFU)", y = "Frequency") +
  theme_bw() +
  theme(text = element_text(size = 15, colour = "black"),
        axis.text.x = element_text(size=14, angle = 45, hjust = 1, vjust = 1))+
 scale_x_continuous(
  breaks = scales::pretty_breaks(n = 6)  #ggplot makes breaks based on data
)


ggsave(file="plots/c3/2025_plots/c3_2025JUNcruz_gen_freq_distribution.png", width = 10, height = 12, dpi = 300)

```



+ Removing outliers from June cruise data

```{r}

c3_Jun25_outlier_detect <- multidetect(data = c3_jun2025_clean, 
                                var = "chl", 
                                var_col = "code",
                                multiple = TRUE,
                                methods = c("adjbox", "iqr", "hampel","jknife",
                                            "seqfences", "mixediqr",
                                            "distboxplot", "semiqr",
                                            "zscore", "logboxplot", "medianrule"),
                       silence_true_errors = TRUE, 
                       missingness = 0.1,
                       sdm = FALSE,
                       na.inform = TRUE)
  
  

#Visualize the number of outliers detected by each method

ggoutliers(c3_Jun25_outlier_detect,
                        select = unique(c3_jun2025_clean$code),
                        color='purple',
                        desc = TRUE,
                        ncol = 10,
                        nrow = 10)

ggsave(file = "plots/c3/2025_plots/outliersJUN25_per_site.png", width = 10, height = 10, dpi = 300)


#Obtaining quality controlled dataset using loess method or data labeling


out_qc_loessJUN25 <- extract_clean_data(refdata = c3_jun2025_clean, 
            outliers = c3_Jun25_outlier_detect, loess = TRUE,
            var_col = "code"
            )  #set threshold to 0.8 to remove the very strong outliers only

#clean dataset
nrow(out_qc_loessJUN25)

#reference data
nrow(c3_jun2025_clean)

out_qc_labeledJUN25 <- classify_data(refdata = c3_jun2025_clean, 
                                  var_col = "code",
                                outliers = c3_Jun25_outlier_detect)


#Visualize labelled quality controlled dataset

ggenvironmentalspace(out_qc_labeledJUN25, 
                     type = '1D',
                     ggxangle = 45, 
                     ncol = 10,
                     nrow = 10,
                     scalecolor = 'viridis',
                     xhjust = 1,
                     legend_position = 'blank',
                     ylab = "Number of records",
                     xlab = "Outlier labels")


ggsave(filename = "plots/c3/2025_plots/filtered_siteJUN25_data.png", width = 14, height = 10, dpi = 300)



#labelled outlier or not 
labeledJUN25 <- classify_data(refdata = c3_jun2025_clean, 
                         var_col = "code",
                         outliers = c3_Jun25_outlier_detect)



#plot clean dataset

out_qc_loessJUN25 %>% 
mutate(code = as.factor(code)) %>% 
ggplot(aes(x = chl)) +
  geom_histogram(bins = 30, binwidth = 2, boundary = 0, fill = "steelblue", color = "black") +
facet_wrap(~ code)+
  labs(x = "Chlorophyll (ug/l)", y = "Frequency") +
  theme_bw() +
  theme(text = element_text(size = 10, colour = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))+
 scale_x_continuous(
  breaks = scales::pretty_breaks(n = 6)  #ggplot makes breaks based on data
)

ggsave(file="plots/2025_plots/c3_2025_loessfilter_sitesJUN_freq_distribution.png", width = 10, height = 10, dpi = 300)



```



+ Get final dataset

```{r}


final_crjun_2025_df <- out_qc_loessJUN25 %>% 
  rename(code_clean = code) %>% 
  group_by(code_clean, year) %>%
  summarise(
    chl = mean(chl, na.rm = TRUE),
    pc = mean(pc, na.rm = TRUE),
    temp = mean(temp, na.rm = TRUE),
    depth = mean(depth, na.rm = TRUE),
    source_file  = unique(source_file),
  ) 

```



+ Join c3 cruise datasets

```{r}

c3_cruises_data <- bind_rows(final_crjun_2023_df, final_crjul_2024_df, final_crnov_2024_df,    final_crapr_2025_df, final_crjun_2025_df) %>% 
  arrange(source_file) %>% 
  mutate(source_file = str_remove(source_file, "\\.csv$")) %>% 
  mutate(code_clean = str_trim(code_clean),          # Removes hidden spaces
         code_clean = str_to_upper(code_clean),       # Ensures 'chl14' matches 'CHL14'
         code_clean = str_remove(code_clean, "^.*_")) # Removes prefixes if they exist


```


+ c3 files with extracted chl

```{r}

file_paths.csv_extc3 <- dir("c3_extracted chls", "*.csv", full.names = TRUE)

c3_j2023_extchl <- read_csv(file_paths.csv_extc3[1]) %>%
    filter(!is.na(ext_chl)) %>% 
  mutate(code_clean = str_remove_all(code, " ")) %>% 
  select(code_clean, ext_chl, srp_ug_l, nitrates_ug_l, sd, source_file) %>% 
  mutate(sd = sd/100)


  
c3_2024_25_extchl <- read_csv(file_paths.csv_extc3[3])  %>% 
  remove_empty(which = "cols") %>% clean_names() %>% 
  filter(study %in% "CR") %>% 
  rename(code_clean = site,
         ext_chl = florometer_chl_ug_l) %>% 
  select(code_clean, ext_chl, srp_ug_l,nitrates_ug_l,an_new_sd_cm, an_old_sd_cm, pbm_new_sd_cm, pbm_old_sd_cm, source_file) %>% 
  arrange(code_clean) %>% 
mutate(sd = rowMeans(across(c(an_new_sd_cm, an_old_sd_cm, pbm_new_sd_cm, pbm_old_sd_cm)), na.rm = TRUE) / 100) %>% 
  dplyr::select(!c(an_new_sd_cm, an_old_sd_cm, pbm_new_sd_cm, pbm_old_sd_cm))



#Join the 2 files

c3_extchl_files <- bind_rows(c3_j2023_extchl, c3_2024_25_extchl) %>% 
  mutate(source_file = str_remove(source_file, "\\.csv$")) %>% # remove csv extension from file name, and $ means at the end of name to avoid accidental removals
mutate(code_clean = str_trim(code_clean),          # Removes hidden spaces
         code_clean = str_to_upper(code_clean),       # Ensures 'chl14' matches 'CHL14'
         code_clean = str_remove(code_clean, "^.*_")) %>% # Removes prefixes if they exist  
arrange(code_clean)



```

+ joining ext_chl with c3 sonde files

```{r}


c3_calib_data <- full_join(c3_extchl_files, c3_cruises_data, by = c("code_clean", "source_file")) %>% 
  mutate(
    source_file_clean = ifelse(
      source_file %in% c("c3_CR_27_June_2023", "Turner_C3_28_June2023_cruise"),
      "cruise-c3data-2023",
      source_file
    )
  ) 

 

# Find codes in chl that ARE NOT in ext_chl
mismatches <- anti_join(c3_calib_data, c3_extchl_files, by = "code_clean")

unique(mismatches$source_file_clean)


c3_calib_data_clean <- c3_calib_data %>% 
    filter(!is.na(ext_chl), !is.na(chl)) %>% 
  rename(code = code_clean)

```



# 2024 experiments

 + c3

```{r}


c3_spiking_expt24 <- exptsjul_24 %>% 
  filter(sos == "c3") %>% 
  mutate(code = site,
         chl = top_chl,
         pc = top_bga,
         file_name = "2024_c3sonde_experiment",
         data_type = "experiment") %>% 
  mutate(year = year(date_time)) %>% 
  dplyr::select(code, date_time, chl, pc, ext_chl, file_name, data_type)



```


+ Linear regression for chl_c3 ~ ext_chl

+   check slopes across years

```{r}

  c3_calib_data_clean %>% 
  arrange(source_file_clean, chl) %>% 
  ggplot(aes(x = chl, y = ext_chl)) +
  #facet_wrap(~source_file_clean) +  
geom_point(color = "steelblue", size = 3.5) +
  geom_smooth(method = "lm", se = FALSE, color = "red", linewidth = 1.5) +
  stat_regline_equation(
    aes(label = paste(after_stat(eq.label), ..rr.label.., sep = "~~~~")),
    size = 15,
    show.legend = FALSE
  ) +
  labs(x= "chl_c3 (RFU)", y = "extracted_chlorophyll (ug/l)")+
  theme_bw(base_size = 10) +
  theme(
    panel.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_line(colour = "gray90", linetype = "dotted", linewidth = 0.5),
    legend.position = "bottom",
      legend.text = element_text(size = 20),   # 👈 increases legend labels
  legend.title = element_text(size = 18),
  
    axis.text = element_text(size = 30, margin = margin(t = 15)),
    axis.title = element_text(size = 30),
    axis.text.x = element_text(vjust = 1, hjust = 1),
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
  ) +
  guides(color = guide_legend(override.aes = list(size = 8)))
  

ggsave(filename = "plots/CALIBc3/plot_c3chl~ext_gen_origdata.png", width = 20, height = 15, dpi = 300)
  
  
#model trial
model_orig <- lm(ext_chl ~ chl, data = c3_calib_data_clean)


plot(model_orig)


#Check prediction error explicitly with the original data

c3_calib_data_clean$pred <- predict(model_orig, newdata = c3_calib_data_clean)

ggplot(c3_calib_data_clean, aes(chl, ext_chl - pred)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed")



```

In the original data model, the residuals vs fiited plot shows points skewed to the left and not spread out. The  qqnorm line is more flattened. The scale location plot increasing variance or spread. The leverage plot shows some influential points. However, the ext_chl~ c3chl model shows: the intercept of .... 

The residuals/error plot shows that, low chl, residuals are somewhat tight (but still spread), at high chl, residuals explode (very large positive & negative errors). There is heteroscedasticity or non-constant variance of data points.

**Systematic bias**
Many residuals at low chl are slightly negative
→ model tends to overpredict at low values
Some very large positive residuals
→ model underpredicts high ext_chl

This suggests the relationship is not linear. Predictions get worse as chl increases. 
The nonlinearity is largely due to scale


try a residual plot
```{r}


model_log <- lm(log10(ext_chl) ~ log10(chl), data = c3_calib_data_clean)
model_sqrt <- lm(sqrt(ext_chl) ~ sqrt(chl), data = c3_calib_data_clean)


c3_calib_data_clean$pred_log <- predict(model_log, newdata = c3_calib_data_clean)
c3_calib_data_clean$pred_sqrt <- predict(model_sqrt, newdata = c3_calib_data_clean)

c3_calib_data_clean$resid_log <- log10(c3_calib_data_clean$ext_chl) - c3_calib_data_clean$pred_log
c3_calib_data_clean$resid_sqrt <- sqrt(c3_calib_data_clean$ext_chl) - c3_calib_data_clean$pred_sqrt


#log
ggplot(c3_calib_data_clean, aes(log10(chl), resid_log)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed")


ggsave(filename = "plots/CALIBc3/rship_ext~chlc3_logtransformed_residual plot.png", width = 20, height = 20, dpi = 300)



#sqrt
ggplot(c3_calib_data_clean, aes(sqrt(chl), resid_sqrt)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed")

ggsave(filename = "plots/CALIBc3/rship_ext~chlc3_sqrttransformed_residual plot.png", width = 20, height = 20, dpi = 300)


```


The log-log transformation
Random Scatter: The points are distributed like a "cloud" with no discernible shape (no curves, no slopes, no funnels). This confirms the linearity assumption is met (1:1).
Centered at Zero: The points are evenly balanced above and below the dashed line. This means your model is no longer systematically over- or under-estimating the values.
Homoscedasticity (Constant Variance): The vertical "spread" of the points is relatively consistent from left to right.

The sqrt-sqrt transformation
Heteroscedasticity is the main issue here... with some outliers influecing. The scatter is also not that great for linearity. 

**Data transformation**



+   log transformed visuals

```{r}

c3_calib_data_clean %>% 
  ggplot(aes(log10(chl), log10(ext_chl))) +
  geom_point(size = 3.5) +
  geom_smooth(method = "lm", se = FALSE, color = "red", linewidth = 1.5) +
  stat_regline_equation(
    aes(label = paste(after_stat(eq.label), ..rr.label.., sep = "~~~~")),
    size = 15,
    show.legend = FALSE
  ) +
  theme_bw(base_size = 16) +
  theme(
    panel.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_line(colour = "gray90", linetype = "dotted", linewidth = 0.5),
    legend.position = "bottom",
      legend.text = element_text(size = 20),   # 👈 increases legend labels
  legend.title = element_text(size = 18),
  
    axis.text = element_text(size = 30, margin = margin(t = 15)),
    axis.title = element_text(size = 30),
    axis.text.x = element_text(vjust = 1, hjust = 1),
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
  ) +
  guides(color = guide_legend(override.aes = list(size = 8)))
  

ggsave(filename = "plots/CALIBc3/plot_extchl~c3chl_logtransformed_gen.png", width = 20, height = 20, dpi = 300)




#facetted
c3_calib_data_clean %>% 
  ggplot(aes(log10(chl), log10(ext_chl), color = source_file_clean)) +
  geom_point(size = 4) +
  geom_smooth(method = "lm", se = FALSE, color = "red", linewidth = 1.5) +
  facet_wrap(~ source_file_clean, scales = "free_y") +
  stat_regline_equation(
    aes(label = paste(after_stat(eq.label), ..rr.label.., sep = "~~~~")),
    size = 6,
    show.legend = FALSE
  ) +
  theme_bw(base_size = 18) +    #for scaling facets so they all look the same
  theme(
    panel.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_line(colour = "gray90", linetype = "dotted", linewidth = 0.5),
    legend.position = "bottom",
    legend.text = element_text(size = 16),
    legend.title = element_text(size = 18),
    axis.text = element_text(size = 18, margin = margin(t = 15)),
    axis.title = element_text(size = 20),
    axis.text.x = element_text(vjust = 1, hjust = 1),
    strip.text = element_text(size = 16, face = "bold"),
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold")
  ) +
  guides(color = guide_legend(override.aes = list(size = 5)))
  
  

ggsave(filename = "plots/CALIBc3/plot_extchl~c3chl_logtransformed_facet.png", width = 20, height = 20, dpi = 300)

```


+ Log visuals transformed no June :)


```{r}


c3_calib_data_clean %>% 
  filter(!source_file_clean%in% "c3_CR_JUN2025") %>% 
  ggplot(aes(log10(chl), log10(ext_chl))) +
  geom_point(size = 3.5) +
  geom_smooth(method = "lm", se = FALSE, color = "red", linewidth = 1.5) +
  stat_regline_equation(
    aes(label = paste(after_stat(eq.label), ..rr.label.., sep = "~~~~")),
    size = 15,
    show.legend = FALSE
  ) +
  theme_bw(base_size = 16) +
  theme(
    panel.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_line(colour = "gray90", linetype = "dotted", linewidth = 0.5),
    legend.position = "bottom",
      legend.text = element_text(size = 20),   # 👈 increases legend labels
  legend.title = element_text(size = 18),
  
    axis.text = element_text(size = 30, margin = margin(t = 15)),
    axis.title = element_text(size = 30),
    axis.text.x = element_text(vjust = 1, hjust = 1),
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
  ) +
  guides(color = guide_legend(override.aes = list(size = 8)))
  

ggsave(filename = "plots/CALIBc3/plot_extchl~c3chl_logtransformed_NOJUNE.png", width = 20, height = 20, dpi = 300)




#facetted
c3_calib_data_clean %>% 
  filter(!source_file_clean%in% "c3_CR_JUN2025") %>% 
  ggplot(aes(log10(chl), log10(ext_chl), color = source_file_clean)) +
  geom_point(size = 4) +
  geom_smooth(method = "lm", se = FALSE, color = "red", linewidth = 1.5) +
  facet_wrap(~ source_file_clean, scales = "free_y") +
  stat_regline_equation(
    aes(label = paste(after_stat(eq.label), ..rr.label.., sep = "~~~~")),
    size = 6,
    show.legend = FALSE
  ) +
  theme_bw(base_size = 18) +    #for scaling facets so they all look the same
  theme(
    panel.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_line(colour = "gray90", linetype = "dotted", linewidth = 0.5),
    legend.position = "bottom",
    legend.text = element_text(size = 16),
    legend.title = element_text(size = 18),
    axis.text = element_text(size = 18, margin = margin(t = 15)),
    axis.title = element_text(size = 20),
    axis.text.x = element_text(vjust = 1, hjust = 1),
    strip.text = element_text(size = 16, face = "bold"),
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold")
  ) +
  guides(color = guide_legend(override.aes = list(size = 5)))
  
  

ggsave(filename = "plots/CALIBc3/plot_extchl~c3chl_logtransformed_facetNOJUNE.png", width = 20, height = 20, dpi = 300)




```


+   sqrt transformed visuals

```{r}

c3_calib_data_clean %>% 
  ggplot(aes(sqrt(chl), sqrt(ext_chl))) +
  geom_point(size = 3.5) +
  geom_smooth(method = "lm", se = FALSE, color = "red", linewidth = 1.5) +
  stat_regline_equation(
    aes(label = paste(after_stat(eq.label), ..rr.label.., sep = "~~~~")),
    size = 15,
    show.legend = FALSE
  ) +
  theme_bw() +
  theme(
    panel.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_line(colour = "gray90", linetype = "dotted", linewidth = 0.5),
    legend.position = "bottom",
      legend.text = element_text(size = 20),   # 👈 increases legend labels
  legend.title = element_text(size = 18),
  
    axis.text = element_text(size = 30, margin = margin(t = 15)),
    axis.title = element_text(size = 30),
    axis.text.x = element_text(vjust = 1, hjust = 1),
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
  ) +
  guides(color = guide_legend(override.aes = list(size = 8)))
  

ggsave(filename = "plots/CALIBc3/rship_c3chl~ext_sqrttransformedfacet.png", width = 30, height = 20, dpi = 300)



```



c3 log model with no env _ log transformed chl


```{r}
#Adding scale avoids collinearity issues and makes interactions interpretable.

#Since sd has many NAs, we will remove this variable from the chl calibration for now


c3_model_data <- c3_calib_data_clean %>%
  dplyr::select(!sd) %>% 
  mutate(
    log_ext_chl = log10(ext_chl), #response variable is not scaled
    log_chl     = log10(chl),
    log_pc      = log10(pc)
  ) %>%
  mutate(
    log_chl_c = scale(log_chl, scale = FALSE),  #this means subtract mean, but do NOT divide by sd
    log_pc_c  = scale(log_pc,  scale = FALSE),
    temp_c    = scale(temp,    scale = FALSE)
  )



```











+ Check model without June data
```{r}
#model no June

c3_model_data_nj <- c3_model_data %>% 
  filter(!source_file_clean %in% c("June 2025"))
model_nj <- lm(log_ext_chl ~ log_chl, data = c3_model_data_nj)
summary(model_nj)

par(mfrow = c(2,2))
plot(model_nj)



#Best model from all cruises without June

#model eqn with no June (y = 0.45x + 0.48)

# Manual Log Prediction: y = (Slope * x) + Intercept
c3_model_data$pred_chl_nj <- (0.45 * c3_model_data$log_chl_c) + 0.48

sigma_global <- summary(model_test)$sigma
bcf_ancova <- exp((sigma_global^2 / 2) * (log(10)^2))


```



TRY ANCOVA MODEL: check cruise interaction: see difference in cruises, are they significant?)

```{r}

model_test <- lm(log_ext_chl ~ log_chl * source_file_clean, data = c3_model_data)
summary(model_test)

#all data_only log
par(mfrow = c(2, 2))
plot(model_test)

```

Check cruises fit using ANCOVA (Analysis of Covariance)

+ The model uses Nov 2024 cruise as a baseline cruise and how other cruises deviate from this baseline. 

ANCOVA combines a continuous variable (log10(chl)) with a categorical variable (cruise_id or source_file_clean). It answers two main questions:

Intercept Shift: Does a specific cruise start at a higher or lower baseline? 
Slope Change: Does the relationship change between cruises? (How much Y changes with a unit increase in X..... since we are doing log10, we consider every 10^(x) increase in X) OR if the sensor's sensitivity (the rate of change) shifted across cruises. 

+ Main Effects (the cruises): These show if a cruise is shifted "up" or "down" (the Intercept). Does the cruise have a higher or lower "starting point?
+ Interaction Terms (the log10(chl):cruise rows): These show if the slope is steeper or flatter.


By using the November cruise as the reference "baseline," we can pinpoint exactly how other cruises deviate in two ways:

+Intercept shifts: Offset: in April 2025 and June 2025 both have significant p-values (0.016 and 0.002)
), meaning their readings are naturally shifted higher than the November baseline, regardless of the chlorophyll concentration

+Slope Change (Interaction Terms):While June 2023 and July 2024 follow the same slope as November, June 2025 is a major outlier. Its interaction term is highly significant (0.001), with a slope that is significantly flatter (by -0.58) than the baseline. For calculations, u have to either add or subtract the estimate of the indiv cruise from the baseline.

+ Better fit by accounting for the cruises. should we make a season or cruise corrected model? YES

+ While a simple linear regression assumes one rule applies to all, an ANCOVA (Analysis of Covariance) allows for different groups to have their own rules or starting point. So when we do the prediction , each cruise is treated independently. 

Conclusion: Because you are using a log-log model (both x and y arelog10 ), we talk in percentages.
For every 1% increase in c3_chl, ext_chl increases by b1*% (where by is your slope). E.g For April, if c3_Chl increases by 1%, ext_chl would increase by 0.58%.


+ Extract model coefficients  per cruise
```{r}

model_ancova <- model_test

# 1. Extract the coefficients
cf <- coef(model_ancova)

# 2. Identify the cruise names (excluding the baseline parts)

cruises <- names(cf)[grepl("source_file_clean", names(cf)) & !grepl("log10", names(cf))]
cruise_names <- gsub("source_file_clean", "", cruises)



# 3. Create the Baseline entry
results <- data.frame(
  Cruise = "Baseline (Reference)",
  Intercept = cf["(Intercept)"],
  Slope = cf["log_chl"]
)

# 4. Loop through and calculate the actual coefficients for the other cruises

for (i in seq_along(cruises)) {
  c_name <- cruise_names[i]
  
  # New Intercept = Baseline Intercept + Cruise Offset
  intercept_val <- cf["(Intercept)"] + cf[cruises[i]]
  
  # New Slope = Baseline Slope + Interaction Offset
  # We find the interaction term that matches the cruise name
  inter_term <- names(cf)[grepl("log_", names(cf)) & grepl(cruises[i], names(cf))]
  slope_val <- cf["log_chl"] + cf[inter_term]
  
  results <- rbind(results, data.frame(
    Cruise = c_name,
    Intercept = unname(intercept_val),
    Slope = unname(slope_val)
  ))
}



print(results)





```

+ Plot visuals

```{r}

#reorder cruises by month and year

#Use the ORIGINAL values as they appear in your data right now


original_values <- c(
  "cruise-c3data-2023", 
  "c3_CR_24JUL2024", 
  "c3_CR_NOV2024", 
  "c3_CR_APR2025", 
  "c3_CR_JUN2025"
)


# Define the chronological order manually
chron_order <- c(
  "June 2023", 
  "July 2024", 
  "November 2024", 
  "April 2025", 
  "June 2025"
)



# 3. Create the factor using both
c3_model_data$source_file_clean <- factor(
  c3_calib_data_clean$source_file_clean, 
  levels = original_values,  # What R looks for
  labels = chron_order     # What R displays on the plot
)


#plot coefficients
ggplot(c3_model_data, aes(x=log_chl, y=log_ext_chl, color = source_file_clean)) +
  geom_point(alpha = 0.5) + # Plots the actual data points with some transparency
  geom_smooth(method = "lm", se = FALSE) + # Adds the separate regression lines
  facet_wrap(~source_file_clean) +
 
  labs(
    title = "Calibration Curves by Cruise",
    x = "log10(Chl)",
    y = "log10(ext_chl)",
    color = "Cruise ID"
  ) +
  
  theme_bw(base_size = 18) +
  theme(
    panel.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_line(colour = "gray90", linetype = "dotted", linewidth = 0.5),
    legend.position = c(0.83, 0.22), # Moves legend to the empty 6th facet
      legend.text = element_text(size = 30),   # 👈 increases legend labels
  legend.title = element_text(size = 25),
  
    axis.text = element_text(size = 16, margin = margin(t = 15)),
    axis.title = element_text(size = 30),
    axis.text.x = element_text(vjust = 1, hjust = 1),
    plot.title = element_text(hjust = 0.5, size = 30, face = "bold"),
  ) +
  guides(color = guide_legend(override.aes = list(size = 8))) + 
  
  stat_regline_equation(
    aes(label = paste(after_stat(eq.label), after_stat(rr.label), sep = "~~~~")),
    size = 12,           
    label.y = 2.5,     
    show.legend = FALSE
  )



#General plot one equation

ggplot(c3_model_data, aes(x = log_chl, y = log_ext_chl)) +
  geom_point(alpha = 0.6, size = 2) + 
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.2) +
  #facet_wrap(~source_file_clean, ncol = 3) + # Ensures 3 columns
  scale_color_viridis_d(option = "plasma", end = 0.8) + # High contrast but professional
  stat_regline_equation(
    aes(label = paste(after_stat(eq.label), after_stat(rr.label), sep = "~~~~")),
    size = 8, label.y.npc = 1 # Keeps text away from the very top
  ) +
  labs(
    #title = "Calibration Curves by Cruise",
    x = "log10(c3_Chl(RFU))",
    y = "log10(extracted chlorophyll_a (ug/l))")+
    
  theme_bw(base_size = 30) +
  theme(
    legend.position = c(0.83, 0.22), #The "Magic" positioning
    legend.title = element_text(size = 22),
    legend.text = element_text(size = 25),
    panel.grid.minor = element_blank(),
    strip.background = element_rect(fill = "gray95"), # Cleaner facet tabs
    strip.text = element_text(size = 30, face = "bold"),
    
    axis.text = element_text(size = 18, margin = margin(t = 15)),
    axis.title = element_text(size = 30),
    axis.text.x = element_text(vjust = 1, hjust = 1),
    plot.title = element_text(hjust = 0.5, size = 30, face = "bold")
    ) 
   #guides(color = guide_legend(override.aes = list(size = 10))) 
  

ggsave(filename = "plots/CALIBc3/GLOBAL_plot_c3chl~ext_gen_logdata.png", width = 20, height = 20, dpi = 300)

#####################

#ANCOVA General plot 2 with all 5 equations

ggplot(c3_model_data, aes(x = log10(chl), y = log10(ext_chl), color = source_file_clean)) +
  geom_point(alpha = 0.6, size = 2) + 
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.2) +
  #facet_wrap(~source_file_clean, ncol = 3) + # Ensures 3 columns
  scale_color_viridis_d(option = "plasma", end = 0.8) + # High contrast but professional
  stat_regline_equation(
    aes(label = paste(after_stat(eq.label), after_stat(rr.label), sep = "~~~~")),
    size = 8, label.y.npc = 1 # Keeps text away from the very top
  ) +
  labs(
    #title = "Calibration Curves by Cruise",
    x = "log10(c3_Chl(RFU))",
    y = "log10(extracted chlorophyll_a (ug/l))")+
    
  theme_bw(base_size = 20) +
  theme(
    legend.position = c(0.83, 0.22), #The "Magic" positioning
    legend.title = element_text(size = 25),
    legend.text = element_text(size = 28),
    panel.grid.minor = element_blank(),
    strip.background = element_rect(fill = "gray95"), # Cleaner facet tabs
    strip.text = element_text(size = 30, face = "bold"),
    
    axis.text = element_text(size = 20, margin = margin(t = 15)),
    axis.title = element_text(size = 30),
    axis.text.x = element_text(vjust = 1, hjust = 1),
    plot.title = element_text(hjust = 0.5, size = 30, face = "bold")
    ) +
   guides(color = guide_legend(override.aes = list(size = 16))) 
  



ggsave(filename = "plots/CALIBc3/coff_plot_ext~c3chl_gen_logdata.png", width = 20, height = 20, dpi = 300)



##################################################


#cleaner plot_Facetted plots
ggplot(c3_model_data, aes(x = log10(chl), y = log10(ext_chl), color = source_file_clean)) +
  geom_point(alpha = 0.6, size = 2) + 
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.2) +
  facet_wrap(~source_file_clean, ncol = 3) + # Ensures 3 columns
  scale_color_viridis_d(option = "plasma", end = 0.8) + # High contrast but professional
  stat_regline_equation(
    aes(label = paste(after_stat(eq.label), after_stat(rr.label), sep = "~~~~")),
    size = 8, label.y.npc = 1 # Keeps text away from the very top
  ) +
  labs(
    title = "Calibration Curves by Cruise",
    x = "log10(c3_Chl(RFU))",
    y = "log10(extracted chlorophyll_a (ug/l))")+
    
  theme_bw(base_size = 16) +
  theme(
    legend.position = "none",
    #legend.title = element_text(size = 22),
    #legend.text = element_text(size = 25),
    panel.grid.minor = element_blank(),
    strip.background = element_rect(fill = "gray95"), # Cleaner facet tabs
    strip.text = element_text(size = 30, face = "bold"),
    
    axis.text = element_text(size = 18, margin = margin(t = 15)),
    axis.title = element_text(size = 30),
    axis.text.x = element_text(vjust = 1, hjust = 1),
    plot.title = element_text(hjust = 0.5, size = 30, face = "bold")
    ) 
  
 # guides(color = guide_legend(override.aes = list(size = 10))) 
  



ggsave(filename = "plots/CALIBc3/coff_plot_c3chl~ext_facet_logdata.png", width = 20, height = 20, dpi = 300)
  


```


The global plot **one size fits all** is not the best because it is being pulled by outlier cruises. Because cruises like April and June 2025 have very high values even at low chlorophyll (the "higher starting points" we discussed), they drag the entire blue line upward. The slope in this image is "flatter" because it is trying to compromise between the good cruises and that very flat June cruise. This is exactly why the R is so low (0.22)). The line doesn't actually represent any of the cruises well; it just goes through the middle of the mess. 



Check diagnostics for ANCOVA model

```{r}
library(car)


#check for multicollinearity

vif_results <- vif(model_ancova, type = "predictor")
print(vif_results)

#vif(model_test) #If VIF > 5–10 → variables are overlapping. Which is not the case here


#check residuals normality..etc
par(mfrow = c(2, 2))
plot(model_2)

```

For interaction terms, look at GVIF^(1/2*F)... VIF is relatively okay <5. So our model doesn't have redundant predictors.



c3 model with env: Does the slope of chl → ext_chl change with environment?

Is my sonde calibration stable across environmental gradients?

+ check visuals with env
```{r}


### check r/ships with ext_chl_Need for data transformation?

ggplot(c3_model_data, aes(temp, log10(ext_chl))) + geom_point()
ggplot(c3_model_data, aes(pc, log10(ext_chl))) + geom_point()
ggplot(c3_model_data, aes(sd, log10(ext_chl))) + geom_point()


#sd relationship _Yes, log trasnform for sd and pc. No need for temp
ggplot(c3_model_data, aes(sd, log10(ext_chl))) +
  geom_point() +
  geom_smooth(method = "loess", se = FALSE, color = "red")



#Plot slopes by environment....... sd seems to show some interaction coz the slope at high sd changes
c3_model_data %>% 
  filter(!is.na(sd)) %>% 
ggplot(aes(log_chl, log_ext_chl)) +
  geom_point() +
  geom_smooth(method = "lm") +
  facet_wrap(~ cut(log10(sd), 3))

```

+ models with interaction

```{r}
#Fit FULL interaction model (Main effects: chl, pc, temp and sd).. plus interaction chl*pc, chl*sd, chl*temp) * cruise



# 1. Build the Complex Interaction Model
# This formula includes 3-way interactions between Chlorophyll, 
# environmental factors, and the Cruise ID.
model_int <- lm(
  log_ext_chl ~ log_chl_c * (log_sd_c + log_pc_c + temp_c) * source_file_clean,
  data = c3_model_data
)



# 2. View the full results
# Warning: This will be a very long table of coefficients!
summary(model_int)

# 3. Check for Multicollinearity
# With this many interactions, expect high GVIF values.
library(car)
vif(model_int, type = "predictor")
```

compare models: ANCOVA Simpler vs Interaction model
```{r}
# 4. Compare to your previous ANCOVA model


# 1. Create a "complete cases" subset
# This keeps only rows where ALL these variables are present
df_clean_subset <- c3_model_data %>%
  filter(!is.na(log_ext_chl), 
         !is.na(log_chl_c), 
         !is.na(log_sd_c), 
         !is.na(log_pc_c), 
         !is.na(temp_c), 
         !is.na(source_file_clean))

# 2. Re-run both models on this IDENTICAL dataset
model_ancova_subset <- lm(log_ext_chl ~ log_chl_c * source_file_clean, 
                          data = df_clean_subset)

model_int_subset <- lm(log_ext_chl ~ log_chl_c * (log_sd_c + log_pc_c + temp_c) * source_file_clean, 
                       data = df_clean_subset)

# 3. Now run the comparison
anova(model_ancova_subset, model_int_subset)


AIC(model_ancova_subset, model_int_subset)


# # This calculates the total 'impact' (Sum of Squares) for each factor
influence_test <- Anova(model_int_subset, type = "II")
print(influence_test)



```

check diagnostics for ANCOVA int model
```{r}

par(mfrow = c(2, 2))
plot(model_int)


```


+ Predicted ext chl using ANCOVA (internal training)
```{r}



# BCF for the ANCOVA Model
model_ancova <- lm(log_ext_chl ~ log_chl * source_file_clean, data = c3_model_data)

c3_model_data$pred_ancova <- predict(model_ancova, newdata = c3_model_data)
sigma_ancova <- summary(model_ancova)$sigma
bcf_ancova <- exp((sigma_ancova^2 / 2) * (log(10)^2))
c3_model_data$pred_extchl_ancova <- (10^c3_model_data$pred_ancova) * bcf_ancova


```




Compare models with the simpler model (sd inclusive)

```{r}

anova(model_ancova_subset, model_int_subset)


AIC(model_ancova_subset, model_int_subset)

```


The AICs in the interaction model is lower than in the simpler model but it only slightly improves the R2 of our chl prediction, and yet it takes up more 27 degrees of freedom.


+ Try a GAM : Leave this code here for now
```{r}


gam_model <- gam(ext_chl ~ s(chl), data = c3_calib_data)

summary(gam_model)
plot(gam_model)


gam_log <- gam(log(ext_chl) ~ s(log(chl)), data = c3_calib_data)

summary(gam_log)
plot(gam_log)







```
GAM is detecting the same issue of nonlinearity but since log transformation fixed the diagnostics for linear data, then we can just use the lm from log transformation.

**effective degrees of freedom**
edf ≈ 1 → relationship is linear
edf > 1 → relationship is nonlinear (curved)
edf >> 1 (e.g., 4–8) → very wiggly → possible overfitting


Because edf ~ 1, more evidence to stick with the LM

**P-value**
The pvalue is < 0.05, which would imply that the chl explains the ext_chl

The deviance explained is 21.8% (model fit).

########################################################################################

+ bias correction on the c3 data we used for calibration.

    + Log transformation compresses large values
    + When you back-transform, you don’t fully recover the spread
    + So predictions are biased low
    + The correction inflates them slightly to compensate so that the high values are not under   estimated
    +10^(pred_log) → gives you the median prediction but it is a biased estimate of the mean..This raises 10 to the power of every number in that column, effectively "un-logging" the data 


+  correct the bias to the series using sigma (if residuals are normally distributed) or Duan's smearing estimator if residuals are not normally dist. Duan's estimator is more robust and doesn't assume perfect normal residues... but our log transformed data has normal.. sigma2 is the residual variance of the model
    
+ Predicted values on training data / internal validation

    + Steps: Predict values in log_space, back transform to the original scale, and then add a bias correction term
    
    
+   log model _ No interaction

```{r}


model_global <- lm(log_ext_chl ~ log_chl, data = c3_model_data)

c3_model_data$pred_log <- predict(model_global, newdata = c3_model_data)

sigma_log <- summary(model_global)$sigma^2

bcf_log <- exp((sigma_log^2 / 2) * (log(10)^2)) #corrects log bias

c3_model_data$pred_ext_chl_global <- 10^(c3_model_data$pred_log) * bcf_log



# Calculate metrics

library(Metrics)

rmse(c3_model_data$ext_chl, c3_model_data$pred_ext_chl_global) #residual mean of squared erros
mae(c3_model_data$ext_chl, c3_model_data$pred_ext_chl_global) #mean abs error


# Plot observed vs predicted


plot(x = c3_model_data$chl, y = c3_model_data$pred_ext_chl_global)

library(ggplot2)

ggplot(c3_model_data, aes(x = ext_chl, y = pred_ext_chl_global)) +
  
  # Points
  geom_point(size = 3, alpha = 0.7, color = "black") +
  
  # 1:1 line (perfect prediction)
  geom_abline(slope = 1, intercept = 0, 
              color = "red", linetype = "dashed", linewidth = 1.2) +
  
  # Regression line (your model fit)
  geom_smooth(method = "lm", se = FALSE, 
              color = "blue", linewidth = 1.5) +
  
  labs(
    x = "Observed Chlorophyll (µg/L)",
    y = "Predicted Chlorophyll (µg/L)",
    title = "Calibration of Sonde Chlorophyll"
  ) +
  
  theme_bw(base_size = 18) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_line(colour = "gray90", linetype = "dotted"),
    plot.title = element_text(hjust = 0.5, face = "bold")
  )


ggsave(file="plots/CALIBc3/plot_av_predictedvsobs_extchl_regression.png", width = 16, height = 10, dpi = 300)




```

*Notes
sigma^2 in your R output is calculated based on log10 steps."
exp expects "natural log" steps.
The (log(10)^2) (which is about 5.3) acts as the translator between the two.
If you don't include that your correction will be about 5 times smaller than it should be!




Mixed effects model?

```{r}

c3_model_data$source_file_clean <- relevel(
  factor(c3_model_data$source_file_clean), 
  ref = "c3_CR_NOV2024" 
)


#using uncentered log data


# 1. Run the Fixed Effects Model (lm)
# This treats the 'Standard' vs 'Outlier' as hard categories
model_fixed <- lm(log_ext_chl ~ log_chl + source_file_clean, data = c3_model_data)
summary(model_fixed)

# 2. Run the Mixed Effects Model (lmer)
# This treats each cruise as a random sample from a larger population
library(lme4)
model_mixed <- lmer(log_ext_chl ~ log_chl + (1 | source_file_clean), data = c3_model_data)
summary(model_mixed)




```

+ Predict values for lmer model (accounts for cruise as random effect)

```{r}


# 1. Extract Fixed Effects and Sigma
coeffs <- fixef(model_mixed)
b0 <- coeffs["(Intercept)"]
b1 <- coeffs["log_chl"]
sig <- sigma(model_mixed)



# 2. Calculate Bias Correction Factor (BCF)
bcf_mm <- exp((sig^2 / 2) * (log(10)^2))

# 3. Predict for your dataset
c3_model_data <- c3_model_data %>%
  mutate(
    log_pred_mm   = b0 + (b1 * log_chl),  #predict data using the model
    pred_ext_mm = (10^log_pred_mm) * bcf_mm   #back transform
  )


```

+ Plot all the predicted values by the various models

```{r}


c3_model_data_long <- c3_model_data %>% 
  pivot_longer(
    cols = c(ext_chl, pred_ext_chl_global, pred_extchl_ancova,pred_ext_mm),
    names_to = "model",
               values_to = "chlorophyll") %>%
mutate(type = ifelse(model == "ext_chl", "Observed", "Predicted"))



ggplot(
  c3_model_data,
  aes(x = ext_chl)
) +
  geom_point(aes(y = pred_ext_chl_global, color = "Global model"), size = 2) +
  geom_point(aes(y = pred_extchl_ancova, color = "ANCOVA model"), size = 2) +
  geom_point(aes(y = pred_ext_mm, color = "Mixed model"), size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  

  labs(
    x = "Observed Extracted Chlorophyll",
    y = "Predicted Extracted Chlorophyll",
    color = "Model"
  ) +
  theme_bw(base_size = 10) +
  theme(
    panel.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_line(colour = "gray90", linetype = "dotted", linewidth = 0.5),
    legend.position = "right",
      legend.text = element_text(size = 12),   # 👈 increases legend labels
  legend.title = element_text(size = 14),
  
    axis.text = element_text(size = 16, margin = margin(t = 15)),
    axis.title = element_text(size = 14),
    axis.text.x = element_text(vjust = 1, hjust = 1),
    #plot.title = element_text(hjust = 0.5, size = 30, face = "bold"),
  ) 

#+  guides(color = guide_legend(override.aes = list(size = 8))) + 
  


ggsave(file="plots/CALIBc3/plot_models_predictedvsobs_extchl.png", width = 16, height = 10, dpi = 300)




# Calculate metrics

library(Metrics)

rmse_global <- rmse(c3_model_data$ext_chl, c3_model_data$pred_ext_chl_global) #residual mean of squared erros
mae_global <- mae(c3_model_data$ext_chl, c3_model_data$pred_ext_chl_global) #mean abs error

rmse_ancova <- rmse(c3_model_data$ext_chl, c3_model_data$pred_extchl_ancova) #residual mean of squared erros
mae_ancova <- mae(c3_model_data$ext_chl, c3_model_data$pred_extchl_ancova) #mean abs error

rmse_mm <- rmse(c3_model_data$ext_chl, c3_model_data$pred_ext_mm) #residual mean of 
mae_mm <- mae(c3_model_data$ext_chl, c3_model_data$pred_ext_mm) #mean abs error


rmse_table <- tibble(rmse_global,mae_global, rmse_ancova, mae_ancova, rmse_mm, mae_mm)


```

Notes:
ANCOVA (Red): Because it knows the specific cruise for every point, it stays closest to the 1:1 line, especially at higher concentrations. It is your most accurate model for the data you already have.
The Global (Green) and Mixed (Blue) are "The Averages": They sit on top of each other because they are both trying to find one universal rule. You can see them start to "peel away" from the 1:1 line and underestimate once observed chlorophyll gets above 25ug/l
.
The High-End Crash: the far right (observed ~85ug/l). Every model failed here, but the ANCOVA (red) was the only one that even tried to get close. This shows that at very high concentrations, the simple linear relationship is struggling.




+ Zoom in at points in the 0-20ug/l range where most points fall

```{r}



ggplot(
  c3_model_data,
  aes(x = ext_chl)
) +
  geom_point(aes(y = pred_ext_chl_global, color = "Global model"), size = 2) +
  geom_point(aes(y = pred_extchl_ancova, color = "ANCOVA model"), size = 2) +
  geom_point(aes(y = pred_ext_mm, color = "Mixed model"), size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  
coord_cartesian(xlim = c(0, 25), ylim = c(0, 25)) +
  scale_color_manual(values = c("ANCOVA model" = "#F8766D", 
                                "Global model" = "#00BA38", 
                                "Mixed model"  = "#619CFF")) +
  
  
  labs(
    x = "Observed Extracted Chlorophyll",
    y = "Predicted Extracted Chlorophyll",
    color = "Model"
  ) +
  theme_bw(base_size = 10) +
  theme(
    panel.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_line(colour = "gray90", linetype = "dotted", linewidth = 0.5),
    legend.position = "right",
      legend.text = element_text(size = 12),   # 👈 increases legend labels
  legend.title = element_text(size = 14),
  
    axis.text = element_text(size = 16, margin = margin(t = 15)),
    axis.title = element_text(size = 14),
    axis.text.x = element_text(vjust = 1, hjust = 1)
    #plot.title = element_text(hjust = 0.5, size = 30, face = "bold"),
  ) 

#+  guides(color = guide_legend(override.aes = list(size = 8))) + 
  


ggsave(file="plots/CALIBc3/plot_models_Zoomedpredictedvsobs_extchl.png", width = 16, height = 10, dpi = 300)






 
  


```


The "Vertical Stacks": Notice how the Blue (Mixed) and Green (Global) dots often form vertical lines. This is because those models only "see" the sonde value. If two different samples on two different days had the same sonde reading, these models give them the exact same prediction, even if the lab results were miles apart.

The "Floor" Effect: There is a heavy cluster of predictions between 5 and 10ug/l. The models are very hesitant to predict anything lower than 5, even when the observed lab data is near 2.

Overestimation at the Low End: Below an observed value of 5, almost all the dots (including the red ANCOVA ones) are above the dashed line. The models are "optimistic" at low concentrations, likely due to the intercept and the BCF multiplier.

ANCOVA (Red) still wins: See the red dots "stretching" better to follow the diagonal. It’s the only model that successfully uses the "cruise context" to move the prediction closer to the truth.


+ RMSE for the Zoomed in range

```{r}

# 1. Filter the data to the 0-20 observed range

model_wide_zoom <- c3_model_data_long %>%
  pivot_wider(names_from = model, values_from = chlorophyll) %>%
  # Filter for the "Zoom" range based on actual lab data
  filter(ext_chl < 25)





# 2. Calculate RMSE for each model in this subset
rmse_zoom <- model_wide_zoom %>%
  summarise(
    RMSE_ANCOVA = sqrt(mean((ext_chl - pred_extchl_ancova)^2, na.rm = TRUE)),
    RMSE_Global = sqrt(mean((ext_chl - pred_ext_chl_global)^2, na.rm = TRUE)),
    RMSE_Mixed  = sqrt(mean((ext_chl - pred_ext_mm)^2, na.rm = TRUE))
  )

print(rmse_zoom)


rmse_global <- rmse(model_wide_zoom$ext_chl, model_wide_zoom$pred_ext_chl_global) #residual mean of squared erros
mae_global <- mae(model_wide_zoom$ext_chl, model_wide_zoom$pred_ext_chl_global) #mean abs error

rmse_ancova <- rmse(model_wide_zoom$ext_chl, model_wide_zoom$pred_extchl_ancova) #residual mean of squared erros
mae_ancova <- mae(model_wide_zoom$ext_chl, model_wide_zoom$pred_extchl_ancova) #mean abs error

rmse_mm <- rmse(model_wide_zoom$ext_chl, model_wide_zoom$pred_ext_mm) #residual mean of 
mae_mm <- mae(model_wide_zoom$ext_chl, model_wide_zoom$pred_ext_mm) #mean abs error


rmse_table_zoom <- tibble(rmse_global,mae_global, rmse_ancova, mae_ancova, rmse_mm, mae_mm)
print(rmse_table_zoom)

```



# Model validation (Mixed effects model)

+   Test/Train 

```{r}

#library(boot)

#cv.glm(c3_calib_data_clean, model_log, K = 10)$delta


library(dplyr)

set.seed(123)

K <- 10
folds <- sample(rep(1:K, length.out = nrow(c3_model_data)))

cv_results <- data.frame()

for (k in 1:K) {
  
  train <- c3_model_data[folds != k, ]
  test  <- c3_model_data[folds == k, ]
  
# 1. Extract Fixed Effects and Sigma

model <- model_mixed
coeffs <- fixef(model_mixed)
b0 <- coeffs["(Intercept)"]
b1 <- coeffs["log_chl"]
sig <- sigma(model_mixed)


# 2. Calculate Bias Correction Factor (BCF)
bcf_mm <- exp((sig^2 / 2) * (log(10)^2))

# 3. Predict for your dataset
test <- test %>%
  mutate(
    log_pred_mm   = b0 + (b1 * log_chl),  #predict data using the model
    pred_ext_mm = (10^log_pred_mm) * bcf_mm   #back transform
  )

  
  # Store results
  cv_results <- rbind(cv_results, test)
}

# Final RMSE: the square root of the average of squared differences between observed and predicted values
rmse_mm_2 <- rmse(cv_results$ext_chl, cv_results$pred_ext_mm)



```

The root mean square error between the actual ext chl and predicted is `r rmse_mm_2` ug/l. It is the same as the rmse from before with all the values. Put simply, it shows, on average, how far the predictions are from the actual values, in the same scale as the outcome variable. 



+ Let's try to break down the chl ranges


```{r}

median_chl <- median(cv_results$chl, na.rm = TRUE)

cv_results <- cv_results %>%
  mutate(chl_group = ifelse(chl <= median_chl, "Low", "High"))


# compare rmse per group

library(dplyr)
library(Metrics)

cv_results %>%
  group_by(chl_group) %>%
  summarise(
    RMSE = rmse(ext_chl, pred_ext_chl),
    MAE  = mae(ext_chl, pred_ext_chl),
    n    = n()
  )


```


At low chl (less than median chl), `r median_chl`, the RMSE is lower while at higher chl, the RMSE increases to 115.


Leave one cruise out cross validation (LOOCV)_____try this again

```{r}

library(caret)


#New try

library(dplyr)
library(Metrics)

# 1. Clean data 

# 2. Get unique cruises
cruises <- unique(c3_calib_data_clean$source_file)

# 3. Initialize results storage
cv_cruise_results <- data.frame()

# 4. LOOCV loop (leave one cruise out)
for (c in cruises) {
  
  train <- c3_calib_data_clean %>% filter(cruises != c)
  test  <- c3_calib_data_clean %>% filter(cruises == c)
  
  # Skip if not enough data
  if (nrow(train) < 5 | nrow(test) < 2) next
  
  # Fit model (log-log)
  model <- lm(log10(ext_chl) ~ log10(chl), data = train)
  
  # Predict on left-out cruise
  test$pred_log <- predict(model, newdata = test)
  
  # Bias correction
  sigma2 <- summary(model)$sigma^2
  
  test$pred_ext_chl_2 <- exp((test$pred_log) + (sigma2 / 2))
  
  # Store cruise label
  test$fold <- c
  
  # Combine results
  cv_cruise_results <- bind_rows(cv_cruise_results, test)
}


#Overall performance
rmse(cv_cruise_results$ext_chl, cv_cruise_results$pred_ext_chl_2)

#Performance by cruise
cv_cruise_results %>%
  group_by(fold) %>%
  summarise(
    RMSE = rmse(ext_chl, pred_ext_chl_2),
    MAE  = mae(ext_chl, pred_ext_chl_2),
    n    = n()
  )




```




+ Apply to long term c3 data series



#   Loading longterm data series for model testing and calibration

+   c3 data
```{r}

file_paths.csv <- dir("c3_Dunga_time_series", pattern = "*.csv", full.names = TRUE)

first_file <- read_csv(file_paths.csv[1])
colnames(first_file)

#use function to read files

#read files

c3_dunga_files <- purrr::map_dfr(file_paths.csv, read_c3_file) 


c3_dunga_files_clean <- c3_dunga_files  %>% 
  mutate(
    date_time = as.character(date_time),
    date_time_parsed = parse_date_time(
      date_time,
      orders = c(
  "Ymd HMS",
  "Ymd HM",
  "dmy HMS",
  "dmy HM",
  "dmY HMS",
  "mdy HMS",
  "mdy HM"
))) %>%
  mutate(date = date(date_time_parsed),
         year= year(date_time_parsed),
         month = month(date_time_parsed),
         mmyy = format(date_time_parsed, "%m-%y")) %>% 
  arrange(date_time) %>% 
  clean_names() 


#check files
table(c3_dunga_files_clean$source_file)




```


The longterm data series are within (1.01 - 200.73) calibration range compared to the c3_chl_data (2.55 - 218.5). From the Dunga Sonde, I filtered out values <1. The values close to zero seemed like non-data.


+ frequency distribution


```{r}

c3_dunga_files_clean %>% 
  mutate(code = as.factor(source_file)) %>% 
ggplot(aes(x = chl)) +
  geom_histogram(bins = 30, binwidth = 2, boundary = 0, fill = "steelblue", color = "black") +
  #facet_wrap(~ code, scales="free_y") +
  labs(x = "Chlorophyll (RFU)", y = "Frequency") +
  theme_bw() +
  theme(text = element_text(size = 15, colour = "black"),
        axis.text.x = element_text(size=14, angle = 45, hjust = 1, vjust = 1))+
 scale_x_continuous(
  breaks = scales::pretty_breaks(n = 6)  #ggplot makes breaks based on data
)


ggsave(file="plots/c3/DG_plots/gen_freq_distribution.png", width = 10, height = 12, dpi = 300)

```

+ monthly plots for raw data

```{r}

# get unique mmyy values (ordered chronologically if needed)
mmyy_levels_raw <- c3_dunga_files_clean %>%
  mutate(mmyy_date = lubridate::my(mmyy)) %>%
  arrange(mmyy_date) %>%
  pull(mmyy) %>%
  unique()

# loop
for (m in mmyy_levels_raw) {
  
  df_sub <- c3_dunga_files_clean %>%
    filter(mmyy == m) %>%
    arrange(date_time)
  
  p <- ggplot(df_sub, aes(x = date_time, y = chl)) +
    geom_line(color = "steelblue") +
    geom_point(size = 0.5, alpha = 0.5) +
   # geom_point(aes(color = chl > 200), size = 0.7)+
    geom_smooth(se = FALSE, color = "red", span = 0.2)+
    labs(
      title = paste("Chlorophyll Time Series -", m),
      x = "Date-Time",
      y = "Chlorophyll (µg/L)"
    ) +
    theme_bw() +
    theme(
      text = element_text(size = 10, colour = "black"),
      axis.text.x = element_text(angle = 45, hjust = 1)
    )
  
  ggsave(
    filename = paste0("DG_plots/raw_monthly_chl_plots/chl_timeseries_", m, ".png"),
    plot = p,
    width = 8,
    height = 4,
    dpi = 300
  )
}




```




+   remove outliers

```{r}
c3_dg_outlier_detect <- multidetect(data = c3_dunga_files_clean, 
                                var = "chl", 
                                var_col = "mmyy",
                                multiple = TRUE,
                                methods = c("adjbox", "iqr", "hampel","jknife",
                                            "seqfences", "mixediqr",
                                            "distboxplot", "semiqr",
                                            "zscore", "logboxplot", "medianrule"),
                       silence_true_errors = TRUE, 
                       missingness = 0.1,
                       sdm = FALSE,
                       na.inform = TRUE)
  
  

#Visualize the number of outliers detected by each method

ggoutliers(c3_dg_outlier_detect,
                        select = unique(c3_dunga_files_clean$mmyy),
                        color='purple',
                        desc = TRUE,
                        ncol = 10,
                        nrow = 10)

ggsave(file = "plots/c3/DG_plots/outliers_per_month.png", width = 25, height = 15, dpi = 300)


#Obtaining quality controlled dataset using loess method or data labeling


out_qc_loessdgc3 <- extract_clean_data(refdata = c3_dunga_files_clean, 
            outliers = c3_dg_outlier_detect, loess = TRUE,
            var_col = "mmyy"
            )  #set threshold to 0.8 to remove the very strong outliers only

#clean dataset
nrow(out_qc_loessdgc3)

#reference data
nrow(c3_dunga_files_clean)

out_qc_labeledgdc3 <- classify_data(refdata = c3_dunga_files_clean, 
                                  var_col = "mmyy",
                                outliers = c3_dg_outlier_detect)


#Visualize labelled quality controlled dataset

ggenvironmentalspace(out_qc_labeledgdc3, 
                     type = '1D',
                     ggxangle = 45, 
                     ncol = 10,
                     nrow = 10,
                     scalecolor = 'viridis',
                     xhjust = 1,
                     legend_position = 'blank',
                     ylab = "Number of records",
                     xlab = "Outlier labels")


ggsave(filename = "plots/c3/DG_plots/filtered_month_data.png", width = 25, height = 15, dpi = 300)



#labelled outlier or not 
labeleddgc3 <- classify_data(refdata = c3_dunga_files_clean, 
                         var_col = "mmyy",
                         outliers = c3_dg_outlier_detect)


##Final cleaning of cleaned dataset

weird_chl <- out_qc_loessdgc3 %>% 
  filter(chl < 1)



wierd_dg <- out_qc_loessdgc3 %>% 
  mutate(
    bad_sensor = (chl < 1) |
         (chl < 2 & mmyy == "03-24") |
         (date_time_parsed >= as.POSIXct("2024-10-16 00:00:00") &
         date_time_parsed <= as.POSIXct("2024-10-30 23:59:59")))

  
final_dg <- wierd_dg %>% 
  filter(!bad_sensor)
  
  
  

#plot clean dataset

final_dg %>% 
  mutate(mmyy_date = my(mmyy),   # converts "01-25" → Jan 2025
  mmyy = factor(mmyy, levels = unique(mmyy[order(mmyy_date)]))       
  ) %>% 
mutate(code = as.factor(source_file)) %>% 
  filter(mmyy == "01-25") %>% 
ggplot(aes(x = chl)) +
  geom_histogram(bins = 30, binwidth = 2, boundary = 0, fill = "steelblue", color = "black") +
facet_wrap(~ mmyy, scales= "free_y")+
  labs(x = "Chlorophyll (ug/l)", y = "Frequency") +
  theme_bw() +
  theme(text = element_text(size = 10, colour = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))+
 scale_x_continuous(
  breaks = scales::pretty_breaks(n = 6)  #ggplot makes breaks based on data
)

ggsave(file="plots/c3/DG_plots/loessfilter_freq_distribution.png", width = 10, height = 10, dpi = 300)



#Plots frequency probability

out_qc_loessdgc3 %>% 
  ggplot(aes(x = chl)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30, binwidth = 2, boundary = 0,
                 fill = "steelblue", color = "black") +
  facet_wrap(~ mmyy, scales = "free_y")

```


clean monthly distribtution plots for clean data

```{r}

final_dg %>% 
  mutate(
    month = as_factor(month),
    year = as_factor(year)
  ) %>%
  arrange(date_time) %>% 
  group_by(month, year) %>%
  summarise(avg_chl = mean(chl), .groups = "drop") %>% 
  ggplot(aes(x = month, y = avg_chl, colour = year, group = year)) +
  geom_line(size = 1) +
  geom_point(size = 3) +
  labs(
    x = "Months",
    y = "Chlorophyll (µg/L)",
    color = "Year"
  ) +
 
theme_bw() +
  theme(
    panel.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_line(colour = "gray90", linetype = "dotted", linewidth = 0.5),
    legend.position = "bottom",
      legend.text = element_text(size = 20),   # 👈 increases legend labels
  legend.title = element_text(size = 18),
  
    axis.text = element_text(size = 30, margin = margin(t = 15)),
    axis.title = element_text(size = 30),
    axis.text.x = element_text(vjust = 1, hjust = 1),
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
  ) +
  guides(color = guide_legend(override.aes = list(size = 8)))
  

ggsave(file="plots/c3/DG_plots/av_rawc3_chl.png", width = 16, height = 10, dpi = 300)



#individual plots every day

# get unique mmyy values (ordered chronologically)

mmyy_levels <- final_dg %>%
  mutate(mmyy_date = lubridate::my(mmyy)) %>%
  arrange(mmyy_date) %>%
  pull(mmyy) %>%
  unique



# loop
for (m in mmyy_levels) {
  
  df_sub <- final_dg %>%
    filter(!(mmyy == "03-24" & chl < 13.43)) %>% 
    filter(mmyy == m) %>%
    arrange(date_time)
  
  
  p <- ggplot() +
   
    #geom_point(size = 0.5, alpha = 0.5, color="red") +
    #geom_point(aes(y=df_sub_2$chl), size = 0.1, alpha = 0.5, color="purple") +
    
    
    
  geom_point(data = df_sub, aes(x = date_time, y = chl), color = "red", size=0.1) +
    
    
    labs(
      title = paste("Chlorophyll Time Series -", m),
      x = "Date-Time",
      y = "Chlorophyll (µg/L)"
    ) +
    theme_bw() +
    theme(
      text = element_text(size = 10, colour = "black"),
      axis.text.x = element_text(size=3, angle = 45, hjust = 1)
    )
  
  ggsave(
    filename = paste0("DG_plots/monthly_chl_plots/chl_timeseries_", m, ".png"),
    plot = p,
    width = 8,
    height = 4,
    dpi = 300
  )
}


 #############################################

#every week plots


# loop
for (m in mmyy_levels) {
  
  df_sub <- final_dg %>%
    filter(mmyy == m) %>%
    arrange(date_time)
  
  
  p <- ggplot() +
   
    #geom_point(size = 0.5, alpha = 0.5, color="red") +
    #geom_point(aes(y=df_sub_2$chl), size = 0.1, alpha = 0.5, color="purple") +
    
    
    
  geom_point(data = df_sub, aes(x = date_time_parsed, y = chl), color = "red", size=0.1) +
    
    
    labs(
      title = paste("Chlorophyll Time Series -", m),
      x = "Date-Time",
      y = "Chlorophyll (µg/L)"
    ) +
    theme_bw() +
    theme(
      text = element_text(size = 10, colour = "black"),
      axis.text.x = element_text(angle = 45, hjust = 1)) +
    scale_x_datetime(
    date_breaks = "3 days",   # Change to "2 weeks" or "10 days" as needed
    date_labels = "%d-%b-%y")      # Formats as "Jan 2023"
  

  ggsave(
    filename = paste0("DG_plots/weekly_chl_plots/chl_timeseries_", m, ".png"),
    plot = p,
    width = 8,
    height = 4,
    dpi = 300
  )
}



```



+   visualize predicted vs observed values


```{r}

#check range before applying

range(c3_model_data$chl)
range(wierd_dg$chl)

#log transform chl

wierd_dg$log_chl <- log10(wierd_dg$chl)


# 1. Extract Fixed Effects and Sigma
coeffs <- fixef(model_mixed)
b0 <- coeffs["(Intercept)"]
b1 <- coeffs["log_chl"]
sig <- sigma(model_mixed)

# 2. Calculate Bias Correction Factor (BCF)
bcf_mm <- exp((sig^2 / 2) * (log(10)^2))

# 3. Predict for your dataset
wierd_dg <- wierd_dg %>%
  mutate(
    log_pred_mm   = b0 + (b1 * log_chl),  #predict data using the model
    pred_ext_mm = (10^log_pred_mm) * bcf_mm   #back transform
  )


write_csv(wierd_dg, file = "Corrected_Dunga_c3_chlorophyll.csv")





#visualise the predicted chl with the c3 chl

c3_long <- out_qc_loessdgc3 %>%
  mutate(
    month = as.numeric(month),
    year  = as.factor(year)
  ) %>%
  group_by(year, month) %>%
  summarise(
    avg_pred = mean(pred_ext_chl, na.rm = TRUE),
    avg_obs  = mean(chl, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_longer(
    cols = c(avg_pred, avg_obs),
    names_to = "type",
    values_to = "chl"
  ) %>%
  mutate(
    type = dplyr::recode(type,
                  avg_pred = "Predicted",
                  avg_obs  = "c3 chl"),
    month = factor(month, levels = 1:12, labels = month.abb)
  )

# Plotting Raw vs Predicted by Year

ggplot(c3_long, aes(x = month, y = chl, color = type, group = type)) +
  geom_line(linewidth = 1.5) +
  geom_point(size = 3) +
  facet_wrap(~ year, scales = "free_y") +
  scale_color_manual(values = c("C3_chlorophyll_RFU" = "black", "Predicted_chl_ug/l" = "red")) +
  labs(
    x = "Month",
    y = "Chlorophyll",
    color = ""
  ) +
  theme_bw(base_size = 18) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_line(colour = "gray90", linetype = "dotted"),
    legend.position = "top",
    strip.text = element_text(size = 16, face = "bold")
  )

ggsave(file="plots/c3/DG_plots/av_predictedvsobs_chl.png", width = 16, height = 10, dpi = 300)




###########################################################




```




+ Apply model to experiment data

```{r}


#c3_spiking_expt24

  #log transform chl
c3_spiking_expt24$log_chl <- log10(c3_spiking_expt24$chl)

#predict new chl log transformed
c3_spiking_expt24$pred_log <- predict(model_log, newdata = c3_spiking_expt24)

#add the log bias correction term: converting predictions from log space back to real chlorophyll units, with bias correction

sigma_log <- summary(model_log)$sigma^2

c3_spiking_expt24$pred_ext_chl <- 10^(c3_spiking_expt24$pred_log) * exp(sigma_log / 2)



#visualize

plot(c3_spiking_expt24$chl, c3_spiking_expt24$pred_ext_chl)

plot_data <- c3_spiking_expt24 %>%
  pivot_longer(
    cols = c(pred_ext_chl, ext_chl),
    names_to = "type",
    values_to = "chl_value"
  ) %>%
  mutate(
    type = dplyr::recode(type,
                         pred_ext_chl = "Predicted",
                         ext_chl  = "Lab_Extracted")
  ) %>% 
  mutate(
    expt = ifelse(
      code %in% paste0("HG", 1:6),
      "Very-green-expt",
      "light-green-expt"
    )
  ) %>% 
  mutate(expt = as_factor(expt))

# Plotting Raw vs Predicted by Year

ggplot(plot_data, aes(x = code, y = chl_value, color = type, group = type)) +
  geom_line(linewidth = 1.5) +
  geom_point(size = 3) +
  facet_wrap(~expt) +
  #scale_color_manual(values = c("Lab_Extracted" = "black", "Predicted" = "red")) +
  labs(
    x = "treatments",
    y = "Chlorophyll (µg/L)",
    color = ""
  ) +
  theme_bw(base_size = 18) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "top",
    strip.text = element_text(size = 16, face = "bold")
  )

ggsave(file="plots/c3/expts24_plot.png", width = 16, height = 10, dpi = 300)

 


```

